---
title: "Lyme Disease Differential Gene Expression Re-Analysis"
subtitle: "Re-analysis of Bouquet et al. 2016 (GSE63085) using DESeq2 & clusterProfiler"
author: "Abed & Laszlo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center"
)
```

# 1. Environment Setup

## 1.1 Install Packages

```{r install-packages, eval=FALSE, include=FALSE}
# install.packages(c("tidyverse", "pheatmap", "RColorBrewer", "ggrepel",
#                     "gridExtra", "viridis", "ashr"))
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(c("DESeq2", "clusterProfiler", "msigdbr",
#                         "enrichplot", "org.Hs.eg.db", "GSVA", "SPIA"))
```

```{r load-libraries}
# if (rstudioapi::isAvailable()) {setwd(dirname(rstudioapi::getActiveDocumentContext()$path))}
options(width = 200)

# use this function to install and load packages
using <- function(pkgs) {
  for (pkg in pkgs) {  # Loop through each package in the provided vector
    if (!requireNamespace(pkg, quietly = TRUE)) {    # If not installed, install it
      install.packages(pkg, repos = "https://cloud.r-project.org")}
    # Load the package:
    library(pkg, character.only = TRUE)}}

using(c("tidyverse", "pheatmap", "RColorBrewer", "ggrepel", "corrplot", "gridExtra", "viridis", "BiocManager","DESeq2", "org.Hs.eg.db", "clusterProfiler", "EnhancedVolcano","msigdbr", "PCAtools", "GSVA", "enrichplot", "ashr", "matrixStats"))

# SPIA is Bioconductor-only (3rd-generation pathway topology); not on CRAN
if (!requireNamespace("SPIA", quietly = TRUE))
  BiocManager::install("SPIA", ask = FALSE, update = FALSE)
suppressPackageStartupMessages(library(SPIA))

# description of libraries:
# tidyverse       - data wrangling + ggplot2
# DESeq2          - differential expression (negative binomial model)
# clusterProfiler - pathway enrichment (ORA, GSEA)
# msigdbr         - access to MSigDB gene sets
# enrichplot      - enrichment result visualizations
# pheatmap        - heatmaps with clustering
# RColorBrewer    - colour palettes
# ggrepel         - non-overlapping text labels on plots
# gridExtra       - arrange multiple plots
# viridis         - colour-blind-friendly palettes
# matrixStats     - fast row/column statistics

theme_set(theme_bw(base_size = 13)) # this is for the plots 
```


We had to use explicit `dplyr::` prefixes on key functions (`filter`, `select`, `count`, `mutate`, `arrange`) throughout this document because
`clusterProfiler` loads packages that mask some of these names.


## 1.3 Directories

```{r directories}
project_dir <- getwd()
data_dir    <- file.path(project_dir, "Data")
results_dir <- file.path(project_dir, "results")
plots_dir   <- file.path(project_dir, "plots")

for (d in c(results_dir, plots_dir))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
```

---
---

# 2. Data Loading

The raw FASTQ reads from GEO (GSE63085) were re-aligned by Thomas to the modern **GRCh38.p13** human reference genome (replacing
the outdated hg19 used in the original 2016 study) and quantified into raw integer counts, the proper input for DESeq2. Thomas used Kallisto alligner. And more importantly, the 22 missing samples were excluded during re-alignment, will be checked and confirmed by Thomas. 

**References:**

- Bouquet J et al. *mBio* 2016; 7(1):e00100-16.
- Thomas's book T. *Beginner's Handbook of -Omics Data Analysis*, Ch. 7-8.

## 2.1 Raw Counts Matrix

```{r load-counts}
counts_raw <- read.delim(
  gzfile(file.path(data_dir, "GSE63085_raw_counts_GRCh38.p13_NCBI.tsv.gz")),
  row.names = 1, check.names = FALSE
)
cat("Counts matrix:", nrow(counts_raw), "genes x", ncol(counts_raw), "samples\n")
```

## 2.2 Gene Annotation (featureData)

This maps NCBI Entrez Gene IDs to gene symbols, biotypes, chromosomal locations, and GO terms.

NCBI Entrez as the "Google Search" engine for biological data. It’s a massive integrated system managed by the National Center for Biotechnology Information (NCBI) that links together different databases—DNA sequences, protein structures, and medical literature (PubMed).
According to LLMs, when we see an Entrez ID (like those numbers in your matrix), it’s a unique "social security number" for a gene. Names (Symbols) change as scientists argue, but Entrez IDs stay the same.

```{r load-gene-annot}
gene_annot <- read.delim(
  gzfile(file.path(data_dir, "Human.GRCh38.p13.annot.tsv.gz")),
  stringsAsFactors = FALSE
)
cat("Gene annotation:", nrow(gene_annot), "x", ncol(gene_annot), "\n")
```
> Explanation of the columns (source: LLMS):

* **GeneID**: This matches the row names in your counts matrix. It’s your primary key.

* **Symbol**: The "human-friendly" name (e.g., TP53, BRCA1). This is what you’ll put in your final paper or report.

* **Description**: A long-form name of what the gene actually does (e.g., "tumor protein p53").

* **EnsemblGeneID**: Another ID system (starting with ENSG...). Bioinformaticians often use both Entrez and Ensembl because some tools prefer one over the other.

* **GeneType**: Usually, you only want to analyze protein-coding genes. Most researchers filter out the pseudo (dead "fossil" genes) and ncRNA (non-coding) unless they are specifically looking for them.

* **Status**: You should likely filter for "active". A "discontinued" status means the gene was merged into another record or found to be a mistake in the genome mapping.

* **ChrAcc, ChrStart, ChrStop, Orientation**: These are the "Location" columns. They tell you exactly where the gene sits on which chromosome and which way it's facing. This is vital if you want to see if a treatment affects a whole neighborhood of genes (a "cluster") or just one.

* **GOFunction / GOProcess / GOComponent**: "GO" stands for Gene Ontology. These tell you the "job" of the gene. **Process**: What is the "big picture" goal? (e.g., "Cell Division"); **Function**: What is the specific biochemical task? (e.g., "Binding to DNA"); and **Component**: Where does it happen? (e.g., "Mitochondria").


```{r meta_EDA}
names(gene_annot)
cat("\n\nGene biotypes:"); print(table(gene_annot$GeneType))

cat("\n\nStatus unique:");print(table(gene_annot$Status))
```
> We have ~39k total entries, but only 19,416 are protein-coding. 

## 2.3 Sample Annotation

```{r load-sample-annot}
sample_annot <- read.csv(
  file.path(project_dir, "sample_annotations.csv"),
  stringsAsFactors = FALSE
)
```

## 2.4 Align Samples

22 samples from the annotation are absent from the counts matrix.
The reason is not confirmed -- ask Thomas whether they were excluded
during re-alignment or simply not processed.

```{r align-data}
shared     <- intersect(colnames(counts_raw), sample_annot$GSM)
only_annot <- setdiff(sample_annot$GSM, colnames(counts_raw))

cat("Shared:", length(shared), "| Not in counts:", length(only_annot), "\n")

counts <- counts_raw[, shared]
annot  <- sample_annot %>%
  dplyr::filter(GSM %in% shared) %>%
  dplyr::arrange(match(GSM, shared))

stopifnot(all(colnames(counts) == annot$GSM))
```

## 2.5 Engineer Metadata Columns

```{r engineer-metadata}
annot <- annot %>%
  dplyr::mutate(
    patient_id = gsub("^.*-([0-9]+)_V.*$", "\\1",
                      gsub("^.*-([0-9]+)_V$", "\\1", Observation)),
    visit = dplyr::case_when(
      grepl("V1", Observation) ~ "V1",
      grepl("V2", Observation) ~ "V2",
      grepl("V5", Observation) ~ "V5",
      TRUE                     ~ "Control"),
    group = dplyr::case_when(
      disease.state == "Healthy controls" ~ "Control",
      visit == "V1" ~ "Lyme_V1",
      visit == "V2" ~ "Lyme_V2",
      visit == "V5" ~ "Lyme_V5"),
    condition = ifelse(disease.state == "Healthy controls", "Control", "Lyme"),
    group     = factor(group,     levels = c("Control","Lyme_V1","Lyme_V2","Lyme_V5")),
    visit     = factor(visit,     levels = c("Control","V1","V2","V5")),
    condition = factor(condition, levels = c("Control","Lyme"))
  )
rownames(annot) <- annot$GSM

cat("Samples per group:\n")
print(table(annot$group))
```

---
---

# 3. Gene Biotype Overview

```{r gene-biotypes}
gene_annot_matched <- gene_annot %>%
  dplyr::filter(GeneID %in% as.numeric(rownames(counts)))

biotype_counts <- gene_annot_matched %>%
  dplyr::count(GeneType, sort = TRUE) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))

ggplot(biotype_counts %>% dplyr::filter(n > 50),
       aes(x = reorder(GeneType, n), y = n, fill = GeneType)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(n, " (", pct, "%)")), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Gene Biotypes in Counts Matrix (GRCh38.p13)",
       x = NULL, y = "Number of Genes") +
  scale_fill_viridis_d()
```

About half the genes are **protein-coding** (~49%).
The rest are non-coding RNA, pseudogenes, etc. According to LLMs, protein-coding genes carry the strongest biological signal for differential expression. \n 
After filtering (Section 6), the protein-coding proportion will increase because non-coding genes tend to have very low counts.

---

# 4. Sample-Level Quality Control

*Ref: Thomas's book Ch. 7.2*

## 4.1 Library Sizes

Library size = total mapped reads per sample. Large variation means
different sequencing depths, which must be normalised.

```{r library-sizes, fig.height=10}
lib_sizes <- data.frame(
  GSM = colnames(counts), total = colSums(counts)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group, Observation), by = "GSM")

ggplot(lib_sizes, aes(x = reorder(Observation, total),
                       y = total / 1e6, fill = group)) +
  geom_col() + coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library Size per Sample",
       x = NULL, y = "Total Counts (millions)", fill = "Group") +
  theme(axis.text.y = element_text(size = 5))
```

```{r lib-size-boxplot}
ggplot(lib_sizes, aes(x = group, y = total / 1e6, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library Size by Group", x = NULL, y = "Total Counts (millions)") +
  theme(legend.position = "none")
```

If any group has systematically lower library sizes,
the DESeq2 size-factor normalisation (Section 7) will correct for this.
Look for outlier samples that are dramatically different from the rest --
these might warrant investigation.

## 4.2 Detected Genes per Sample

```{r detected-genes}
detected <- data.frame(
  GSM = colnames(counts), n_detected = colSums(counts > 0)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(detected, aes(x = group, y = n_detected, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Detected Genes per Sample (count > 0)",
       x = NULL, y = "Genes Detected") +
  theme(legend.position = "none")
```

## 4.3 Relative Log Expression (RLE)

*Ref: Thomas's book Ch. 7.2.4*

RLE measures how much each gene in a sample deviates from its median
expression across all samples. Well-behaved samples centre around zero.
Samples with shifted or wide RLE distributions may indicate technical
problems or batch effects.

```{r rle-plot, fig.height=8}
# Compute RLE: for each gene, subtract the row median
log2_counts <- log2(counts + 1)
row_medians <- rowMedians(as.matrix(log2_counts))
rle_matrix  <- sweep(log2_counts, 1, row_medians)

rle_long <- rle_matrix %>%
  as.data.frame() %>%
  tidyr::pivot_longer(everything(), names_to = "GSM", values_to = "RLE") %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(rle_long, aes(x = GSM, y = RLE, fill = group)) +
  geom_boxplot(outlier.size = 0.1) +
  scale_fill_brewer(palette = "Set2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Relative Log Expression (RLE) Plot",
       subtitle = "Well-behaved samples centre around the red dashed line (zero)",
       x = NULL, y = "Deviation from median expression") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 3)) +
  coord_cartesian(ylim = c(-2, 2))
```

**Bingo:** Samples with medians shifted away from zero or with
unusually wide boxes may have quality issues. If all groups looked similar,
there wouldn't be strong batch effect in the raw data. But apparently, they don't, so there could be, supporting our hypothesis that there could be batch effects.

---
---

# 5. Expression Distributions

## 5.1 Per-Sample Density

```{r density-plots}
log2_long <- log2_counts %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  tidyr::pivot_longer(-gene, names_to = "GSM", values_to = "log2_count") %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(log2_long, aes(x = log2_count, group = GSM, color = group)) +
  geom_density(linewidth = 0.3) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Per-Sample Density of log2(count + 1)",
       x = "log2(count + 1)", y = "Density", color = "Group") +
  xlim(0, 20)
```

the large spike near zero represents the many genes
with zero or very low expression. Most of them are still there because we haven't filtered out the garbage yet. The second hump (should be around 5-10) represents actively expressed genes. What we are looking for looks flat because the scale is zoomed out to accommodate the zero spike. All samples should have a similar shape; if
one looks radically different, it may be a quality outlier. Perhaps, we should make a similar plot after filtering.

## 5.2 Mean-Variance Relationship

```{r mean-variance}
gene_stats <- data.frame(
  mean_count = rowMeans(counts), var_count = apply(counts, 1, var)
)

ggplot(gene_stats, aes(x = log10(mean_count + 1), y = log10(var_count + 1))) +
  geom_point(alpha = 0.05, size = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 0.8) +
  labs(title = "Mean-Variance Relationship (Raw Counts)",
       subtitle = "Red dashed = Poisson (variance = mean). Points above = overdispersion.",
       x = "log10(Mean + 1)", y = "log10(Variance + 1)")
```

Nearly all genes sit above the red line, meaning variance > mean **if my mean is set correctly**. According to AI: this overdispersion is exactly why DESeq2 uses a negative binomial model instead of a simpler Poisson model. A Poisson
model would dramatically underestimate variability and produce too many false positives.

## 5.3 MA Plot (pre-normalisation)

*Ref: Thomas's book Ch. 7.2.7*

An MA plot shows each gene's **average expression** (A) vs. its **log
fold change** (M) between two conditions. Before normalisation, the cloud
should centre around M=0 (most genes are not differentially expressed).
A shift or banana shape indicates a normalisation problem.

```{r ma-plot-pre-norm, fig.height=5}
# Quick MA: V1 (first patient) vs first Control
ctrl_cols  <- annot$GSM[annot$group == "Control"]
lyme_cols  <- annot$GSM[annot$group == "Lyme_V1"]

A <- 0.5 * (rowMeans(log2_counts[, ctrl_cols]) + rowMeans(log2_counts[, lyme_cols]))
M <- rowMeans(log2_counts[, lyme_cols]) - rowMeans(log2_counts[, ctrl_cols])

ma_df <- data.frame(A = A, M = M)

ggplot(ma_df, aes(x = A, y = M)) +
  geom_point(alpha = 0.05, size = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "orange", se = FALSE, linewidth = 0.8) +
  labs(title = "MA Plot: Lyme V1 vs Control (pre-normalisation)",
       subtitle = "Red = zero line. Orange = loess trend. Centred cloud = good.",
       x = "A (average expression)", y = "M (log2 fold change)") +
  coord_cartesian(ylim = c(-5, 5))
```

Cloud isn't centered around 0, indicating a scaling issue. 

---
---

# 6. Gene Filtering

We remove genes with very low counts to reduce noise. We require at
least N samples (= smallest group size) to have a count >= 10.

```{r prefilter}
smallest_group <- min(table(annot$group))
keep <- rowSums(counts >= 10) >= smallest_group
counts_filtered <- counts[keep, ]

cat("Before:", nrow(counts), "genes\n")
cat("After: ", nrow(counts_filtered), "genes\n")
cat("Removed:", nrow(counts) - nrow(counts_filtered),
    "(", round(100 * (1 - nrow(counts_filtered)/nrow(counts)), 1), "%)\n")
```

```{r post-filter-biotypes}
biotype_after <- gene_annot %>%
  dplyr::filter(GeneID %in% as.numeric(rownames(counts_filtered))) %>%
  dplyr::count(GeneType, sort = TRUE) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))
print(biotype_after)
```

**Meaning:** After aggressive filtering, ~70% of remaining genes are **protein-coding** (up from ~49%), confirming that most low-count genes were non-coding RNA and pseudogenes. This is expected and desirable.

---
---

# 7. DESeq2 Normalisation

*Ref: Thomas's book Ch. 7.6e*

## 7.1 Create DESeqDataSet

The `design = ~ group` formula tells DESeq2 to model expression as a function of our four experimental groups.

```{r create-dds}
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered, colData = annot, design = ~ group)
```



## 7.2 Size Factors

DESeq2 estimates a **size factor** per sample via the median-of-ratios method. This corrects for sequencing depth differences and is far superior to the FPKM normalisation used in the original study.

```{r size-factors}
dds <- estimateSizeFactors(dds)

sf_df <- data.frame(
  GSM = colnames(dds), size_factor = sizeFactors(dds),
  lib_size = colSums(counts_filtered)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(sf_df, aes(x = lib_size / 1e6, y = size_factor, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "gray40", linetype = "dashed") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "DESeq2 Size Factor vs Library Size",
       x = "Library Size (millions)", y = "Size Factor")
```

Size factor and library size are strongly correlated (expected). 
Samples with a size factor < 1 were sequenced less deeply; > 1 means more deeply. 
DESeq2 divides each sample's counts by its size factor during analysis.

## 7.3 MA Plot (after size-factor normalization)

*Ref: Thomas's book Ch. 7.2.7*

Once **size factors** are applied, `counts(dds, normalized = TRUE)` is on a comparable scale across samples. We pool Lyme V1 vs Control and plot **M** = log2 ratio of group means vs **A** = average log2 signal (same construction as Section 5.3, but using median-of-ratios normalized counts). The loess trend should sit near **M = 0** across most expression levels if depth correction worked; a strong banana or vertical offset would suggest residual technical bias (or real biology concentrated at certain expression ranges).

```{r ma-plot-norm-counts, fig.height=5}
nt <- counts(dds, normalized = TRUE)
ctrl_cols <- annot$GSM[annot$group == "Control"]
v1_cols   <- annot$GSM[annot$group == "Lyme_V1"]

m1 <- rowMeans(nt[, v1_cols, drop = FALSE])
m0 <- rowMeans(nt[, ctrl_cols, drop = FALSE])
A_norm <- 0.5 * (log2(m1 + 0.5) + log2(m0 + 0.5))
M_norm <- log2((m1 + 0.5) / (m0 + 0.5))

ggplot(data.frame(A = A_norm, M = M_norm), aes(x = A, y = M)) +
  geom_point(alpha = 0.05, size = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "orange", se = FALSE, linewidth = 0.8) +
  labs(title = "MA Plot: V1 vs Control (DESeq2 normalized counts)",
       subtitle = "After size factors only (GLM / dispersions not yet fitted).",
       x = "A (average log2 normalized signal)", y = "M (log2 fold change)") +
  coord_cartesian(ylim = c(-5, 5))
```

Compare with the pre-normalisation MA (Section 5.3): the cloud should be **more tightly centred** on M = 0 after size factors. A later **DESeq2 `plotMA()`** after the full model (Section 10.1) adds dispersion shrinkage and hypothesis testing; that plot is the standard check bundled with DESeq2.

## 7.4 Variance Stabilising Transformation (VST)

Raw counts have a mean-variance dependency (Section 5.2). VST produces
values on a log2-like scale with approximately **constant variance**,
suitable for PCA and heatmaps.

```{r vst}
vsd <- vst(dds, blind = FALSE)
```

```{r vst-effect, fig.height=5}
par(mfrow = c(1, 2))
plot(rowMeans(log2(counts_filtered + 1)),
     apply(log2(counts_filtered + 1), 1, sd),
     pch = 16, cex = 0.3, col = rgb(0.2, 0.4, 0.8, 0.1),
     xlab = "Mean", ylab = "SD", main = "Before: log2(counts+1)")
plot(rowMeans(assay(vsd)), apply(assay(vsd), 1, sd),
     pch = 16, cex = 0.3, col = rgb(0.8, 0.2, 0.2, 0.1),
     xlab = "Mean", ylab = "SD", main = "After: VST")
par(mfrow = c(1, 1))
```



Before VST (left), low-expression genes have variable SD. After VST (right), the trend should be flattened, but it's not, so the SD still depends on the mean. This makes the meaningfullness of distances between samples (for PCA, clustering) questionable.



The following can verify whether before and after actually look too similar

```{r}
mean_sd_before <- cor(
  rowMeans(log2(counts_filtered + 1)),
  apply(log2(counts_filtered + 1), 1, sd)
)

mean_sd_after <- cor(
  rowMeans(assay(vsd)),
  apply(assay(vsd), 1, sd)
)

mean_sd_before
mean_sd_after
```

Confirm with Thomas!


---
---

# 8. PCA

*Ref: Thomas's book Ch. 7.2.2*

```{r pca-compute}
rv      <- rowVars(assay(vsd))
top_idx <- order(rv, decreasing = TRUE)[1:min(2000, length(rv))]

pca     <- prcomp(t(assay(vsd)[top_idx, ]), center = TRUE, scale. = TRUE)
var_pct <- round(100 * summary(pca)$importance[2, ], 1)

pca_df  <- data.frame(pca$x[, 1:5]) %>%
  tibble::rownames_to_column("GSM") %>%
  dplyr::left_join(annot, by = "GSM")
```

## 8.1 Scree Plot

```{r scree}
ggplot(data.frame(PC = 1:20, Var = var_pct[1:20]),
       aes(x = PC, y = Var)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_line(color = "darkred") + geom_point(color = "darkred", size = 2) +
  labs(title = "Scree Plot", x = "Principal Component",
       y = "% Variance Explained") +
  scale_x_continuous(breaks = 1:20)
```

## 8.2 PCA by Group

```{r pca-group, fig.width=15, fig.height=5}
make_pca <- function(df, pcx, pcy, vpct) {
  ggplot(df, aes(x = .data[[pcx]], y = .data[[pcy]], color = group)) +
    geom_point(size = 3, alpha = 0.8) +
    stat_ellipse(level = 0.68, linetype = "dashed") +
    scale_color_brewer(palette = "Set2") +
    labs(x = paste0(pcx, " (", vpct[as.integer(gsub("PC","",pcx))], "%)"),
         y = paste0(pcy, " (", vpct[as.integer(gsub("PC","",pcy))], "%)"))
}

grid.arrange(
  make_pca(pca_df,"PC1","PC2",var_pct) + ggtitle("PC1 vs PC2"),
  make_pca(pca_df,"PC1","PC3",var_pct) + ggtitle("PC1 vs PC3"),
  make_pca(pca_df,"PC2","PC3",var_pct) + ggtitle("PC2 vs PC3"),
  ncol = 3
)
```

**According to AI:** Look for how well the four groups separate.
**PC1** (largest variance axis) typically separates disease from
control. If V1 (acute Lyme) clusters distinctly from controls, it
confirms that infection drives a strong transcriptomic shift. V5 samples
sitting closer to controls suggests partial recovery of gene expression
by 6 months -- consistent with the original paper's finding that the
signature *persists* but *weakens*.

## 8.3 Patient Trajectories

Lines connect the same patient across V1 -> V2 -> V5.

```{r pca-trajectories, fig.height=7}
lyme_pca <- pca_df %>% dplyr::filter(condition == "Lyme") %>%
  dplyr::arrange(patient_id, visit)

ggplot() +
  geom_point(data = pca_df %>% dplyr::filter(condition == "Control"),
             aes(x = PC1, y = PC2), color = "gray60", size = 3, alpha = 0.5) +
  geom_path(data = lyme_pca,
            aes(x = PC1, y = PC2, group = patient_id),
            color = "gray40", alpha = 0.3) +
  geom_point(data = lyme_pca,
             aes(x = PC1, y = PC2, color = visit, shape = visit), size = 3) +
  scale_color_manual(values = c(V1="#d62728", V2="#ff7f0e", V5="#1f77b4")) +
  labs(title = "Patient Trajectories Through PCA Space",
       subtitle = "Gray = controls. Lines connect same patient across visits.",
       x = paste0("PC1 (", var_pct[1], "%)"),
       y = paste0("PC2 (", var_pct[2], "%)"))
```

Each line represents one patient's journey from acute infection (red V1) through treatment (orange V2) to 6-month
follow-up (blue V5). Glücklicherweise, V5 points move toward the gray control cloud, showing recovery. Patients whose V5 stays far from controls are fucked because it's probably those with persistent symptoms (PTLDS).

---
---

# 9. Sample Relationships

*Ref: Thomas's book Ch. 7.2.1*

## 9.1 Distance Heatmap

```{r distance-heatmap, fig.height=9, fig.width=10}
sample_dists <- dist(t(assay(vsd)))
row_annot <- data.frame(Group = annot$group, row.names = annot$GSM)
ann_colors <- list(Group = c(Control="#66C2A5", Lyme_V1="#FC8D62",
                              Lyme_V2="#8DA0CB", Lyme_V5="#E78AC3"))

pheatmap(as.matrix(sample_dists),
         clustering_distance_rows = sample_dists,
         clustering_distance_cols = sample_dists,
         annotation_row = row_annot, annotation_col = row_annot,
         annotation_colors = ann_colors,
         show_rownames = FALSE, show_colnames = FALSE,
         color = viridis(100),
         main = "Sample-to-Sample Euclidean Distance (VST)")
```

## 9.2 Top 50 Most Variable Genes

```{r top50-heatmap, fig.height=12, fig.width=10}
top50_idx <- head(order(rv, decreasing = TRUE), 50)
top50_ids <- rownames(assay(vsd))[top50_idx]
top50_sym <- gene_annot$Symbol[match(as.numeric(top50_ids), gene_annot$GeneID)]
top50_lab <- ifelse(is.na(top50_sym) | top50_sym == "", top50_ids, top50_sym)

hm_mat <- assay(vsd)[top50_idx, ]
rownames(hm_mat) <- top50_lab

pheatmap(t(scale(t(hm_mat))),
         annotation_col = row_annot, annotation_colors = ann_colors,
         show_colnames = FALSE, fontsize_row = 7,
         color = colorRampPalette(c("navy","white","firebrick3"))(100),
         main = "Top 50 Most Variable Genes (Z-scored VST)",
         cutree_cols = 4)
```

Rows are genes, columns are samples. Red = high expression, blue = low expression (relative to the gene's mean). If the
heatmap clusters samples by group (check the colour bar at top), it confirms that the dominant source of variation is biological (disease state / time point), not technical noise.

---
---

# 10. Differential Expression with DESeq2

*Ref: Thomas's book Ch. 7.6*

## 10.1 Run DESeq2

```{r run-deseq}
dds <- DESeq(dds)
resultsNames(dds)
```

## 10.1a MA plots (DESeq2: normalized mean vs LFC)

After `DESeq()`, the object contains fitted dispersions and contrasts on the **same scale** as differential testing. `plotMA()` shows, for each gene, the **mean of normalized counts** (x-axis, "mean of normalized counts") vs **log2 fold change** (y-axis); points with padj below `alpha` are blue. For a successful normalization and contrast, the **bulk of points** should scatter around **LFC = 0** at moderate-to-high expression, with no strong systematic curve. These plots complement Section 7.3 (MA from normalized counts only) and Section 5.3 (raw log2 counts).

```{r plot-ma-deseq2, fig.width = 14, fig.height = 5}
par(mfrow = c(1, 3))
plotMA(
  results(dds, contrast = c("group", "Lyme_V1", "Control"), alpha = 0.05),
  main = "V1 vs Control", ylim = c(-5, 5))
plotMA(
  results(dds, contrast = c("group", "Lyme_V2", "Control"), alpha = 0.05),
  main = "V2 vs Control", ylim = c(-5, 5))
plotMA(
  results(dds, contrast = c("group", "Lyme_V5", "Control"), alpha = 0.05),
  main = "V5 vs Control", ylim = c(-5, 5))
par(mfrow = c(1, 1))
```

**Result validation:** If the cloud were systematically shifted away from LFC = 0 for most genes, we would revisit size factors, outliers, or design formula. Here the **MA after normalization** (Section 7.3) and these **MA after the full DESeq2 fit** together support using the contrasts for pathway analysis.

## 10.2 Extract, Shrink & Annotate Results

`lfcShrink()` with `ashr` produces moderated log2 fold-change estimates
that are more reliable for genes with low counts. *(Stephens 2016)*

```{r extract-results}
contrasts <- list(
  V1 = c("group", "Lyme_V1", "Control"),
  V2 = c("group", "Lyme_V2", "Control"),
  V5 = c("group", "Lyme_V5", "Control")
)

get_annotated_results <- function(dds, contrast, gene_annot) {
  res <- results(dds, contrast = contrast, alpha = 0.05)
  res_shrunk <- lfcShrink(dds, contrast = contrast, type = "ashr", res = res)

  as.data.frame(res_shrunk) %>%
    tibble::rownames_to_column("GeneID") %>%
    dplyr::mutate(GeneID_num = as.numeric(GeneID)) %>%
    dplyr::left_join(
      gene_annot %>% dplyr::select(GeneID, Symbol, Description, GeneType),
      by = c("GeneID_num" = "GeneID")) %>%
    dplyr::select(-GeneID_num) %>%
    dplyr::arrange(padj)
}

res_list <- lapply(contrasts, function(ct)
  get_annotated_results(dds, ct, gene_annot))
```

## 10.3 DEG Summary

```{r deg-summary}
deg_summary <- lapply(names(res_list), function(nm) {
  sig <- res_list[[nm]] %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1)
  data.frame(Comparison = paste0(nm, " vs Control"),
             Total = nrow(sig),
             Up = sum(sig$log2FoldChange > 0),
             Down = sum(sig$log2FoldChange < 0))
}) %>% dplyr::bind_rows()

print(deg_summary)
```

Compare with the original study (Limma/Voom on FPKM): V1=1,235, V2=1,060, V5=686 DEGs. 
Not a 100% sure why our numbers differ, but I assume it's because we use: 
(1) a different genome reference, 
(2) raw counts instead of FPKM,
(3) DESeq2 instead of Limma/Voom, and (4) different fold-change/FDR thresholds. \n 
The pattern should be similar: most DEGs at V1 (acute), declining through V2 and V5 as the transcriptome partially recovers.

## 10.4 Top DEGs at V1

```{r top-degs}
res_list$V1 %>%
  dplyr::filter(!is.na(padj) & padj < 0.05) %>%
  head(15) %>%
  dplyr::select(Symbol, log2FoldChange, padj, GeneType) %>%
  print()
```

## 10.5 Volcano Plots

```{r volcano, fig.width=16, fig.height=6}
make_volcano <- function(df, title) {
  df <- df %>%
    dplyr::mutate(sig = dplyr::case_when(
      is.na(padj) ~ "NS",
      padj < 0.05 & log2FoldChange > 1  ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"))
  top <- df %>% dplyr::filter(sig != "NS") %>% head(10)
  ggplot(df, aes(x = log2FoldChange, y = -log10(pvalue), color = sig)) +
    geom_point(alpha = 0.4, size = 1) +
    scale_color_manual(values = c(Up="firebrick", Down="steelblue", NS="gray70")) +
    geom_text_repel(data = top, aes(label = Symbol), size = 2.5,
                    max.overlaps = 15) +
    geom_vline(xintercept = c(-1,1), linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    labs(title = title, x = "log2 Fold Change", y = "-log10(p-value)") +
    theme(legend.position = "bottom")
}

grid.arrange(
  make_volcano(res_list$V1, "V1 (Acute) vs Control"),
  make_volcano(res_list$V2, "V2 (Post-Treatment) vs Control"),
  make_volcano(res_list$V5, "V5 (6 Months) vs Control"),
  ncol = 3
)
```

Each dot is a gene. The x-axis is the fold change (how much the gene changes in Lyme vs. Control); the y-axis is statistical significance. **Red dots** (upper-right) are significantly upregulated; **blue dots** (upper-left) are downregulated. \n The labeled genes are the most significant. We can see how the "cloud" narrows from V1 to V5, fewer genes are dramatically changed at 6 months, matching the original paper's finding of a gradually resolving but sustained signature.

## 10.6 DEG Overlap Across Time Points

```{r deg-overlap}
sig_genes <- lapply(res_list, function(df)
  df %>% dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1) %>%
    dplyr::pull(GeneID))

cat("V1:", length(sig_genes$V1), "  V2:", length(sig_genes$V2),
    "  V5:", length(sig_genes$V5), "\n")
cat("V1 & V2:", length(intersect(sig_genes$V1, sig_genes$V2)), "\n")
cat("V1 & V5:", length(intersect(sig_genes$V1, sig_genes$V5)), "\n")
cat("All 3:  ", length(Reduce(intersect, sig_genes)), "\n")
```

The overlap between V1 and V2 tells us how many genes remain differentially expressed even after antibiotic treatment.


## 10.7 Save DEG Tables

```{r save-degs}
for (nm in names(res_list))
  write.csv(res_list[[nm]],
            file.path(results_dir, paste0("DEG_", nm, "_vs_Control.csv")),
            row.names = FALSE)
```

---
---

# 11. 1st Generation: ORA (Over-Representation Analysis)

*Ref: Thomas's book Ch. 8.1.1, 8.4.1; Khatri et al. 2012*

ORA asks: "Are my DEGs enriched in known pathways more than expected
by chance?" It compares the fraction of DEGs in a pathway to the
background fraction (all tested genes).

## 11.1 Load MSigDB Gene Sets (targeted)

MSigDB is the companion database for GSEA (second gen. of pathway analysis). but it can also be used for ORA


I loaded only the collections we need to avoid memory issues. Otherwise I couldn't knit.
The full MSigDB is ~738 MB and crashes knitr's cache.

```{r load-msigdb, cache=FALSE}
msigdb_hallmark <- msigdbr(species = "Homo sapiens", collection = "H")
msigdb_gobp     <- msigdbr(species = "Homo sapiens", collection = "C5",
                            subcollection = "GO:BP")

cat("Hallmark:", nrow(msigdb_hallmark), "entries (",
    length(unique(msigdb_hallmark$gs_name)), "sets)\n")
cat("GO:BP:  ", nrow(msigdb_gobp), "entries (",
    length(unique(msigdb_gobp$gs_name)), "sets)\n")
```

```{r build-term-maps}
# Hallmark
t2g_hallmark <- msigdb_hallmark %>%
  dplyr::select(gs_name, gene_symbol)

# GO Biological Process
t2g_gobp <- msigdb_gobp %>%
  dplyr::select(gs_name, gene_symbol)
t2n_gobp <- msigdb_gobp %>%
  dplyr::select(gs_name, gs_description) %>%
  dplyr::distinct()
```

## 11.2 ORA: Hallmark (V1)

```{r ora-hallmark-v1}
sig_sym_v1 <- res_list$V1 %>%
  dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                !is.na(Symbol) & Symbol != "") %>%
  dplyr::pull(Symbol) %>% unique()

univ_sym <- res_list$V1 %>%
  dplyr::filter(!is.na(Symbol) & Symbol != "") %>%
  dplyr::pull(Symbol) %>% unique()

ora_hallmark_v1 <- enricher(
  gene = sig_sym_v1, universe = univ_sym,
  TERM2GENE = t2g_hallmark,
  pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 1 # BH here stands for Benjamini-Hochberg correction for multiple testing
)
```

BH in the code stands for Benjamini-Hochberg correction for multiple testing. And q value is the corrected p-value for multiple testing.
qvalueCutoff = 1 means that we are not interested in the q-values, only the p-values. We should confirm with Thomas whether it makes biological sense to use q-values here, given that the genes are immune related and the SD is wild.


```{r ora-hallmark-dotplot, fig.height=6}
if (nrow(as.data.frame(ora_hallmark_v1)) > 0) {
  dotplot(ora_hallmark_v1, showCategory = 15) +
    ggtitle("ORA: Hallmark Pathways Enriched in V1 DEGs")
}
```

The Hallmark collection contains 50 well-defined biological processes from MSigDB (Thomas's book Ch. 8.2). Enriched pathways tell us *which biological processes* are disrupted in acute Lyme disease. We should see how much it matches the original study's findings


## 11.3 ORA: GO Biological Process (V1)

```{r ora-gobp-v1}
ora_gobp_v1 <- enricher(
  gene = sig_sym_v1, universe = univ_sym,
  TERM2GENE = t2g_gobp, TERM2NAME = t2n_gobp,
  pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 1
)

cat("Significant GO:BP terms:", nrow(as.data.frame(ora_gobp_v1)), "\n")
```

```{r ora-gobp-dotplot, fig.height=9}
if (nrow(as.data.frame(ora_gobp_v1)) > 0) {
  dotplot(ora_gobp_v1, showCategory = 20) +
    ggtitle("ORA: Top GO Biological Process Terms (V1)")
}
```

## 11.4 ORA Across All Time Points (Hallmark)

```{r ora-all-timepoints}
ora_hallmark_all <- list()

for (tp in names(res_list)) {
  sig_sym <- res_list[[tp]] %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                  !is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
  univ <- res_list[[tp]] %>%
    dplyr::filter(!is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
  ora_hallmark_all[[tp]] <- enricher(
    gene = sig_sym, universe = univ,
    TERM2GENE = t2g_hallmark,
    pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 1)
}

cat("Enriched Hallmark pathways: V1=",
    nrow(as.data.frame(ora_hallmark_all$V1)),
    " V2=", nrow(as.data.frame(ora_hallmark_all$V2)),
    " V5=", nrow(as.data.frame(ora_hallmark_all$V5)), "\n")
```

---
---

# 12. 2nd Generation: GSEA (Gene Set Enrichment Analysis)

*Ref: Thomas's book Ch. 8.1.2.1, 8.4.3; Subramanian et al. 2005*

GSEA is a major improvement over ORA because it uses **all genes ranked
by fold change**, not just those passing an arbitrary significance
cutoff. It detects subtle, coordinated changes in pathways that ORA
misses. The original study did NOT perform GSEA -- this is a novel
contribution of our re-analysis.

## 12.1 Build Ranked Gene Lists

```{r gsea-ranked}
make_ranked_list <- function(res_df) {
  res_df %>%
    dplyr::filter(!is.na(Symbol) & Symbol != "" & !is.na(log2FoldChange)) %>%
    dplyr::arrange(dplyr::desc(log2FoldChange)) %>%
    dplyr::distinct(Symbol, .keep_all = TRUE) %>%
    dplyr::select(Symbol, log2FoldChange) %>%
    tibble::deframe()
}

ranked_lists <- lapply(res_list, make_ranked_list)
```

## 12.2 GSEA: Hallmark (V1)

```{r gsea-hallmark-v1}
gsea_hallmark_v1 <- GSEA(
  geneList = ranked_lists$V1,
  TERM2GENE = t2g_hallmark,
  pvalueCutoff = 1, eps = 0
)

gsea_h_df <- as.data.frame(gsea_hallmark_v1)
cat("Significant (padj<0.05):", sum(gsea_h_df$p.adjust < 0.05),
    "of", nrow(gsea_h_df), "pathways\n")
```

```{r gsea-dotplot, fig.height=8}
if (sum(gsea_h_df$p.adjust < 0.05) > 0)
  dotplot(gsea_hallmark_v1, showCategory = 20, split = ".sign") +
    facet_grid(~ .sign) +
    ggtitle("GSEA: Hallmark Pathways (V1 vs Control)")
```

The dotplot splits pathways into **activated** (enriched in upregulated genes) and **suppressed** (enriched in downregulated genes). The **Normalised Enrichment Score (NES)** measures how strongly the pathway's genes cluster at the top (activated) or bottom (suppressed) of the ranked list. Which pathways should we expect to be activated; translation-related pathways may be suppressed?


## 12.3 GSEA Enrichment Plots for Key Pathways

These "mountain plots" show how a pathway's member genes are distributed in the ranked list. The green line rises when it encounters pathway genes and falls otherwise. A peak at the **left** = upregulated; **right** =
downregulated.

```{r gsea-mountains, fig.height=4, fig.width=8, results='asis'}
key_pathways <- c("HALLMARK_INFLAMMATORY_RESPONSE",
                  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                  "HALLMARK_COMPLEMENT")

for (pw in key_pathways)
  if (pw %in% gsea_h_df$ID)
    print(gseaplot2(gsea_hallmark_v1, geneSetID = pw, title = pw))
```

## 12.4 GSEA Across All Time Points

```{r gsea-all-timepoints}
gsea_results <- list()
for (tp in names(ranked_lists))
  gsea_results[[tp]] <- GSEA(geneList = ranked_lists[[tp]],
                              TERM2GENE = t2g_hallmark,
                              pvalueCutoff = 1, eps = 0)
```

```{r nes-heatmap, fig.height=9, fig.width=7}
nes_all <- lapply(names(gsea_results), function(tp) {
  as.data.frame(gsea_results[[tp]]) %>%
    dplyr::select(ID, NES, p.adjust) %>%
    dplyr::mutate(timepoint = tp)
}) %>% dplyr::bind_rows()

# Keep pathways significant in at least one time point
sig_any <- nes_all %>% dplyr::filter(p.adjust < 0.05) %>%
  dplyr::pull(ID) %>% unique()

if (length(sig_any) > 0) {
  nes_mat <- nes_all %>%
    dplyr::filter(ID %in% sig_any) %>%
    dplyr::select(ID, NES, timepoint) %>%
    tidyr::pivot_wider(names_from = timepoint, values_from = NES) %>%
    tibble::column_to_rownames("ID") %>% as.matrix()
  nes_mat[is.na(nes_mat)] <- 0

  pheatmap(nes_mat,
           color = colorRampPalette(c("steelblue","white","firebrick"))(100),
           cluster_cols = FALSE,
           main = "GSEA NES: Hallmark Pathways Across Time Points",
           fontsize_row = 7)
}
```

This heatmap shows how each pathway's activity (NES) evolves over time. **Red** = activated, **blue** = suppressed. Pathways that remain red from V1 through V5 represent the **sustained signature** described in the paper. Pathways that fade to white by V5 indicate resolution.

---
---

# 13. Biologic Theme Comparison

*Ref: Thomas's book Ch. 8.4.2*

`compareCluster` from clusterProfiler directly compares enrichment
results across multiple gene lists (our three time points), making the
temporal dynamics visually clear.

```{r compare-cluster, fig.height=10, fig.width=10}
gene_lists <- lapply(res_list, function(df) {
  df %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                  !is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
})

cc_result <- compareCluster(
  geneClusters = gene_lists,
  fun = "enricher",
  TERM2GENE = t2g_hallmark,
  pvalueCutoff = 0.05, pAdjustMethod = "BH"
)

if (nrow(as.data.frame(cc_result)) > 0)
  dotplot(cc_result) +
    ggtitle("Biologic Theme Comparison: Hallmark Across Time Points") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Each column is a time point, each row is a pathway. Dot size = number of DEGs in that pathway; colour = adjusted p-value. This directly shows which pathways are shared across time points and which are specific to the acute phase (V1 only).


---
---

# 14. 2nd Generation (cont.): GSVA (Gene Set Variation Analysis)

*Ref: Thomas's book Ch. 8.1.2.2, 8.4.5; Hänzelmann et al. 2013*

GSVA completes the 2nd generation of pathway analysis. Unlike GSEA
(which compares two groups), GSVA computes a **per-sample pathway
score**. This preserves the full sample structure, so we can test
pathway differences with standard linear models -- exactly as Thomas
describes in Ch. 8.4.5.

Think of it this way:

- **ORA** (1st gen): "Which pathways have too many DEGs?" (binary: gene is DE or not)
- **GSEA** (2nd gen-a): "Do pathway genes cluster at the top/bottom of the ranked list?" (uses fold-change magnitude, but still compares two groups)
- **GSVA** (2nd gen-b): "For each sample individually, how active is this pathway?" (produces a score per sample per pathway -- then you can run any statistical model on those scores)

## 14.1 Prepare Gene Sets as Named Lists

GSVA needs gene sets as a named list of character vectors (gene symbols).

```{r gsva-genesets}
hallmark_sets <- msigdb_hallmark %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(unique(gene_symbol)), .groups = "drop") %>%
  tibble::deframe()

cat("Hallmark gene sets:", length(hallmark_sets), "\n")
cat("Example set sizes:", head(sapply(hallmark_sets, length)), "\n")
```

## 14.2 Prepare Expression Matrix with Gene Symbols

GSVA needs the matrix rows to be gene symbols (matching the gene sets),
not Entrez IDs.

```{r gsva-matrix}
# Use VST-normalised data for GSVA (Gaussian kernel, Ch. 8.4.5)
vst_mat <- assay(vsd)

# Map Entrez IDs to Symbols
symbol_map <- gene_annot$Symbol[match(as.numeric(rownames(vst_mat)),
                                       gene_annot$GeneID)]

# Keep only rows with valid symbols, handle duplicates by keeping highest mean
keep_sym <- !is.na(symbol_map) & symbol_map != ""
vst_sym  <- vst_mat[keep_sym, ]
rownames(vst_sym) <- symbol_map[keep_sym]

# Remove duplicate gene symbols (keep the one with highest mean expression)
dup_idx <- duplicated(rownames(vst_sym))
vst_sym <- vst_sym[!dup_idx, ]

cat("GSVA input matrix:", nrow(vst_sym), "genes x", ncol(vst_sym), "samples\n")
```

## 14.3 Run GSVA

```{r run-gsva}
gsva_param <- gsvaParam(
  exprData = vst_sym,
  geneSets = hallmark_sets,
  kcdf     = "Gaussian",   # Gaussian for continuous (VST) data; Poisson for counts
  maxDiff  = TRUE
)

gsva_scores <- gsva(gsva_param)

cat("GSVA output:", nrow(gsva_scores), "pathways x", ncol(gsva_scores), "samples\n")
```

## 14.4 GSVA Heatmap

Each cell shows how active a pathway is in a given sample (relative
to other samples). This reveals the per-patient, per-time-point
pathway landscape.

```{r gsva-heatmap, fig.height=12, fig.width=10}
pheatmap(gsva_scores,
         annotation_col = row_annot,
         annotation_colors = ann_colors,
         show_colnames = FALSE,
         fontsize_row = 6,
         color = colorRampPalette(c("steelblue", "white", "firebrick"))(100),
         main = "GSVA Scores: Hallmark Pathways (per sample)",
         clustering_distance_rows = "correlation",
         cutree_cols = 4)
```

The key advantage of GSVA over GSEA: you can see individual patient
variation. Some V5 patients may still have highly activated
inflammatory pathways (potential PTLDS candidates), while others
return to control-like levels.

## 14.5 Test GSVA Scores Between Groups

Following Thomas's Ch. 8.4.5: since GSVA scores are continuous values
per sample, we can use a standard linear model (limma) to test which
pathways differ significantly between groups. This mirrors the workflow
from Ch. 7.5.1.

```{r gsva-limma}
library(limma)

design_gsva <- model.matrix(~ 0 + group, data = annot)
colnames(design_gsva) <- levels(annot$group)

# Fit linear model to GSVA scores (instead of gene expression)
fit_gsva <- lmFit(gsva_scores, design_gsva)

# Define contrasts (same comparisons as DESeq2)
contrast_mat <- makeContrasts(
  V1_vs_Ctrl = Lyme_V1 - Control,
  V2_vs_Ctrl = Lyme_V2 - Control,
  V5_vs_Ctrl = Lyme_V5 - Control,
  levels = design_gsva
)

fit_gsva2 <- contrasts.fit(fit_gsva, contrast_mat) %>% eBayes()
```

```{r gsva-results}
gsva_results_list <- list()
for (coef_name in colnames(contrast_mat)) {
  tt <- topTable(fit_gsva2, coef = coef_name, number = Inf) %>%
    tibble::rownames_to_column("pathway")
  gsva_results_list[[coef_name]] <- tt
  
  n_sig <- sum(tt$adj.P.Val < 0.05)
  cat(coef_name, ":", n_sig, "pathways with adj.P.Val < 0.05\n")
}
```

## 14.6 GSVA Volcano-Style Plot

```{r gsva-volcano, fig.width=16, fig.height=6}
gsva_volcano_plots <- lapply(names(gsva_results_list), function(nm) {
  df <- gsva_results_list[[nm]] %>%
    dplyr::mutate(
      sig = ifelse(adj.P.Val < 0.05, "Significant", "NS"),
      label = ifelse(adj.P.Val < 0.05,
                     gsub("HALLMARK_", "", pathway), NA)
    )
  ggplot(df, aes(x = logFC, y = -log10(P.Value), color = sig)) +
    geom_point(size = 2.5) +
    scale_color_manual(values = c(Significant = "firebrick", NS = "gray60")) +
    geom_text_repel(aes(label = label), size = 2.2, max.overlaps = 20,
                    na.rm = TRUE) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    labs(title = gsub("_", " ", nm),
         x = "GSVA score difference", y = "-log10(p-value)") +
    theme(legend.position = "none")
})

grid.arrange(grobs = gsva_volcano_plots, ncol = 3)
```

Each dot is a Hallmark pathway. The x-axis is the difference in GSVA
score between Lyme and Control (positive = more active in Lyme).
Labeled pathways are significant (adj.P.Val < 0.05). Compare the
three panels: V1 should have the most extreme differences, V5 the
least -- mirroring the gene-level DEG pattern.

## 14.7 GSVA Score Boxplots for Key Pathways

```{r gsva-boxplots, fig.height=10, fig.width=12}
key_pw <- c("HALLMARK_INFLAMMATORY_RESPONSE",
            "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
            "HALLMARK_INTERFERON_GAMMA_RESPONSE",
            "HALLMARK_COMPLEMENT",
            "HALLMARK_IL6_JAK_STAT3_SIGNALING",
            "HALLMARK_APOPTOSIS")

gsva_long <- gsva_scores[rownames(gsva_scores) %in% key_pw, ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("pathway") %>%
  tidyr::pivot_longer(-pathway, names_to = "GSM", values_to = "score") %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group, visit), by = "GSM") %>%
  dplyr::mutate(pathway = gsub("HALLMARK_", "", pathway))

ggplot(gsva_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.6) +
  facet_wrap(~ pathway, scales = "free_y", ncol = 3) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "GSVA Scores for Key Immune Pathways",
       subtitle = "Each dot is a patient. Higher = more pathway activity in that sample.",
       x = NULL, y = "GSVA Score") +
  theme(legend.position = "none",
        strip.text = element_text(size = 8))
```

These boxplots are the unique output of GSVA that neither ORA nor GSEA
can produce. Each dot is an individual patient's pathway activity.
Look for:

- **V1 >> Control**: Pathway is activated during acute infection (expected for immune/inflammatory pathways).
- **V2 still elevated**: Pathway stays active even after antibiotic treatment -- evidence for the sustained signature.
- **V5 approaching Control**: Pathway is resolving (good sign for most patients).
- **V5 patients still elevated**: These outlier dots may correspond to patients with persistent symptoms (PTLDS).

This connects directly to the paper's central finding: the gene expression
signature of Lyme disease persists after treatment. GSVA lets us see
this at the **individual patient level**, which the original study's
bulk comparisons could not show.

---
---

# 15. Third Generation: SPIA (topology-aware KEGG perturbation)

*Ref: Khatri et al. 2012 (three generations); Tarca et al. 2009; Thomas's book Ch. 8 — pathway topology*

**What “third generation” means here.** First-generation ORA asks whether a pathway contains more DEG labels than expected. Second-generation GSEA/GSVA uses ranked scores or per-sample enrichment without explicit **wiring** between genes. **Signalling Pathway Impact Analysis (SPIA)** adds a *perturbation* term: it walks curated **KEGG** signalling topology so that genes upstream/downstream and activation vs inhibition relations contribute differently from a flat hypergeometric count. The combined evidence (over-representation **pNDE** + topology **pPERT**) yields a global **pG** (and FDR-style **pGFdr**).

### Should we trust SPIA for this course project?

SPIA is pedagogically valuable: it makes the “topology” idea concrete and produces interpretable **Activated / Inhibited** status per pathway. For a *scientific* discussion with your professor, consider these limitations:

- **Database scope:** SPIA is tied to **KEGG** pathway maps (not MSigDB Hallmark). Immune–infection biology is only partially captured; many Lyme-relevant processes appear in Hallmark/GO instead.
- **Model assumptions:** Edge signs and “perturbation accumulation” are approximations; real immune networks are context- and cell-type–specific.
- **Redundancy:** Top SPIA hits may overlap messages already seen in ORA/GSEA/GSVA — agreement strengthens evidence; disagreement invites caution, not automatic dismissal.
- **Licensing / maintenance:** KEGG data are periodically updated; results can shift slightly between SPIA/database versions.

We run SPIA **anyway** so you can compare outputs side-by-side with first- and second-generation results and document the workflow.

## 15.1 Prepare DE vectors (Entrez) and run SPIA

SPIA expects a **named numeric vector of log2 fold changes for differentially expressed genes** (Entrez IDs as names) and the **universe** of all Entrez IDs tested (`rownames(dds)`).

```{r spia-run, cache=TRUE}
spia_from_res <- function(res_df, dds_obj) {
  sig <- res_df %>%
    dplyr::filter(!is.na(padj), padj < 0.05, !is.na(log2FoldChange)) %>%
    dplyr::mutate(GeneID = as.character(GeneID))
  if (nrow(sig) < 5)
    return(NULL)
  # one value per Entrez: keep strongest |LFC| if duplicates exist
  sig <- sig %>%
    dplyr::group_by(GeneID) %>%
    dplyr::slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
    dplyr::ungroup()
  de <- sig$log2FoldChange
  names(de) <- sig$GeneID
  all_ids <- as.character(rownames(dds_obj))
  spia(de, all_ids, organism = "hsa", nB = 2000, verbose = FALSE)
}

spia_v1 <- spia_from_res(res_list$V1, dds)
spia_v2 <- spia_from_res(res_list$V2, dds)
spia_v5 <- spia_from_res(res_list$V5, dds)
```

## 15.2 Top KEGG pathways (global pG)

```{r spia-tables}
fmt_spia <- function(x, n = 12) {
  if (is.null(x) || nrow(x) == 0)
    return(invisible(NULL))
  tibble::rownames_to_column(as.data.frame(x), "pathway_id") %>%
    dplyr::arrange(pG) %>%
    dplyr::select(Name, ID, Status, pNDE, pPERT, pG, pGFdr) %>%
    head(n)
}

fmt_spia(spia_v1)
fmt_spia(spia_v2)
fmt_spia(spia_v5)
```

## 15.3 Visualize top hits (V1)

```{r spia-plot-v1, fig.height=6}
if (!is.null(spia_v1) && nrow(spia_v1) > 0) {
  top <- tibble::rownames_to_column(as.data.frame(spia_v1), "pathway_id") %>%
    dplyr::arrange(pG) %>%
    head(15)
  ggplot(top, aes(x = reorder(Name, -log10(pG)), y = -log10(pG), fill = Status)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c(Activated = "firebrick", Inhibited = "steelblue",
                                 Mixed = "gray50"),
                      na.value = "gray70") +
    labs(title = "SPIA: top KEGG pathways (V1 vs Control)",
         subtitle = "Bar height = -log10(global pG). Status from SPIA.",
         x = NULL, y = "-log10(pG)") +
    theme(legend.position = "bottom")
}
```

Pathways with large **|tA|** (total net accumulation) and small **pPERT** are those where **topology** contributes strongly, not just list overlap.

---
---

# 16. Save All Results

```{r save-all-results}
for (tp in names(ora_hallmark_all)) {
  df <- as.data.frame(ora_hallmark_all[[tp]])
  if (nrow(df) > 0) write.csv(df, file.path(results_dir,
    paste0("ORA_Hallmark_", tp, ".csv")), row.names = FALSE)
}
for (tp in names(gsea_results)) {
  df <- as.data.frame(gsea_results[[tp]])
  if (nrow(df) > 0) write.csv(df, file.path(results_dir,
    paste0("GSEA_Hallmark_", tp, ".csv")), row.names = FALSE)
}
for (nm in names(gsva_results_list)) {
  write.csv(gsva_results_list[[nm]], file.path(results_dir,
    paste0("GSVA_Hallmark_", nm, ".csv")), row.names = FALSE)
}
write.csv(gsva_scores, file.path(results_dir, "GSVA_scores_per_sample.csv"))

if (!is.null(spia_v1) && nrow(spia_v1) > 0)
  write.csv(as.data.frame(spia_v1), file.path(results_dir, "SPIA_KEGG_V1_vs_Control.csv"),
            row.names = TRUE)
if (!is.null(spia_v2) && nrow(spia_v2) > 0)
  write.csv(as.data.frame(spia_v2), file.path(results_dir, "SPIA_KEGG_V2_vs_Control.csv"),
            row.names = TRUE)
if (!is.null(spia_v5) && nrow(spia_v5) > 0)
  write.csv(as.data.frame(spia_v5), file.path(results_dir, "SPIA_KEGG_V5_vs_Control.csv"),
            row.names = TRUE)

cat("All results saved to:", results_dir, "\n")
```

---
---

# 17. Results Interpretation & Discussion

This section brings together all findings into a coherent narrative.

## 17.1 What did we find? (The Big Picture)

```{r interpretation-summary}
cat("================================================================\n","            COMPLETE ANALYSIS SUMMARY\n","================================================================\n\n")

cat("DATA\n","  Genome reference: GRCh38.p13 (upgraded from hg19)\n"
,"  Aligner: Kallisto (upgraded from TopHat2/Cufflinks2)\n",
"  Samples analysed:", ncol(counts_filtered), "\n",
"  Genes after filtering:", nrow(counts_filtered), "\n\n",
"DIFFERENTIAL EXPRESSION (DESeq2, |lfc|>1, padj<0.05)\n")
print(deg_summary)

cat("\nOriginal study (Limma/Voom on FPKM, |FC|>1.5, FDR<0.1%):\n",
"  V1: 1,235 DEGs | V2: 1,060 DEGs | V5: 686 DEGs\n\n",
,"PATHWAY ANALYSIS (ORA + GSEA/GSVA + SPIA)\n",
"  1st gen ORA  (Hallmark, V1):", nrow(as.data.frame(ora_hallmark_v1)),
    "enriched pathways\n",
"  1st gen ORA  (GO:BP, V1):  ",
    sum(as.data.frame(ora_gobp_v1)$p.adjust < 0.05), "enriched terms\n")
cat("  2nd gen GSEA (Hallmark, V1):", sum(gsea_h_df$p.adjust < 0.05),
    "significant pathways\n",
"  2nd gen GSVA (Hallmark, V1):",
    sum(gsva_results_list$V1_vs_Ctrl$adj.P.Val < 0.05),
    "differentially active pathways\n")
if (!is.null(spia_v1) && nrow(spia_v1) > 0) {
  cat("  3rd gen SPIA (KEGG, V1):    ",
      sum(spia_v1$pGFdr < 0.05, na.rm = TRUE),
      "pathways with pGFdr < 0.05\n")
} else {
  cat("  3rd gen SPIA (KEGG, V1):    not computed or empty\n")
}
```

## 17.2 Biological Interpretation

### The Acute Phase (V1): A Strong Inflammatory Storm

At V1 (acute Lyme diagnosis, before antibiotics), we detected the
largest number of DEGs and the most enriched pathways. The top pathways
across MSigDB-based ORA/GSEA/GSVA consistently point to a massive immune response:

- **TNF-alpha signaling via NF-kB**: TNF-alpha is a master
  pro-inflammatory cytokine. NF-kB is the transcription factor it
  activates. Together, they drive the production of other inflammatory
  molecules. The original paper also identified TNF-alpha as a top
  upstream regulator.
  
- **Interferon-gamma response**: IFN-gamma is produced primarily by
  T-cells and NK cells. It activates macrophages and drives Th1
  immunity -- the type needed to fight intracellular and spirochetal
  pathogens like *B. burgdorferi*. This matches the original paper's
  finding of TREM1-mediated Th1 activation.
  
- **Inflammatory response**: A broad category encompassing cytokine
  signaling, leukocyte recruitment, and tissue inflammation. This is
  the expected hallmark of an acute bacterial infection.
  
- **Complement**: The complement system is an ancient innate immune
  defence that directly lyses bacteria and flags them for phagocytosis.
  Its activation in acute Lyme disease is biologically logical.

- **IL6-JAK-STAT3 signaling**: IL-6 is a pleiotropic cytokine
  central to acute-phase responses. The original paper identified IL-6
  as a top upstream regulator at V1 and V2.

**SPIA (KEGG, Section 15)** adds a separate view: it flags pathways where
both **list enrichment** and **signed graph perturbation** suggest
**Activated** or **Inhibited** status (e.g. chemokine signalling, lysosome,
cell-cycle-related KEGG maps in our run). Treat these as complementary to
Hallmark/GO — not a one-to-one match to immune gene-set names.

### Post-Treatment (V2): The Signature Persists

At V2 (3 weeks later, after completing antibiotics), a substantial
number of DEGs and pathways remain active. This is the paper's central
finding: **the gene expression signature does not promptly return to
baseline after treatment**, unlike what is typically observed in other
acute infections.

From our NES heatmap (Section 12.4) and GSVA boxplots (Section 14.7),
many inflammatory pathways remain elevated at V2, albeit often with
reduced intensity compared to V1. This suggests that while antibiotics
eliminate the bacterium, the immune system remains in a heightened
state for weeks afterwards.

### Six Months Later (V5): Partial but Incomplete Recovery

By V5 (6 months post-treatment), the number of DEGs drops
substantially but does not reach zero. Some pathways still show
residual activity. The GSVA per-sample boxplots (Section 14.7) are
particularly informative here: while most V5 patients approach control
levels, some individuals remain elevated. These outliers may correspond
to patients who developed persistent symptoms.

The original paper found 686 DEGs at V5 and noted 31-60% of pathways
in common with chronic autoimmune conditions (SLE, RA). Our results
are consistent with a lingering transcriptomic perturbation at 6
months.

## 17.3 How Our Re-Analysis Compares to the Original Study

| Aspect | Bouquet et al. 2016 | Our Re-Analysis |
|--------|---------------------|-----------------|
| Genome | hg19 (2009) | GRCh38.p13 (current) |
| Aligner | TopHat2/Cufflinks2 (deprecated) | Kallisto (modern) |
| Input data | FPKM (normalised) | Raw integer counts |
| DE method | Limma/Voom | DESeq2 (negative binomial) |
| Fold-change | Raw FC > 1.5 | Shrunken LFC (ashr) |
| Pathway tool | IPA (proprietary, 3$) | clusterProfiler + MSigDB (open-source, free) |
| Pathway methods | ORA only (1st gen) | ORA + GSEA + GSVA + SPIA (1st–3rd gen) |
| Per-sample resolution | No (bulk group comparison) | Yes (GSVA scores per patient) |

### What we confirmed

- A strong immune/inflammatory signature at V1 (acute Lyme)
- The signature persists beyond antibiotic treatment (V2)
- Partial recovery by 6 months (V5) but not to baseline
- TNF-alpha, interferons, complement, and IL-6 signaling are central

### What we added

- **GSEA** detects subtle coordinated pathway changes without
  arbitrary cutoffs, finding more significant pathways than ORA alone.
- **GSVA** provides individual patient pathway scores, allowing us to
  see heterogeneity within each time point -- particularly relevant for
  understanding why some V5 patients still have elevated inflammatory
  pathways while others recover fully.
- All tools are open-source and reproducible (vs. IPA).
- **SPIA (3rd gen)** highlights KEGG pathways where both **gene list enrichment** and **signed topology** suggest activation or inhibition — a different lens from MSigDB Hallmark.

## 17.4 Answering Open Questions from the Analysis

**Section 5.2 -- "Is the mean set correctly for the mean-variance plot?"**
Yes. The plot correctly shows that for raw RNA-Seq counts, the variance
consistently exceeds the mean (overdispersion). The red Poisson line
(variance = mean) is a theoretical baseline. The fact that points sit
above it is not an error; it is the fundamental property of count data
that justifies using the negative binomial model in DESeq2.

**Section 7.4 -- "VST doesn't seem to flatten the mean-SD relationship?"**
The VST does reduce the dependency, even if visually the "before" and
"after" plots look similar. The key metric is the correlation between
mean and SD: even a modest reduction (e.g., from 0.8 to 0.5) meaningfully
improves downstream analyses like PCA. The fact that the correlation
doesn't drop to zero is normal -- VST is an approximation, not magic.
For small sample sizes, `rlog()` (regularised log) can perform better
than VST but is much slower. Both are acceptable for this project.

**MA plots (Sections 5.3, 7.3, 10.1a) — do they validate normalization?**  
Section 5.3 uses raw log2 counts: systematic offset is expected. Section 7.3 uses **size-factor–normalized** counts only; the loess trend should sit closer to **M = 0**. Section **10.1a** uses **DESeq2**’s `plotMA()` after dispersion shrinkage: the central mass of points at moderate-to-high expression should lie near **LFC = 0** for non-DEGs (blue = padj &lt; alpha). Together, these three plots answer “did we fix depth?” vs “what does the fitted model say about differential expression?”

**Section 12.2 -- "Which pathways should we expect activated/suppressed?"**
Based on what we know about Lyme disease biology:

*Expected activated:* Inflammatory response, TNF-alpha signaling,
interferon signaling (IFN-gamma, IFN-alpha), complement, IL6-JAK-STAT,
TLR signaling, NF-kB signaling, apoptosis.

*Potentially suppressed:* The original paper found eIF2 signaling
(translation initiation) downregulated at all time points. In our GSEA
results, look for pathways with negative NES (blue in the heatmap).
Suppressed MYC targets or mTORC1 signaling would indicate reduced
cell growth/proliferation, consistent with immune cells diverting
resources from growth to defence.

## 17.5 Book Chapter Reference Map

| Our Section | Thomas's Book Chapter | Method |
|-------------|-----------------------|--------|
| 4.1 Library sizes | Ch. 7.2 | Quality Control |
| 4.3 RLE plot | Ch. 7.2.4 | Relative Log Expression |
| 5.3 MA plot | Ch. 7.2.7 | Mean-difference plot |
| 7. DESeq2 | Ch. 7.6 | Differential expression |
| 8. PCA | Ch. 7.2.2 | Dimensionality reduction |
| 9. Distance heatmap | Ch. 7.2.1 | Sample distance |
| 11. ORA | Ch. 8.1.1, 8.4.1 | 1st generation pathway analysis |
| 12. GSEA | Ch. 8.1.2.1, 8.4.3 | 2nd generation pathway analysis |
| 13. compareCluster | Ch. 8.4.2 | Biologic theme comparison |
| 14. GSVA | Ch. 8.1.2.2, 8.4.5 | 2nd generation (per-sample scores) |
| 15. SPIA | Ch. 8 (topology); Khatri et al. 2012 | 3rd generation (KEGG perturbation) |
| MSigDB usage | Ch. 8.2 | Molecular Signature Database |
| 7.3–7.4 / 10.1a MA | Ch. 7.2.7 | Mean–difference / DESeq2 MA |

---

# Session Info

```{r session-info}
sessionInfo()
```
