---
title: "Lyme Disease Differential Gene Expression Re-Analysis"
subtitle: "Re-analysis of Bouquet et al. 2016 (GSE63085) using DESeq2 & clusterProfiler"
author: "Abed & Laszlo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center"
)
```

# 1. Environment Setup

## 1.1 Install Packages (run once, then set eval=FALSE), or use "using" function

```{r install-packages, eval=FALSE, include=FALSE}
# install.packages(c("tidyverse", "pheatmap", "RColorBrewer", "ggrepel",
#                     "gridExtra", "viridis", "ashr"))
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(c("DESeq2", "clusterProfiler", "msigdbr",
#                         "enrichplot", "org.Hs.eg.db", "GSVA"))
```

```{r load-libraries}
# if (rstudioapi::isAvailable()) {setwd(dirname(rstudioapi::getActiveDocumentContext()$path))}
options(width = 200)

# use this function to install and load packages
using <- function(pkgs) {
  for (pkg in pkgs) {  # Loop through each package in the provided vector
    if (!requireNamespace(pkg, quietly = TRUE)) {    # If not installed, install it
      install.packages(pkg, repos = "https://cloud.r-project.org")}
    # Load the package:
    library(pkg, character.only = TRUE)}}

using(c("tidyverse", "pheatmap", "RColorBrewer", "ggrepel", "corrplot", "gridExtra", "viridis", "BiocManager","DESeq2", "org.Hs.eg.db", "clusterProfiler", "EnhancedVolcano","msigdbr", "PCAtools", "GSVA", "enrichplot", "ashr", "matrixStats"))

# description of libraries:
# tidyverse       - data wrangling + ggplot2
# DESeq2          - differential expression (negative binomial model)
# clusterProfiler - pathway enrichment (ORA, GSEA)
# msigdbr         - access to MSigDB gene sets
# enrichplot      - enrichment result visualizations
# pheatmap        - heatmaps with clustering
# RColorBrewer    - colour palettes
# ggrepel         - non-overlapping text labels on plots
# gridExtra       - arrange multiple plots
# viridis         - colour-blind-friendly palettes
# matrixStats     - fast row/column statistics

theme_set(theme_bw(base_size = 13)) # this is for the plots 
```


We had to use explicit `dplyr::` prefixes on key functions (`filter`, `select`, `count`, `mutate`, `arrange`) throughout this document because
`clusterProfiler` loads packages that mask some of these names.


## 1.3 Directories

```{r directories}
project_dir <- getwd()
data_dir    <- file.path(project_dir, "Data")
results_dir <- file.path(project_dir, "results")
plots_dir   <- file.path(project_dir, "plots")

for (d in c(results_dir, plots_dir))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
```

---
---

# 2. Data Loading

The raw FASTQ reads from GEO (GSE63085) were re-aligned by Thomas to the modern **GRCh38.p13** human reference genome (replacing
the outdated hg19 used in the original 2016 study) and quantified into raw integer counts, the proper input for DESeq2. Let's not forget to ask Thomas which aligner tool he used (STAR + featureCounts? HISAT2 + HTSeq?). And more importantly, why are there 22 missing samples were excluded during re-alignment? or are they simply not processed at all?

**References:**

- Bouquet J et al. *mBio* 2016; 7(1):e00100-16.
- Thomas's book T. *Beginner's Handbook of -Omics Data Analysis*, Ch. 7-8.

## 2.1 Raw Counts Matrix

```{r load-counts}
counts_raw <- read.delim(
  gzfile(file.path(data_dir, "GSE63085_raw_counts_GRCh38.p13_NCBI.tsv.gz")),
  row.names = 1, check.names = FALSE
)
cat("Counts matrix:", nrow(counts_raw), "genes x", ncol(counts_raw), "samples\n")
```

## 2.2 Gene Annotation (featureData)

This maps NCBI Entrez Gene IDs to gene symbols, biotypes, chromosomal locations, and GO terms.

NCBI Entrez as the "Google Search" engine for biological data. It’s a massive integrated system managed by the National Center for Biotechnology Information (NCBI) that links together different databases—DNA sequences, protein structures, and medical literature (PubMed).
According to LLMs, when we see an Entrez ID (like those numbers in your matrix), it’s a unique "social security number" for a gene. Names (Symbols) change as scientists argue, but Entrez IDs stay the same.

```{r load-gene-annot}
gene_annot <- read.delim(
  gzfile(file.path(data_dir, "Human.GRCh38.p13.annot.tsv.gz")),
  stringsAsFactors = FALSE
)
cat("Gene annotation:", nrow(gene_annot), "x", ncol(gene_annot), "\n")
```
> Explanation of the columns (source: LLMS):

* **GeneID**: This matches the row names in your counts matrix. It’s your primary key.

* **Symbol**: The "human-friendly" name (e.g., TP53, BRCA1). This is what you’ll put in your final paper or report.

* **Description**: A long-form name of what the gene actually does (e.g., "tumor protein p53").

* **EnsemblGeneID**: Another ID system (starting with ENSG...). Bioinformaticians often use both Entrez and Ensembl because some tools prefer one over the other.

* **GeneType**: Usually, you only want to analyze protein-coding genes. Most researchers filter out the pseudo (dead "fossil" genes) and ncRNA (non-coding) unless they are specifically looking for them.

* **Status**: You should likely filter for "active". A "discontinued" status means the gene was merged into another record or found to be a mistake in the genome mapping.

* **ChrAcc, ChrStart, ChrStop, Orientation**: These are the "Location" columns. They tell you exactly where the gene sits on which chromosome and which way it's facing. This is vital if you want to see if a treatment affects a whole neighborhood of genes (a "cluster") or just one.

* **GOFunction / GOProcess / GOComponent**: "GO" stands for Gene Ontology. These tell you the "job" of the gene. **Process**: What is the "big picture" goal? (e.g., "Cell Division"); **Function**: What is the specific biochemical task? (e.g., "Binding to DNA"); and **Component**: Where does it happen? (e.g., "Mitochondria").


```{r meta_EDA}
names(gene_annot)
cat("\n\nGene biotypes:"); print(table(gene_annot$GeneType))

cat("\n\nStatus unique:");print(table(gene_annot$Status))
```
> We have ~39k total entries, but only 19,416 are protein-coding. 

## 2.3 Sample Annotation

```{r load-sample-annot}
sample_annot <- read.csv(
  file.path(project_dir, "sample_annotations.csv"),
  stringsAsFactors = FALSE
)
```

## 2.4 Align Samples

22 samples from the annotation are absent from the counts matrix.
The reason is not confirmed -- ask Thomas whether they were excluded
during re-alignment or simply not processed.

```{r align-data}
shared     <- intersect(colnames(counts_raw), sample_annot$GSM)
only_annot <- setdiff(sample_annot$GSM, colnames(counts_raw))

cat("Shared:", length(shared), "| Not in counts:", length(only_annot), "\n")

counts <- counts_raw[, shared]
annot  <- sample_annot %>%
  dplyr::filter(GSM %in% shared) %>%
  dplyr::arrange(match(GSM, shared))

stopifnot(all(colnames(counts) == annot$GSM))
```

## 2.5 Engineer Metadata Columns

```{r engineer-metadata}
annot <- annot %>%
  dplyr::mutate(
    patient_id = gsub("^.*-([0-9]+)_V.*$", "\\1",
                      gsub("^.*-([0-9]+)_V$", "\\1", Observation)),
    visit = dplyr::case_when(
      grepl("V1", Observation) ~ "V1",
      grepl("V2", Observation) ~ "V2",
      grepl("V5", Observation) ~ "V5",
      TRUE                     ~ "Control"),
    group = dplyr::case_when(
      disease.state == "Healthy controls" ~ "Control",
      visit == "V1" ~ "Lyme_V1",
      visit == "V2" ~ "Lyme_V2",
      visit == "V5" ~ "Lyme_V5"),
    condition = ifelse(disease.state == "Healthy controls", "Control", "Lyme"),
    group     = factor(group,     levels = c("Control","Lyme_V1","Lyme_V2","Lyme_V5")),
    visit     = factor(visit,     levels = c("Control","V1","V2","V5")),
    condition = factor(condition, levels = c("Control","Lyme"))
  )
rownames(annot) <- annot$GSM

cat("Samples per group:\n")
print(table(annot$group))
```

---
---

# 3. Gene Biotype Overview

```{r gene-biotypes}
gene_annot_matched <- gene_annot %>%
  dplyr::filter(GeneID %in% as.numeric(rownames(counts)))

biotype_counts <- gene_annot_matched %>%
  dplyr::count(GeneType, sort = TRUE) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))

ggplot(biotype_counts %>% dplyr::filter(n > 50),
       aes(x = reorder(GeneType, n), y = n, fill = GeneType)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(n, " (", pct, "%)")), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Gene Biotypes in Counts Matrix (GRCh38.p13)",
       x = NULL, y = "Number of Genes") +
  scale_fill_viridis_d()
```

About half the genes are **protein-coding** (~49%).
The rest are non-coding RNA, pseudogenes, etc. According to LLMs, protein-coding genes carry the strongest biological signal for differential expression. \n 
After filtering (Section 6), the protein-coding proportion will increase because non-coding genes tend to have very low counts.

---

# 4. Sample-Level Quality Control

*Ref: Thomas's book Ch. 7.2*

## 4.1 Library Sizes

Library size = total mapped reads per sample. Large variation means
different sequencing depths, which must be normalised.

```{r library-sizes, fig.height=10}
lib_sizes <- data.frame(
  GSM = colnames(counts), total = colSums(counts)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group, Observation), by = "GSM")

ggplot(lib_sizes, aes(x = reorder(Observation, total),
                       y = total / 1e6, fill = group)) +
  geom_col() + coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library Size per Sample",
       x = NULL, y = "Total Counts (millions)", fill = "Group") +
  theme(axis.text.y = element_text(size = 5))
```

```{r lib-size-boxplot}
ggplot(lib_sizes, aes(x = group, y = total / 1e6, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library Size by Group", x = NULL, y = "Total Counts (millions)") +
  theme(legend.position = "none")
```

**Interpretation:** If any group has systematically lower library sizes,
the DESeq2 size-factor normalisation (Section 7) will correct for this.
Look for outlier samples that are dramatically different from the rest --
these might warrant investigation.

## 4.2 Detected Genes per Sample

```{r detected-genes}
detected <- data.frame(
  GSM = colnames(counts), n_detected = colSums(counts > 0)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(detected, aes(x = group, y = n_detected, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Detected Genes per Sample (count > 0)",
       x = NULL, y = "Genes Detected") +
  theme(legend.position = "none")
```

## 4.3 Relative Log Expression (RLE)

*Ref: Thomas's book Ch. 7.2.4*

RLE measures how much each gene in a sample deviates from its median
expression across all samples. Well-behaved samples centre around zero.
Samples with shifted or wide RLE distributions may indicate technical
problems or batch effects.

```{r rle-plot, fig.height=8}
# Compute RLE: for each gene, subtract the row median
log2_counts <- log2(counts + 1)
row_medians <- rowMedians(as.matrix(log2_counts))
rle_matrix  <- sweep(log2_counts, 1, row_medians)

rle_long <- rle_matrix %>%
  as.data.frame() %>%
  tidyr::pivot_longer(everything(), names_to = "GSM", values_to = "RLE") %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(rle_long, aes(x = GSM, y = RLE, fill = group)) +
  geom_boxplot(outlier.size = 0.1) +
  scale_fill_brewer(palette = "Set2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Relative Log Expression (RLE) Plot",
       subtitle = "Well-behaved samples centre around the red dashed line (zero)",
       x = NULL, y = "Deviation from median expression") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 3)) +
  coord_cartesian(ylim = c(-2, 2))
```

**Bingo:** Samples with medians shifted away from zero or with
unusually wide boxes may have quality issues. If all groups looked similar,
there wouldn't be strong batch effect in the raw data. But apparently, they don't, so there could be, supporting our hypothesis that there could be batch effects.

---
---

# 5. Expression Distributions

## 5.1 Per-Sample Density

```{r density-plots}
log2_long <- log2_counts %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  tidyr::pivot_longer(-gene, names_to = "GSM", values_to = "log2_count") %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(log2_long, aes(x = log2_count, group = GSM, color = group)) +
  geom_density(linewidth = 0.3) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Per-Sample Density of log2(count + 1)",
       x = "log2(count + 1)", y = "Density", color = "Group") +
  xlim(0, 20)
```

**Interpretation:**  the large spike near zero represents the many genes
with zero or very low expression. Most of them are still there because we haven't filtered out the garbage yet. The second hump (should be around 5-10) represents actively expressed genes. What we are looking for looks flat because the scale is zoomed out to accommodate the zero spike. All samples should have a similar shape; if
one looks radically different, it may be a quality outlier. Perhaps, we should make a similar plot after filtering.

## 5.2 Mean-Variance Relationship

```{r mean-variance}
gene_stats <- data.frame(
  mean_count = rowMeans(counts), var_count = apply(counts, 1, var)
)

ggplot(gene_stats, aes(x = log10(mean_count + 1), y = log10(var_count + 1))) +
  geom_point(alpha = 0.05, size = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 0.8) +
  labs(title = "Mean-Variance Relationship (Raw Counts)",
       subtitle = "Red dashed = Poisson (variance = mean). Points above = overdispersion.",
       x = "log10(Mean + 1)", y = "log10(Variance + 1)")
```

Nearly all genes sit above the red line, meaning variance > mean **if my mean is set correctly**. According to AI: this overdispersion is exactly why DESeq2 uses a negative binomial model instead of a simpler Poisson model. A Poisson
model would dramatically underestimate variability and produce too many false positives.

## 5.3 MA Plot (pre-normalisation)

*Ref: Thomas's book Ch. 7.2.7*

An MA plot shows each gene's **average expression** (A) vs. its **log
fold change** (M) between two conditions. Before normalisation, the cloud
should centre around M=0 (most genes are not differentially expressed).
A shift or banana shape indicates a normalisation problem.

```{r ma-plot-pre-norm, fig.height=5}
# Quick MA: V1 (first patient) vs first Control
ctrl_cols  <- annot$GSM[annot$group == "Control"]
lyme_cols  <- annot$GSM[annot$group == "Lyme_V1"]

A <- 0.5 * (rowMeans(log2_counts[, ctrl_cols]) + rowMeans(log2_counts[, lyme_cols]))
M <- rowMeans(log2_counts[, lyme_cols]) - rowMeans(log2_counts[, ctrl_cols])

ma_df <- data.frame(A = A, M = M)

ggplot(ma_df, aes(x = A, y = M)) +
  geom_point(alpha = 0.05, size = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "orange", se = FALSE, linewidth = 0.8) +
  labs(title = "MA Plot: Lyme V1 vs Control (pre-normalisation)",
       subtitle = "Red = zero line. Orange = loess trend. Centred cloud = good.",
       x = "A (average expression)", y = "M (log2 fold change)") +
  coord_cartesian(ylim = c(-5, 5))
```

Cloud isn't centered around 0, indicating a scaling issue. 

---
---

# 6. Gene Filtering

We remove genes with very low counts to reduce noise. We require at
least N samples (= smallest group size) to have a count >= 10.

```{r prefilter}
smallest_group <- min(table(annot$group))
keep <- rowSums(counts >= 10) >= smallest_group
counts_filtered <- counts[keep, ]

cat("Before:", nrow(counts), "genes\n")
cat("After: ", nrow(counts_filtered), "genes\n")
cat("Removed:", nrow(counts) - nrow(counts_filtered),
    "(", round(100 * (1 - nrow(counts_filtered)/nrow(counts)), 1), "%)\n")
```

```{r post-filter-biotypes}
biotype_after <- gene_annot %>%
  dplyr::filter(GeneID %in% as.numeric(rownames(counts_filtered))) %>%
  dplyr::count(GeneType, sort = TRUE) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))
print(biotype_after)
```

**Meaning:** After aggressive filtering, ~70% of remaining genes are **protein-coding** (up from ~49%), confirming that most low-count genes were non-coding RNA and pseudogenes. This is expected and desirable.

---
---

# 7. DESeq2 Normalisation

*Ref: Thomas's book Ch. 7.6e*

## 7.1 Create DESeqDataSet

The `design = ~ group` formula tells DESeq2 to model expression as a function of our four experimental groups.

```{r create-dds}
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered, colData = annot, design = ~ group)
```



## 7.2 Size Factors

DESeq2 estimates a **size factor** per sample via the median-of-ratios method. This corrects for sequencing depth differences and is far superior to the FPKM normalisation used in the original study.

```{r size-factors}
dds <- estimateSizeFactors(dds)

sf_df <- data.frame(
  GSM = colnames(dds), size_factor = sizeFactors(dds),
  lib_size = colSums(counts_filtered)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(sf_df, aes(x = lib_size / 1e6, y = size_factor, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "gray40", linetype = "dashed") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "DESeq2 Size Factor vs Library Size",
       x = "Library Size (millions)", y = "Size Factor")
```

Size factor and library size are strongly correlated (expected). 
Samples with a size factor < 1 were sequenced less deeply; > 1 means more deeply. 
DESeq2 divides each sample's counts by its size factor during analysis.

## 7.3 Variance Stabilising Transformation (VST)

Raw counts have a mean-variance dependency (Section 5.2). VST produces
values on a log2-like scale with approximately **constant variance**,
suitable for PCA and heatmaps.

```{r vst}
vsd <- vst(dds, blind = FALSE)
```

```{r vst-effect, fig.height=5}
par(mfrow = c(1, 2))
plot(rowMeans(log2(counts_filtered + 1)),
     apply(log2(counts_filtered + 1), 1, sd),
     pch = 16, cex = 0.3, col = rgb(0.2, 0.4, 0.8, 0.1),
     xlab = "Mean", ylab = "SD", main = "Before: log2(counts+1)")
plot(rowMeans(assay(vsd)), apply(assay(vsd), 1, sd),
     pch = 16, cex = 0.3, col = rgb(0.8, 0.2, 0.2, 0.1),
     xlab = "Mean", ylab = "SD", main = "After: VST")
par(mfrow = c(1, 1))
```



Before VST (left), low-expression genes have variable SD. After VST (right), the trend should be flattened, but it's not, so the SD still depends on the mean. This makes the meaningfullness of distances between samples (for PCA, clustering) questionable.



The following can verify whether before and after actually look too similar

```{r}
mean_sd_before <- cor(
  rowMeans(log2(counts_filtered + 1)),
  apply(log2(counts_filtered + 1), 1, sd)
)

mean_sd_after <- cor(
  rowMeans(assay(vsd)),
  apply(assay(vsd), 1, sd)
)

mean_sd_before
mean_sd_after
```

Confirm with Thomas!


---
---

# 8. PCA

*Ref: Thomas's book Ch. 7.2.2*

```{r pca-compute}
rv      <- rowVars(assay(vsd))
top_idx <- order(rv, decreasing = TRUE)[1:min(2000, length(rv))]

pca     <- prcomp(t(assay(vsd)[top_idx, ]), center = TRUE, scale. = TRUE)
var_pct <- round(100 * summary(pca)$importance[2, ], 1)

pca_df  <- data.frame(pca$x[, 1:5]) %>%
  tibble::rownames_to_column("GSM") %>%
  dplyr::left_join(annot, by = "GSM")
```

## 8.1 Scree Plot

```{r scree}
ggplot(data.frame(PC = 1:20, Var = var_pct[1:20]),
       aes(x = PC, y = Var)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_line(color = "darkred") + geom_point(color = "darkred", size = 2) +
  labs(title = "Scree Plot", x = "Principal Component",
       y = "% Variance Explained") +
  scale_x_continuous(breaks = 1:20)
```

## 8.2 PCA by Group

```{r pca-group, fig.width=15, fig.height=5}
make_pca <- function(df, pcx, pcy, vpct) {
  ggplot(df, aes(x = .data[[pcx]], y = .data[[pcy]], color = group)) +
    geom_point(size = 3, alpha = 0.8) +
    stat_ellipse(level = 0.68, linetype = "dashed") +
    scale_color_brewer(palette = "Set2") +
    labs(x = paste0(pcx, " (", vpct[as.integer(gsub("PC","",pcx))], "%)"),
         y = paste0(pcy, " (", vpct[as.integer(gsub("PC","",pcy))], "%)"))
}

grid.arrange(
  make_pca(pca_df,"PC1","PC2",var_pct) + ggtitle("PC1 vs PC2"),
  make_pca(pca_df,"PC1","PC3",var_pct) + ggtitle("PC1 vs PC3"),
  make_pca(pca_df,"PC2","PC3",var_pct) + ggtitle("PC2 vs PC3"),
  ncol = 3
)
```

**According to AI:** Look for how well the four groups separate.
**PC1** (largest variance axis) typically separates disease from
control. If V1 (acute Lyme) clusters distinctly from controls, it
confirms that infection drives a strong transcriptomic shift. V5 samples
sitting closer to controls suggests partial recovery of gene expression
by 6 months -- consistent with the original paper's finding that the
signature *persists* but *weakens*.

## 8.3 Patient Trajectories

Lines connect the same patient across V1 -> V2 -> V5.

```{r pca-trajectories, fig.height=7}
lyme_pca <- pca_df %>% dplyr::filter(condition == "Lyme") %>%
  dplyr::arrange(patient_id, visit)

ggplot() +
  geom_point(data = pca_df %>% dplyr::filter(condition == "Control"),
             aes(x = PC1, y = PC2), color = "gray60", size = 3, alpha = 0.5) +
  geom_path(data = lyme_pca,
            aes(x = PC1, y = PC2, group = patient_id),
            color = "gray40", alpha = 0.3) +
  geom_point(data = lyme_pca,
             aes(x = PC1, y = PC2, color = visit, shape = visit), size = 3) +
  scale_color_manual(values = c(V1="#d62728", V2="#ff7f0e", V5="#1f77b4")) +
  labs(title = "Patient Trajectories Through PCA Space",
       subtitle = "Gray = controls. Lines connect same patient across visits.",
       x = paste0("PC1 (", var_pct[1], "%)"),
       y = paste0("PC2 (", var_pct[2], "%)"))
```

Each line represents one patient's journey from acute infection (red V1) through treatment (orange V2) to 6-month
follow-up (blue V5). Glücklicherweise, V5 points move toward the gray control cloud, showing recovery. Patients whose V5 stays far from controls are fucked because it's probably those with persistent symptoms (PTLDS).

---
---

# 9. Sample Relationships

*Ref: Thomas's book Ch. 7.2.1*

## 9.1 Distance Heatmap

```{r distance-heatmap, fig.height=9, fig.width=10}
sample_dists <- dist(t(assay(vsd)))
row_annot <- data.frame(Group = annot$group, row.names = annot$GSM)
ann_colors <- list(Group = c(Control="#66C2A5", Lyme_V1="#FC8D62",
                              Lyme_V2="#8DA0CB", Lyme_V5="#E78AC3"))

pheatmap(as.matrix(sample_dists),
         clustering_distance_rows = sample_dists,
         clustering_distance_cols = sample_dists,
         annotation_row = row_annot, annotation_col = row_annot,
         annotation_colors = ann_colors,
         show_rownames = FALSE, show_colnames = FALSE,
         color = viridis(100),
         main = "Sample-to-Sample Euclidean Distance (VST)")
```

## 9.2 Top 50 Most Variable Genes

```{r top50-heatmap, fig.height=12, fig.width=10}
top50_idx <- head(order(rv, decreasing = TRUE), 50)
top50_ids <- rownames(assay(vsd))[top50_idx]
top50_sym <- gene_annot$Symbol[match(as.numeric(top50_ids), gene_annot$GeneID)]
top50_lab <- ifelse(is.na(top50_sym) | top50_sym == "", top50_ids, top50_sym)

hm_mat <- assay(vsd)[top50_idx, ]
rownames(hm_mat) <- top50_lab

pheatmap(t(scale(t(hm_mat))),
         annotation_col = row_annot, annotation_colors = ann_colors,
         show_colnames = FALSE, fontsize_row = 7,
         color = colorRampPalette(c("navy","white","firebrick3"))(100),
         main = "Top 50 Most Variable Genes (Z-scored VST)",
         cutree_cols = 4)
```

Rows are genes, columns are samples. Red = high expression, blue = low expression (relative to the gene's mean). If the
heatmap clusters samples by group (check the colour bar at top), it confirms that the dominant source of variation is biological (disease state / time point), not technical noise.

---
---

# 10. Differential Expression with DESeq2

*Ref: Thomas's book Ch. 7.6*

## 10.1 Run DESeq2

```{r run-deseq}
dds <- DESeq(dds)
resultsNames(dds)
```

## 10.2 Extract, Shrink & Annotate Results

`lfcShrink()` with `ashr` produces moderated log2 fold-change estimates
that are more reliable for genes with low counts. *(Stephens 2016)*

```{r extract-results}
contrasts <- list(
  V1 = c("group", "Lyme_V1", "Control"),
  V2 = c("group", "Lyme_V2", "Control"),
  V5 = c("group", "Lyme_V5", "Control")
)

get_annotated_results <- function(dds, contrast, gene_annot) {
  res <- results(dds, contrast = contrast, alpha = 0.05)
  res_shrunk <- lfcShrink(dds, contrast = contrast, type = "ashr", res = res)

  as.data.frame(res_shrunk) %>%
    tibble::rownames_to_column("GeneID") %>%
    dplyr::mutate(GeneID_num = as.numeric(GeneID)) %>%
    dplyr::left_join(
      gene_annot %>% dplyr::select(GeneID, Symbol, Description, GeneType),
      by = c("GeneID_num" = "GeneID")) %>%
    dplyr::select(-GeneID_num) %>%
    dplyr::arrange(padj)
}

res_list <- lapply(contrasts, function(ct)
  get_annotated_results(dds, ct, gene_annot))
```

## 10.3 DEG Summary

```{r deg-summary}
deg_summary <- lapply(names(res_list), function(nm) {
  sig <- res_list[[nm]] %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1)
  data.frame(Comparison = paste0(nm, " vs Control"),
             Total = nrow(sig),
             Up = sum(sig$log2FoldChange > 0),
             Down = sum(sig$log2FoldChange < 0))
}) %>% dplyr::bind_rows()

print(deg_summary)
```

Compare with the original study (Limma/Voom on FPKM): V1=1,235, V2=1,060, V5=686 DEGs. 
Not a 100% sure why our numbers differ, but I assume it's because we use: 
(1) a different genome reference, 
(2) raw counts instead of FPKM,
(3) DESeq2 instead of Limma/Voom, and (4) different fold-change/FDR thresholds. \n 
The pattern should be similar: most DEGs at V1 (acute), declining through V2 and V5 as the transcriptome partially recovers.

## 10.4 Top DEGs at V1

```{r top-degs}
res_list$V1 %>%
  dplyr::filter(!is.na(padj) & padj < 0.05) %>%
  head(15) %>%
  dplyr::select(Symbol, log2FoldChange, padj, GeneType) %>%
  print()
```

## 10.5 Volcano Plots

```{r volcano, fig.width=16, fig.height=6}
make_volcano <- function(df, title) {
  df <- df %>%
    dplyr::mutate(sig = dplyr::case_when(
      is.na(padj) ~ "NS",
      padj < 0.05 & log2FoldChange > 1  ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"))
  top <- df %>% dplyr::filter(sig != "NS") %>% head(10)
  ggplot(df, aes(x = log2FoldChange, y = -log10(pvalue), color = sig)) +
    geom_point(alpha = 0.4, size = 1) +
    scale_color_manual(values = c(Up="firebrick", Down="steelblue", NS="gray70")) +
    geom_text_repel(data = top, aes(label = Symbol), size = 2.5,
                    max.overlaps = 15) +
    geom_vline(xintercept = c(-1,1), linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    labs(title = title, x = "log2 Fold Change", y = "-log10(p-value)") +
    theme(legend.position = "bottom")
}

grid.arrange(
  make_volcano(res_list$V1, "V1 (Acute) vs Control"),
  make_volcano(res_list$V2, "V2 (Post-Treatment) vs Control"),
  make_volcano(res_list$V5, "V5 (6 Months) vs Control"),
  ncol = 3
)
```

Each dot is a gene. The x-axis is the fold change (how much the gene changes in Lyme vs. Control); the y-axis is statistical significance. **Red dots** (upper-right) are significantly upregulated; **blue dots** (upper-left) are downregulated. \n The labeled genes are the most significant. We can see how the "cloud" narrows from V1 to V5, fewer genes are dramatically changed at 6 months, matching the original paper's finding of a gradually resolving but sustained signature.

## 10.6 DEG Overlap Across Time Points

```{r deg-overlap}
sig_genes <- lapply(res_list, function(df)
  df %>% dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1) %>%
    dplyr::pull(GeneID))

cat("V1:", length(sig_genes$V1), "  V2:", length(sig_genes$V2),
    "  V5:", length(sig_genes$V5), "\n")
cat("V1 & V2:", length(intersect(sig_genes$V1, sig_genes$V2)), "\n")
cat("V1 & V5:", length(intersect(sig_genes$V1, sig_genes$V5)), "\n")
cat("All 3:  ", length(Reduce(intersect, sig_genes)), "\n")
```

The overlap between V1 and V2 tells us how many genes remain differentially expressed even after antibiotic treatment.


## 10.7 Save DEG Tables

```{r save-degs}
for (nm in names(res_list))
  write.csv(res_list[[nm]],
            file.path(results_dir, paste0("DEG_", nm, "_vs_Control.csv")),
            row.names = FALSE)
```

---
---

# 11. 1st Generation: ORA (Over-Representation Analysis)

*Ref: Thomas's book Ch. 8.1.1, 8.4.1; Khatri et al. 2012*

ORA asks: "Are my DEGs enriched in known pathways more than expected
by chance?" It compares the fraction of DEGs in a pathway to the
background fraction (all tested genes).

## 11.1 Load MSigDB Gene Sets (targeted)

MSigDB is the companion database for GSEA (second gen. of pathway analysis). but it can also be used for ORA


I loaded only the collections we need to avoid memory issues. Otherwise I couldn't knit.
The full MSigDB is ~738 MB and crashes knitr's cache.

```{r load-msigdb, cache=FALSE}
msigdb_hallmark <- msigdbr(species = "Homo sapiens", collection = "H")
msigdb_gobp     <- msigdbr(species = "Homo sapiens", collection = "C5",
                            subcollection = "GO:BP")

cat("Hallmark:", nrow(msigdb_hallmark), "entries (",
    length(unique(msigdb_hallmark$gs_name)), "sets)\n")
cat("GO:BP:  ", nrow(msigdb_gobp), "entries (",
    length(unique(msigdb_gobp$gs_name)), "sets)\n")
```

```{r build-term-maps}
# Hallmark
t2g_hallmark <- msigdb_hallmark %>%
  dplyr::select(gs_name, gene_symbol)

# GO Biological Process
t2g_gobp <- msigdb_gobp %>%
  dplyr::select(gs_name, gene_symbol)
t2n_gobp <- msigdb_gobp %>%
  dplyr::select(gs_name, gs_description) %>%
  dplyr::distinct()
```

## 11.2 ORA: Hallmark (V1)

```{r ora-hallmark-v1}
sig_sym_v1 <- res_list$V1 %>%
  dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                !is.na(Symbol) & Symbol != "") %>%
  dplyr::pull(Symbol) %>% unique()

univ_sym <- res_list$V1 %>%
  dplyr::filter(!is.na(Symbol) & Symbol != "") %>%
  dplyr::pull(Symbol) %>% unique()

ora_hallmark_v1 <- enricher(
  gene = sig_sym_v1, universe = univ_sym,
  TERM2GENE = t2g_hallmark,
  pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 1
)
```

```{r ora-hallmark-dotplot, fig.height=6}
if (nrow(as.data.frame(ora_hallmark_v1)) > 0) {
  dotplot(ora_hallmark_v1, showCategory = 15) +
    ggtitle("ORA: Hallmark Pathways Enriched in V1 DEGs")
}
```

The Hallmark collection contains 50 well-defined biological processes from MSigDB (Thomas's book Ch. 8.2). Enriched pathways tell us *which biological processes* are disrupted in acute Lyme disease. We should see how much it matches the original study's findings


## 11.3 ORA: GO Biological Process (V1)

```{r ora-gobp-v1}
ora_gobp_v1 <- enricher(
  gene = sig_sym_v1, universe = univ_sym,
  TERM2GENE = t2g_gobp, TERM2NAME = t2n_gobp,
  pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 1
)

cat("Significant GO:BP terms:", nrow(as.data.frame(ora_gobp_v1)), "\n")
```

```{r ora-gobp-dotplot, fig.height=9}
if (nrow(as.data.frame(ora_gobp_v1)) > 0) {
  dotplot(ora_gobp_v1, showCategory = 20) +
    ggtitle("ORA: Top GO Biological Process Terms (V1)")
}
```

## 11.4 ORA Across All Time Points (Hallmark)

```{r ora-all-timepoints}
ora_hallmark_all <- list()

for (tp in names(res_list)) {
  sig_sym <- res_list[[tp]] %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                  !is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
  univ <- res_list[[tp]] %>%
    dplyr::filter(!is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
  ora_hallmark_all[[tp]] <- enricher(
    gene = sig_sym, universe = univ,
    TERM2GENE = t2g_hallmark,
    pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 1)
}

cat("Enriched Hallmark pathways: V1=",
    nrow(as.data.frame(ora_hallmark_all$V1)),
    " V2=", nrow(as.data.frame(ora_hallmark_all$V2)),
    " V5=", nrow(as.data.frame(ora_hallmark_all$V5)), "\n")
```

---
---

# 12. 2nd Generation: GSEA (Gene Set Enrichment Analysis)

*Ref: Thomas's book Ch. 8.1.2.1, 8.4.3; Subramanian et al. 2005*

GSEA is a major improvement over ORA because it uses **all genes ranked
by fold change**, not just those passing an arbitrary significance
cutoff. It detects subtle, coordinated changes in pathways that ORA
misses. The original study did NOT perform GSEA -- this is a novel
contribution of our re-analysis.

## 12.1 Build Ranked Gene Lists

```{r gsea-ranked}
make_ranked_list <- function(res_df) {
  res_df %>%
    dplyr::filter(!is.na(Symbol) & Symbol != "" & !is.na(log2FoldChange)) %>%
    dplyr::arrange(dplyr::desc(log2FoldChange)) %>%
    dplyr::distinct(Symbol, .keep_all = TRUE) %>%
    dplyr::select(Symbol, log2FoldChange) %>%
    tibble::deframe()
}

ranked_lists <- lapply(res_list, make_ranked_list)
```

## 12.2 GSEA: Hallmark (V1)

```{r gsea-hallmark-v1}
gsea_hallmark_v1 <- GSEA(
  geneList = ranked_lists$V1,
  TERM2GENE = t2g_hallmark,
  pvalueCutoff = 1, eps = 0
)

gsea_h_df <- as.data.frame(gsea_hallmark_v1)
cat("Significant (padj<0.05):", sum(gsea_h_df$p.adjust < 0.05),
    "of", nrow(gsea_h_df), "pathways\n")
```

```{r gsea-dotplot, fig.height=8}
if (sum(gsea_h_df$p.adjust < 0.05) > 0)
  dotplot(gsea_hallmark_v1, showCategory = 20, split = ".sign") +
    facet_grid(~ .sign) +
    ggtitle("GSEA: Hallmark Pathways (V1 vs Control)")
```

**Interpretation:** The dotplot splits pathways into **activated**
(enriched in upregulated genes) and **suppressed** (enriched in downregulated genes). The **Normalised Enrichment Score (NES)** measures how strongly the pathway's genes cluster at the top (activated) or bottom (suppressed) of the ranked list. Which pathways should we expect to be activated; translation-related pathways may be suppressed?


## 12.3 GSEA Enrichment Plots for Key Pathways

These "mountain plots" show how a pathway's member genes are distributed in the ranked list. The green line rises when it encounters pathway genes and falls otherwise. A peak at the **left** = upregulated; **right** =
downregulated.

```{r gsea-mountains, fig.height=4, fig.width=8, results='asis'}
key_pathways <- c("HALLMARK_INFLAMMATORY_RESPONSE",
                  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                  "HALLMARK_COMPLEMENT")

for (pw in key_pathways)
  if (pw %in% gsea_h_df$ID)
    print(gseaplot2(gsea_hallmark_v1, geneSetID = pw, title = pw))
```

## 12.4 GSEA Across All Time Points

```{r gsea-all-timepoints}
gsea_results <- list()
for (tp in names(ranked_lists))
  gsea_results[[tp]] <- GSEA(geneList = ranked_lists[[tp]],
                              TERM2GENE = t2g_hallmark,
                              pvalueCutoff = 1, eps = 0)
```

```{r nes-heatmap, fig.height=9, fig.width=7}
nes_all <- lapply(names(gsea_results), function(tp) {
  as.data.frame(gsea_results[[tp]]) %>%
    dplyr::select(ID, NES, p.adjust) %>%
    dplyr::mutate(timepoint = tp)
}) %>% dplyr::bind_rows()

# Keep pathways significant in at least one time point
sig_any <- nes_all %>% dplyr::filter(p.adjust < 0.05) %>%
  dplyr::pull(ID) %>% unique()

if (length(sig_any) > 0) {
  nes_mat <- nes_all %>%
    dplyr::filter(ID %in% sig_any) %>%
    dplyr::select(ID, NES, timepoint) %>%
    tidyr::pivot_wider(names_from = timepoint, values_from = NES) %>%
    tibble::column_to_rownames("ID") %>% as.matrix()
  nes_mat[is.na(nes_mat)] <- 0

  pheatmap(nes_mat,
           color = colorRampPalette(c("steelblue","white","firebrick"))(100),
           cluster_cols = FALSE,
           main = "GSEA NES: Hallmark Pathways Across Time Points",
           fontsize_row = 7)
}
```



---
---

# 13. Biologic Theme Comparison

*Ref: Thomas's book Ch. 8.4.2*

`compareCluster` from clusterProfiler directly compares enrichment
results across multiple gene lists (our three time points), making the
temporal dynamics visually clear.

```{r compare-cluster, fig.height=10, fig.width=10}
gene_lists <- lapply(res_list, function(df) {
  df %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                  !is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
})

cc_result <- compareCluster(
  geneClusters = gene_lists,
  fun = "enricher",
  TERM2GENE = t2g_hallmark,
  pvalueCutoff = 0.05, pAdjustMethod = "BH"
)

if (nrow(as.data.frame(cc_result)) > 0)
  dotplot(cc_result) +
    ggtitle("Biologic Theme Comparison: Hallmark Across Time Points") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



---
---

# 14. Save All Downstream Results

```{r save-downstream}
for (tp in names(ora_hallmark_all)) {
  df <- as.data.frame(ora_hallmark_all[[tp]])
  if (nrow(df) > 0) write.csv(df, file.path(results_dir,
    paste0("ORA_Hallmark_", tp, ".csv")), row.names = FALSE)
}
for (tp in names(gsea_results)) {
  df <- as.data.frame(gsea_results[[tp]])
  if (nrow(df) > 0) write.csv(df, file.path(results_dir,
    paste0("GSEA_Hallmark_", tp, ".csv")), row.names = FALSE)
}
cat("All results saved to:", results_dir, "\n")
```

---
---

---

# Session Info

```{r session-info}
sessionInfo()
```
