---
title: "Lyme Disease Differential Gene Expression Re-Analysis"
subtitle: "Re-analysis of Bouquet et al. 2016 (GSE63085) using DESeq2 & clusterProfiler"
author: "Abed & Laszlo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 7
---


# Introduction

In this notebook, we re-analyse the publicly available RNA-seq dataset GSE63085 from Bouquet et al. 2016, which profiled peripheral blood gene expression in Lyme disease patients at multiple time points (acute infection, post-treatment, and 6-month follow-up) compared to healthy controls. The original study used: 

* outdated hg19 reference genome, 
* TopHat2/Cufflinks2 for alignment and quantification, and 
* Limma/Voom for differential expression analysis, with pathway enrichment via the proprietary IPA tool.


Lyme disease is a tick-borne infection caused by *Borrelia burgdorferi*. It can lead to a range of symptoms, from acute flu-like illness to chronic fatigue and joint pain. Understanding the host immune response at the transcriptomic level can provide insights into disease mechanisms and potential therapeutic targets.



# Summary of the re-analysis Implementation

## How Our Re-Analysis Compares to the Original Study

| Aspect | Bouquet et al. 2016 | Our Re-Analysis |
|--------|---------------------|-----------------|
| Genome | hg19 (2009) | GRCh38.p13 (current) |
| Aligner | TopHat2/Cufflinks2 (deprecated) | Kallisto (modern) |
| Input data | FPKM (normalised) | Raw integer counts |
| DE method | Limma/Voom | DESeq2 (negative binomial) |
| Fold-change | Raw FC > 1.5 | Shrunken LFC (ashr) |
| Pathway tool | IPA (proprietary, 3$) | clusterProfiler + MSigDB (open-source, free) |
| Pathway methods | ORA only (1st gen) | ORA + GSEA + GSVA + SPIA (1st–3rd gen) |
| Per-sample resolution | No (bulk group comparison) | Yes (GSVA scores per patient) |

### What we confirmed

- A strong immune/inflammatory signature at V1 (acute Lyme)
- The signature persists beyond antibiotic treatment (V2)
- Partial recovery by 6 months (V5) but not to baseline
- TNF-alpha, interferons, complement, and IL-6 signaling are central

### What we added

- **GSEA** detects subtle coordinated pathway changes without
  arbitrary cutoffs, finding more significant pathways than ORA alone.
- **GSVA** provides individual patient pathway scores, allowing us to
  see heterogeneity within each time point -- particularly relevant for
  understanding why some V5 patients still have elevated inflammatory
  pathways while others recover fully.
- All tools are open-source and reproducible (vs. IPA).
- **SPIA (3rd gen)** highlights KEGG pathways where both **gene list enrichment** and **signed topology** suggest activation or inhibition, a different lens from MSigDB Hallmark.



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center"
)
```

# 1. Environment Setup

## 1.1 Install Packages

```{r install-packages, eval=FALSE, include=FALSE}
# install.packages(c("tidyverse", "pheatmap", "RColorBrewer", "ggrepel",
#                     "gridExtra", "viridis", "ashr"))
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(c("DESeq2", "clusterProfiler", "msigdbr",
#                         "enrichplot", "org.Hs.eg.db", "GSVA", "SPIA"))
```

```{r load-libraries}
# if (rstudioapi::isAvailable()) {setwd(dirname(rstudioapi::getActiveDocumentContext()$path))}
options(width = 200)

# use this function to install and load packages
using <- function(pkgs) {
  for (pkg in pkgs) {  # Loop through each package in the provided vector
    if (!requireNamespace(pkg, quietly = TRUE)) {    # If not installed, install it
      install.packages(pkg, repos = "https://cloud.r-project.org")}
    # Load the package:
    library(pkg, character.only = TRUE)}}

using(c("tidyverse", "pheatmap", "RColorBrewer", "ggrepel", "corrplot", "gridExtra", "viridis", "BiocManager","DESeq2", "org.Hs.eg.db", "clusterProfiler", "EnhancedVolcano","msigdbr", "PCAtools", "GSVA", "enrichplot", "ashr", "matrixStats"))

# SPIA is Bioconductor-only (3rd-generation pathway topology); not on CRAN
if (!requireNamespace("SPIA", quietly = TRUE))
  BiocManager::install("SPIA", ask = FALSE, update = FALSE)
suppressPackageStartupMessages(library(SPIA))

# description of libraries:
# tidyverse       - data wrangling + ggplot2
# DESeq2          - differential expression (negative binomial model)
# clusterProfiler - pathway enrichment (ORA, GSEA)
# msigdbr         - access to MSigDB gene sets
# enrichplot      - enrichment result visualizations
# pheatmap        - heatmaps with clustering
# RColorBrewer    - colour palettes
# ggrepel         - non-overlapping text labels on plots
# gridExtra       - arrange multiple plots
# viridis         - colour-blind-friendly palettes
# matrixStats     - fast row/column statistics

theme_set(theme_bw(base_size = 13)) # this is for the plots 
```


We had to use explicit `dplyr::` prefixes on key functions (`filter`, `select`, `count`, `mutate`, `arrange`) throughout this document because
`clusterProfiler` loads packages that mask some of these names.


## 1.3 Directories

```{r directories}
project_dir <- getwd()
data_dir    <- file.path(project_dir, "Data")
results_dir <- file.path(project_dir, "results")
plots_dir   <- file.path(project_dir, "plots")

for (d in c(results_dir, plots_dir))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
```

---
---

# 2. Data Loading

The raw FASTQ reads from GEO (GSE63085) were re-aligned by Thomas to the modern **GRCh38.p13** human reference genome (replacing
the outdated hg19 used in the original 2016 study) and quantified into raw integer counts, the proper input for DESeq2. Thomas used Kallisto alligner. And more importantly, the 22 missing samples were excluded during re-alignment, will be checked and confirmed by Thomas. 

**References:**

- Bouquet J et al. *mBio* 2016; 7(1):e00100-16.
- Thomas's book T. *Beginner's Handbook of -Omics Data Analysis*, Ch. 7-8.

## 2.1 Raw Counts Matrix

```{r load-counts}
counts_raw <- read.delim(
  gzfile(file.path(data_dir, "GSE63085_raw_counts_GRCh38.p13_NCBI.tsv.gz")),
  row.names = 1, check.names = FALSE
)
cat("Counts matrix:", nrow(counts_raw), "genes x", ncol(counts_raw), "samples\n")
```

## 2.2 Gene Annotation (featureData)

This maps NCBI Entrez Gene IDs to gene symbols, biotypes, chromosomal locations, and GO terms.

NCBI Entrez as the "Google Search" engine for biological data. It’s a massive integrated system managed by the National Center for Biotechnology Information (NCBI) that links together different databases, DNA sequences, protein structures, and medical literature (PubMed).
According to online sources, when we see an Entrez ID (like those numbers in your matrix), it’s a unique "social security number" for a gene. Names (Symbols) change as scientists argue, but Entrez IDs stay the same.

```{r load-gene-annot}
gene_annot <- read.delim(
  gzfile(file.path(data_dir, "Human.GRCh38.p13.annot.tsv.gz")),
  stringsAsFactors = FALSE
)
cat("Gene annotation:", nrow(gene_annot), "x", ncol(gene_annot), "\n")
```
> Explanation of the columns (source: online):

* **GeneID**: This matches the row names in your counts matrix. It’s your primary key.

* **Symbol**: The "human-friendly" name (e.g., TP53, BRCA1). This is what you’ll put in your final paper or report.

* **Description**: A long-form name of what the gene actually does (e.g., "tumor protein p53").

* **EnsemblGeneID**: Another ID system (starting with ENSG...). Bioinformaticians often use both Entrez and Ensembl because some tools prefer one over the other.

* **GeneType**: Usually, you only want to analyze protein-coding genes. Most researchers filter out the pseudo (dead "fossil" genes) and ncRNA (non-coding) unless they are specifically looking for them.

* **Status**: You should likely filter for "active". A "discontinued" status means the gene was merged into another record or found to be a mistake in the genome mapping.

* **ChrAcc, ChrStart, ChrStop, Orientation**: These are the "Location" columns. They tell you exactly where the gene sits on which chromosome and which way it's facing. This is vital if you want to see if a treatment affects a whole neighborhood of genes (a "cluster") or just one.

* **GOFunction / GOProcess / GOComponent**: "GO" stands for Gene Ontology. These tell you the "job" of the gene. **Process**: What is the "big picture" goal? (e.g., "Cell Division"); **Function**: What is the specific biochemical task? (e.g., "Binding to DNA"); and **Component**: Where does it happen? (e.g., "Mitochondria").


```{r meta_EDA}
names(gene_annot)

cat("Status unique:");print(table(gene_annot$Status))
```
> We have ~39k total entries, but only 19,416 are protein-coding. 

## 2.3 Sample Annotation

```{r load-sample-annot}
sample_annot <- read.csv(
  file.path(project_dir, "sample_annotations.csv"),
  stringsAsFactors = FALSE
)
```

## 2.4 Align Samples

22 samples from the annotation are absent from the counts matrix.
The reason is not confirmed.

```{r align-data}
shared     <- intersect(colnames(counts_raw), sample_annot$GSM)
only_annot <- setdiff(sample_annot$GSM, colnames(counts_raw))

cat("Shared:", length(shared), "| Not in counts:", length(only_annot), "\n")

counts <- counts_raw[, shared]
annot  <- sample_annot %>%
  dplyr::filter(GSM %in% shared) %>%
  dplyr::arrange(match(GSM, shared))

stopifnot(all(colnames(counts) == annot$GSM))
```

## 2.5 Engineer Metadata Columns

We derive three columns from the raw `Observation` string:

- `patient_id`, the numeric patient identifier embedded in the observation name (e.g. `LymeP-12_V1` → `12`). Parsed via two stacked `gsub()` calls so both `_V1/V2/V5` and bare `_V` formats match.
- {`visit`: [`V1`, `V2`, `V5`]}, or {`Control`}.
- `group`, the four-level factor used as the DESeq2 design: `Control / Lyme_V1 / Lyme_V2 / Lyme_V5`.

The patient-ID parser is fragile by nature (it depends on the exact observation naming convention from GEO), so we add a hard sanity check immediately after: every Lyme sample must yield a *purely numeric* patient ID. If GEO ever changes the format, the check fires here rather than producing silently wrong patient trajectories in the PCA (Section 8.3).

```{r engineer-metadata}
annot <- annot %>%
  dplyr::mutate(
    patient_id = gsub("^.*-([0-9]+)_V.*$", "\\1",
                      gsub("^.*-([0-9]+)_V$", "\\1", Observation)),
    visit = dplyr::case_when(
      grepl("V1", Observation) ~ "V1",
      grepl("V2", Observation) ~ "V2",
      grepl("V5", Observation) ~ "V5",
      TRUE                     ~ "Control"),
    group = dplyr::case_when(
      disease.state == "Healthy controls" ~ "Control",
      visit == "V1" ~ "Lyme_V1",
      visit == "V2" ~ "Lyme_V2",
      visit == "V5" ~ "Lyme_V5"),
    condition = ifelse(disease.state == "Healthy controls", "Control", "Lyme"),
    group     = factor(group,     levels = c("Control","Lyme_V1","Lyme_V2","Lyme_V5")),
    visit     = factor(visit,     levels = c("Control","V1","V2","V5")),
    condition = factor(condition, levels = c("Control","Lyme"))
  )
rownames(annot) <- annot$GSM

# Sanity check: every Lyme patient_id should be purely numeric.
# If GEO ever changes the Observation format, this will fail loudly here
# instead of producing silently wrong patient trajectories downstream.
lyme_pids <- annot$patient_id[annot$condition == "Lyme"]
stopifnot(
  "patient_id parsing produced NAs for Lyme samples"      = !any(is.na(lyme_pids)),
  "patient_id contains non-numeric values for Lyme samples"= all(grepl("^[0-9]+$", lyme_pids))
)
cat("patient_id parser OK:",
    length(unique(lyme_pids)), "unique Lyme patients across",
    length(lyme_pids), "Lyme samples.\n\n")

cat("Samples per group:\n")
print(table(annot$group))
```

---
---

# 3. Gene Biotype Overview

```{r gene-biotypes}
gene_annot_matched <- gene_annot %>%
  dplyr::filter(GeneID %in% as.numeric(rownames(counts)))

biotype_counts <- gene_annot_matched %>%
  dplyr::count(GeneType, sort = TRUE) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))

ggplot(biotype_counts %>% dplyr::filter(n > 50),
       aes(x = reorder(GeneType, n), y = n, fill = GeneType)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(n, " (", pct, "%)")), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Gene Biotypes in Counts Matrix (GRCh38.p13)",
       x = NULL, y = "Number of Genes") +
  scale_fill_viridis_d()
```

About half the genes are **protein-coding** (~49%).
The rest are non-coding RNA, pseudogenes, etc. According to online sources, protein-coding genes carry the strongest biological signal for differential expression. \n 
After filtering (Section 6), the protein-coding proportion will increase because non-coding genes tend to have very low counts.

---

# 4. Sample-Level Quality Control

*Ref: Thomas's book Ch. 7.2*

## 4.1 Library Sizes

Library size = total mapped reads per sample. Large variation means
different sequencing depths, which must be normalised.

```{r library-sizes, fig.height=10}
lib_sizes <- data.frame(
  GSM = colnames(counts), total = colSums(counts)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group, Observation), by = "GSM")

ggplot(lib_sizes, aes(x = reorder(Observation, total),
                       y = total / 1e6, fill = group)) +
  geom_col() + coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library Size per Sample",
       x = NULL, y = "Total Counts (millions)", fill = "Group") +
  theme(axis.text.y = element_text(size = 5))
```

```{r lib-size-boxplot}
ggplot(lib_sizes, aes(x = group, y = total / 1e6, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library Size by Group", x = NULL, y = "Total Counts (millions)") +
  theme(legend.position = "none")
```

### How extreme is the library-size imbalance, and is it addressed?

The bar plot looks visually dramatic, so we quantify it: we report the
**fold-range** (max / min library size) and the **coefficient of variation
(CV = SD / mean)**. A CV > 0.5 or a fold-range > 5 would be a red flag for
naive between-sample comparison (e.g. raw FPKM-style normalisation of the
2016 paper). DESeq2's median-of-ratios method, however, is **robust** to
this kind of imbalance because it estimates per-sample size factors from
the geometric mean of gene-wise ratios and not from total counts.

```{r lib-size-cv}
lib_summary <- lib_sizes %>%
  dplyr::summarise(
    n           = dplyr::n(),
    min_M       = round(min(total) / 1e6, 1),
    median_M    = round(median(total) / 1e6, 1),
    max_M       = round(max(total) / 1e6, 1),
    fold_range  = round(max(total) / min(total), 2),
    CV          = round(sd(total) / mean(total), 3)
  )
print(lib_summary)
```

The visual imbalance reflects real depth differences across the group, but **the imbalance is fully handled downstream**:

1. **Section 7.2** estimates DESeq2 size factors via median-of-ratios
   (Anders & Huber 2010) — these scale each sample's counts to a common
   reference *without* assuming equal library size or composition.
2. **Section 7.3 / 10.1a** verifies, via MA plots after normalisation,
   that the M = 0 baseline is restored.
3. **Section 7.4** further stabilises variance for visualisation (PCA,
   heatmaps) using VST.

If a single sample looked like a strong outlier (e.g. > 3× median library
size *and* a wildly different RLE in 4.3), we would consider dropping it.
Otherwise, the model handles depth differences correctly.

## 4.2 Detected Genes per Sample

```{r detected-genes}
detected <- data.frame(
  GSM = colnames(counts), n_detected = colSums(counts > 0)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(detected, aes(x = group, y = n_detected, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Detected Genes per Sample (count > 0)",
       x = NULL, y = "Genes Detected") +
  theme(legend.position = "none")
```

## 4.3 Relative Log Expression (RLE)

*Ref: Thomas's book Ch. 7.2.4*

RLE measures how much each gene in a sample deviates from its median
expression across all samples. Well-behaved samples centre around zero.
Samples with shifted or wide RLE distributions may indicate technical
problems or batch effects.

```{r rle-plot, fig.height=8}
# Compute RLE: for each gene, subtract the row median
log2_counts <- log2(counts + 1)
row_medians <- rowMedians(as.matrix(log2_counts))
rle_matrix  <- sweep(log2_counts, 1, row_medians)

rle_long <- rle_matrix %>%
  as.data.frame() %>%
  tidyr::pivot_longer(everything(), names_to = "GSM", values_to = "RLE") %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(rle_long, aes(x = GSM, y = RLE, fill = group)) +
  geom_boxplot(outlier.size = 0.1) +
  scale_fill_brewer(palette = "Set2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Relative Log Expression (RLE) Plot",
       subtitle = "Well-behaved samples centre around the red dashed line (zero)",
       x = NULL, y = "Deviation from median expression") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 3)) +
  coord_cartesian(ylim = c(-2, 2))
```

**Bingo:** Samples with medians shifted away from zero or with
unusually wide boxes may have quality issues. If all groups looked similar,
there wouldn't be strong batch effect in the raw data. But apparently, they don't, so there could be, supporting our hypothesis that there could be batch effects.

---
---

# 5. Expression Distributions

## 5.1 Per-Sample Density

```{r density-plots}
log2_long <- log2_counts %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  tidyr::pivot_longer(-gene, names_to = "GSM", values_to = "log2_count") %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(log2_long, aes(x = log2_count, group = GSM, color = group)) +
  geom_density(linewidth = 0.3) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Per-Sample Density of log2(count + 1)",
       x = "log2(count + 1)", y = "Density", color = "Group") +
  xlim(0, 20)
```

the large spike near zero represents the many genes
with zero or very low expression. Most of them are still there because we haven't filtered out the garbage yet. The second hump (should be around 5-10) represents actively expressed genes. What we are looking for looks flat because the scale is zoomed out to accommodate the zero spike. All samples should have a similar shape: if one looks radically different, it may be a quality outlier. Perhaps, we should make a similar plot after filtering.

## 5.2 Mean-Variance Relationship

```{r mean-variance}
gene_stats <- data.frame(
  mean_count = rowMeans(counts), var_count = apply(counts, 1, var)
)

ggplot(gene_stats, aes(x = log10(mean_count + 1), y = log10(var_count + 1))) +
  geom_point(alpha = 0.05, size = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 0.8) +
  labs(title = "Mean-Variance Relationship (Raw Counts)",
       subtitle = "Red dashed = Poisson (variance = mean). Points above = overdispersion.",
       x = "log10(Mean + 1)", y = "log10(Variance + 1)")
```

Nearly all genes sit above the red line, meaning variance > mean **if my mean is set correctly**. According to online sources: this overdispersion is exactly why DESeq2 uses a negative binomial model instead of a simpler Poisson model. A Poisson
model would dramatically underestimate variability and produce too many false positives.

## 5.3 MA Plot (pre-normalisation)

*Ref: Thomas's book Ch. 7.2.7*

An MA plot shows each gene's **average expression** (A) vs. its **log
fold change** (M) between two conditions. Before normalisation, the cloud
should centre around M=0 (most genes are not differentially expressed).
A shift or banana shape indicates a normalisation problem.

```{r ma-plot-pre-norm, fig.height=5}
# Quick MA: V1 (first patient) vs first Control
ctrl_cols  <- annot$GSM[annot$group == "Control"]
lyme_cols  <- annot$GSM[annot$group == "Lyme_V1"]

A <- 0.5 * (rowMeans(log2_counts[, ctrl_cols]) + rowMeans(log2_counts[, lyme_cols]))
M <- rowMeans(log2_counts[, lyme_cols]) - rowMeans(log2_counts[, ctrl_cols])

ma_df <- data.frame(A = A, M = M)

ggplot(ma_df, aes(x = A, y = M)) +
  geom_point(alpha = 0.05, size = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "orange", se = FALSE, linewidth = 0.8) +
  labs(title = "MA Plot: Lyme V1 vs Control (pre-normalisation)",
       subtitle = "Red = zero line. Orange = loess trend. Centred cloud = good.",
       x = "A (average expression)", y = "M (log2 fold change)") +
  coord_cartesian(ylim = c(-5, 5))
```

Cloud isn't centered around 0, indicating a scaling issue. 

---
---

# 6. Gene Filtering

We remove genes with very low counts to reduce noise. We require at
least N samples (= smallest group size) to have a count >= 10.

```{r prefilter}
smallest_group <- min(table(annot$group))
keep <- rowSums(counts >= 10) >= smallest_group
counts_filtered <- counts[keep, ]

cat("Before:", nrow(counts), "genes\n")
cat("After: ", nrow(counts_filtered), "genes\n")
cat("Removed:", nrow(counts) - nrow(counts_filtered),
    "(", round(100 * (1 - nrow(counts_filtered)/nrow(counts)), 1), "%)\n")
```

```{r post-filter-biotypes}
biotype_after <- gene_annot %>%
  dplyr::filter(GeneID %in% as.numeric(rownames(counts_filtered))) %>%
  dplyr::count(GeneType, sort = TRUE) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))
print(biotype_after)
```

**Meaning:** After aggressive filtering, ~70% of remaining genes are **protein-coding** (up from ~49%), confirming that most low-count genes were non-coding RNA and pseudogenes. This is expected and desirable.

---
---

# 7. DESeq2 Normalisation

*Ref: Thomas's book Ch. 7.6e*

## 7.1 Create DESeqDataSet

The `design = ~ group` formula tells DESeq2 to model expression as a function of our four experimental groups.

```{r create-dds}
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered, colData = annot, design = ~ group)
```



## 7.2 Size Factors

DESeq2 estimates a **size factor** per sample via the median-of-ratios method (Anders & Huber 2010). This **directly addresses the library-size imbalance flagged in Section 4.1**: each sample's counts are divided by its size factor before any statistical test, so a sample sequenced 3× deeper than another is not falsely flagged as having "more expression". This is far superior to the FPKM normalisation used in the original 2016 study, which only corrects for length and depth without controlling for sample composition.

```{r size-factors}
dds <- estimateSizeFactors(dds)

sf_df <- data.frame(
  GSM = colnames(dds), size_factor = sizeFactors(dds),
  lib_size = colSums(counts_filtered)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(sf_df, aes(x = lib_size / 1e6, y = size_factor, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "gray40", linetype = "dashed") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "DESeq2 Size Factor vs Library Size",
       x = "Library Size (millions)", y = "Size Factor")

cat("Size factor range: ", round(range(sizeFactors(dds)), 3),
    " (fold:", round(max(sizeFactors(dds)) / min(sizeFactors(dds)), 2), ")\n")
```

Size factor and library size are strongly correlated (expected). 
Samples with a size factor < 1 were sequenced less deeply; > 1 means more deeply. 
DESeq2 divides each sample's counts by its size factor during analysis, so the apparent depth imbalance from Section 4.1 has no effect on the downstream DE tests. The MA plots in **Sections 7.3 and 10.1a** confirm visually that the bulk of genes settles around log fold-change = 0 after this correction.

> **Note on dataset variability.** A wide spread of size factors *and* a strong group separation in PCA (Section 8) together indicate that the dataset has substantial **between-sample variability**. This is biologically expected, the cohort is a mix of healthy controls, acute Lyme, post-treatment, and 6-month follow-up patients with heterogeneous immune trajectories. The negative-binomial **dispersion** estimated by DESeq2 (visualised in Section 10.1b) absorbs this variability, which is why DESeq2 typically gives more conservative (and more reliable) p-values than naive Limma/Voom on FPKM in noisy clinical data.

## 7.3 MA Plot (after size-factor normalization)

*Ref: Thomas's book Ch. 7.2.7*

Once **size factors** are applied, `counts(dds, normalized = TRUE)` is on a comparable scale across samples. We pool Lyme V1 vs Control and plot **M** = log2 ratio of group means vs **A** = average log2 signal (same construction as Section 5.3, but using median-of-ratios normalized counts). The loess trend should sit near **M = 0** across most expression levels if depth correction worked; a strong banana or vertical offset would suggest residual technical bias (or real biology concentrated at certain expression ranges).

```{r ma-plot-norm-counts, fig.height=5}
nt <- counts(dds, normalized = TRUE)
ctrl_cols <- annot$GSM[annot$group == "Control"]
v1_cols   <- annot$GSM[annot$group == "Lyme_V1"]

m1 <- rowMeans(nt[, v1_cols, drop = FALSE])
m0 <- rowMeans(nt[, ctrl_cols, drop = FALSE])
A_norm <- 0.5 * (log2(m1 + 0.5) + log2(m0 + 0.5))
M_norm <- log2((m1 + 0.5) / (m0 + 0.5))

ggplot(data.frame(A = A_norm, M = M_norm), aes(x = A, y = M)) +
  geom_point(alpha = 0.05, size = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "orange", se = FALSE, linewidth = 0.8) +
  labs(title = "MA Plot: V1 vs Control (DESeq2 normalized counts)",
       subtitle = "After size factors only (GLM / dispersions not yet fitted).",
       x = "A (average log2 normalized signal)", y = "M (log2 fold change)") +
  coord_cartesian(ylim = c(-5, 5))
```

Compare with the pre-normalisation MA (Section 5.3): the cloud should be **more tightly centred** on M = 0 after size factors. A later **DESeq2 `plotMA()`** after the full model (Section 10.1) adds dispersion shrinkage and hypothesis testing; that plot is the standard check bundled with DESeq2.

## 7.4 Variance Stabilising Transformation (VST)

Raw counts have a mean-variance dependency (Section 5.2). VST produces
values on a log2-like scale with approximately **constant variance**,
suitable for PCA and heatmaps.

```{r vst}
vsd <- vst(dds, blind = FALSE)
```

```{r vst-effect, fig.height=5}
par(mfrow = c(1, 2))
plot(rowMeans(log2(counts_filtered + 1)),
     apply(log2(counts_filtered + 1), 1, sd),
     pch = 16, cex = 0.3, col = rgb(0.2, 0.4, 0.8, 0.1),
     xlab = "Mean", ylab = "SD", main = "Before: log2(counts+1)")
plot(rowMeans(assay(vsd)), apply(assay(vsd), 1, sd),
     pch = 16, cex = 0.3, col = rgb(0.8, 0.2, 0.2, 0.1),
     xlab = "Mean", ylab = "SD", main = "After: VST")
par(mfrow = c(1, 1))
```



Before VST (left), low-expression genes have variable SD. After VST (right), the trend should be flattened, but it's not, so the SD still depends on the mean. This makes the meaningfullness of distances between samples (for PCA, clustering) questionable.



The following quantifies the mean–SD coupling before vs after VST so we don't have to read it off the eye.

```{r vst-correlation-check}
mean_sd_before <- cor(
  rowMeans(log2(counts_filtered + 1)),
  apply(log2(counts_filtered + 1), 1, sd)
)
mean_sd_after <- cor(
  rowMeans(assay(vsd)),
  apply(assay(vsd), 1, sd)
)

cat("Mean-SD correlation BEFORE VST (log2(counts+1)) :",
    round(mean_sd_before, 3), "\n")
cat("Mean-SD correlation AFTER  VST                  :",
    round(mean_sd_after,  3), "\n")
cat("Reduction (closer to zero is better)            :",
    round(mean_sd_before - mean_sd_after, 3), "\n")

if (mean_sd_after < mean_sd_before) {
  cat("VST DID reduce mean-variance coupling — usable for PCA / heatmaps.\n")
} else {
  cat("VST did NOT reduce mean-variance coupling — investigate (try rlog or check for outlier samples).\n")
}
```

A **positive** reduction (before − after > 0) means VST did its job. A typical drop is from ~0.7–0.9 (raw log2) to ~0.2–0.5 (after VST); the residual coupling is normal, VST is an approximation, not a guarantee of zero correlation. Confirm with Thomas if the reduction is small (<0.1) or if VST made things worse, in which case `rlog()` may be preferable for this dataset.


---
---

# 8. PCA

*Ref: Thomas's book Ch. 7.2.2*

```{r pca-compute}
rv      <- rowVars(assay(vsd))
top_idx <- order(rv, decreasing = TRUE)[1:min(2000, length(rv))]

pca     <- prcomp(t(assay(vsd)[top_idx, ]), center = TRUE, scale. = TRUE)
var_pct <- round(100 * summary(pca)$importance[2, ], 1)

pca_df  <- data.frame(pca$x[, 1:5]) %>%
  tibble::rownames_to_column("GSM") %>%
  dplyr::left_join(annot, by = "GSM")
```

## 8.1 Scree Plot

```{r scree}
ggplot(data.frame(PC = 1:20, Var = var_pct[1:20]),
       aes(x = PC, y = Var)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_line(color = "darkred") + geom_point(color = "darkred", size = 2) +
  labs(title = "Scree Plot", x = "Principal Component",
       y = "% Variance Explained") +
  scale_x_continuous(breaks = 1:20)
```

## 8.2 PCA by Group

```{r pca-group, fig.width=15, fig.height=5}
make_pca <- function(df, pcx, pcy, vpct) {
  ggplot(df, aes(x = .data[[pcx]], y = .data[[pcy]], color = group)) +
    geom_point(size = 3, alpha = 0.8) +
    stat_ellipse(level = 0.68, linetype = "dashed") +
    scale_color_brewer(palette = "Set2") +
    labs(x = paste0(pcx, " (", vpct[as.integer(gsub("PC","",pcx))], "%)"),
         y = paste0(pcy, " (", vpct[as.integer(gsub("PC","",pcy))], "%)"))
}

grid.arrange(
  make_pca(pca_df,"PC1","PC2",var_pct) + ggtitle("PC1 vs PC2"),
  make_pca(pca_df,"PC1","PC3",var_pct) + ggtitle("PC1 vs PC3"),
  make_pca(pca_df,"PC2","PC3",var_pct) + ggtitle("PC2 vs PC3"),
  ncol = 3
)
```


## 8.3 Patient Trajectories

Lines connect the same patient across V1 -> V2 -> V5.

```{r pca-trajectories, fig.height=7}
lyme_pca <- pca_df %>% dplyr::filter(condition == "Lyme") %>%
  dplyr::arrange(patient_id, visit)

ggplot() +
  geom_point(data = pca_df %>% dplyr::filter(condition == "Control"),
             aes(x = PC1, y = PC2), color = "gray60", size = 3, alpha = 0.5) +
  geom_path(data = lyme_pca,
            aes(x = PC1, y = PC2, group = patient_id),
            color = "gray40", alpha = 0.3) +
  geom_point(data = lyme_pca,
             aes(x = PC1, y = PC2, color = visit, shape = visit), size = 3) +
  scale_color_manual(values = c(V1="#d62728", V2="#ff7f0e", V5="#1f77b4")) +
  labs(title = "Patient Trajectories Through PCA Space",
       subtitle = "Gray = controls. Lines connect same patient across visits.",
       x = paste0("PC1 (", var_pct[1], "%)"),
       y = paste0("PC2 (", var_pct[2], "%)"))
```

Each line represents one patient's journey from acute infection (red V1) through treatment (orange V2) to 6-month follow-up (blue V5). Reassuringly, most V5 points move toward the gray control cloud, showing recovery. Patients whose V5 stays far from controls are the ones to worry about, they are likely candidates for **post-treatment Lyme disease syndrome (PTLDS)**.

---
---

# 9. Sample Relationships

*Ref: Thomas's book Ch. 7.2.1*

## 9.1 Distance Heatmap

```{r distance-heatmap, fig.height=9, fig.width=10}
sample_dists <- dist(t(assay(vsd)))
row_annot <- data.frame(Group = annot$group, row.names = annot$GSM)
ann_colors <- list(Group = c(Control="#66C2A5", Lyme_V1="#FC8D62",
                              Lyme_V2="#8DA0CB", Lyme_V5="#E78AC3"))

pheatmap(as.matrix(sample_dists),
         clustering_distance_rows = sample_dists,
         clustering_distance_cols = sample_dists,
         annotation_row = row_annot, annotation_col = row_annot,
         annotation_colors = ann_colors,
         show_rownames = FALSE, show_colnames = FALSE,
         color = viridis(100),
         main = "Sample-to-Sample Euclidean Distance (VST)")
```

## 9.2 Top 50 Most Variable Genes

```{r top50-heatmap, fig.height=12, fig.width=10}
top50_idx <- head(order(rv, decreasing = TRUE), 50)
top50_ids <- rownames(assay(vsd))[top50_idx]
top50_sym <- gene_annot$Symbol[match(as.numeric(top50_ids), gene_annot$GeneID)]
top50_lab <- ifelse(is.na(top50_sym) | top50_sym == "", top50_ids, top50_sym)

hm_mat <- assay(vsd)[top50_idx, ]
rownames(hm_mat) <- top50_lab

pheatmap(t(scale(t(hm_mat))),
         annotation_col = row_annot, annotation_colors = ann_colors,
         show_colnames = FALSE, fontsize_row = 7,
         color = colorRampPalette(c("navy","white","firebrick3"))(100),
         main = "Top 50 Most Variable Genes (Z-scored VST)",
         cutree_cols = 4)
```

Rows are genes, columns are samples. Red = high expression, blue = low expression (relative to the gene's mean). If the heatmap clusters samples by group (check the colour bar at top), it confirms that the dominant source of variation is biological (disease state / time point), not technical noise.

---
---

# 10. Differential Expression with DESeq2

*Ref: Thomas's book Ch. 7.6*

## 10.1 Run DESeq2

```{r run-deseq}
dds <- DESeq(dds)
resultsNames(dds)
```

## 10.1a Dispersion estimates (*quantifying how variant the dataset is*)

The single best diagnostic for "is this dataset noisy?" is the **dispersion plot**. DESeq2 fits a gene-wise dispersion (black dots), shrinks each estimate towards a fitted trend (red curve), and uses the shrunken values (blue) for testing. The y-axis is the **dispersion parameter α** of the negative binomial, which captures biological + technical variability *beyond* Poisson noise.

```{r plot-disp, fig.height=5}
plotDispEsts(dds, main = "DESeq2 Dispersion Estimates")
```



- The fitted (red) trend should slope **down** as expression increases, then flatten, that is the textbook shape and indicates the model assumptions hold.
- If many gene-wise dots sit **far above** the fitted line, the dataset is **highly variable** (lots of biological / between-patient noise). For our Lyme cohort this is expected: patients are sampled at different disease stages, with heterogeneous immune trajectories, and individual variation in the inflammatory response is well known. A median dispersion of ~0.05–0.1 is typical for clean cell-line RNA-seq; values of ~0.2–0.5 are typical for human PBMC studies like this one.
- Genes flagged as outliers (circled) are pulled down to the fitted curve by `lfcShrink()` (Section 10.2), which is precisely why we use shrunken LFCs.

```{r dispersion-summary}
disp <- dispersions(dds)
cat("Dispersion summary across", length(disp), "genes:\n")
print(summary(disp))
cat("\nMedian dispersion:", round(median(disp, na.rm = TRUE), 3),
    "  (PBMC studies typically 0.1–0.5)\n")
```

**Practical takeaway.** A relatively high median dispersion confirms what the wide library sizes (Section 4.1) and the spread of PCA points (Section 8) already hinted at: this is a **biologically noisy clinical dataset**, not a clean cell-line experiment. DESeq2's shrinkage absorbs that noise. Downstream DE calls are correspondingly conservative, which is why we report fewer DEGs than the original 2016 paper.

## 10.1b MA plots (DESeq2: normalized mean vs LFC)

After `DESeq()`, the object contains fitted dispersions and contrasts on the **same scale** as differential testing. `plotMA()` shows, for each gene, the **mean of normalized counts** (x-axis, "mean of normalized counts") vs **log2 fold change** (y-axis); points with padj below `alpha` are blue. For a successful normalization and contrast, the **bulk of points** should scatter around **LFC = 0** at moderate-to-high expression, with no strong systematic curve. These plots complement Section 7.3 (MA from normalized counts only) and Section 5.3 (raw log2 counts).

```{r plot-ma-deseq2, fig.width = 14, fig.height = 5}
par(mfrow = c(1, 3))
plotMA(
  results(dds, contrast = c("group", "Lyme_V1", "Control"), alpha = 0.05),
  main = "V1 vs Control", ylim = c(-5, 5))
plotMA(
  results(dds, contrast = c("group", "Lyme_V2", "Control"), alpha = 0.05),
  main = "V2 vs Control", ylim = c(-5, 5))
plotMA(
  results(dds, contrast = c("group", "Lyme_V5", "Control"), alpha = 0.05),
  main = "V5 vs Control", ylim = c(-5, 5))
par(mfrow = c(1, 1))
```

**Result validation:** If the cloud were systematically shifted away from LFC = 0 for most genes, we would revisit size factors, outliers, or design formula. Here the **MA after normalization** (Section 7.3) and these **MA after the full DESeq2 fit** together support using the contrasts for pathway analysis.

## 10.2 Extract, Shrink & Annotate Results

`lfcShrink()` with `ashr` produces moderated log2 fold-change estimates that are more reliable for genes with low counts. *(Stephens 2016)*

```{r extract-results}
contrasts <- list(
  V1 = c("group", "Lyme_V1", "Control"),
  V2 = c("group", "Lyme_V2", "Control"),
  V5 = c("group", "Lyme_V5", "Control")
)

get_annotated_results <- function(dds, contrast, gene_annot) {
  res <- results(dds, contrast = contrast, alpha = 0.05)
  res_shrunk <- lfcShrink(dds, contrast = contrast, type = "ashr", res = res)

  as.data.frame(res_shrunk) %>%
    tibble::rownames_to_column("GeneID") %>%
    dplyr::mutate(GeneID_num = as.numeric(GeneID)) %>%
    dplyr::left_join(
      gene_annot %>% dplyr::select(GeneID, Symbol, Description, GeneType),
      by = c("GeneID_num" = "GeneID")) %>%
    dplyr::select(-GeneID_num) %>%
    dplyr::arrange(padj)
}

res_list <- lapply(contrasts, function(ct)
  get_annotated_results(dds, ct, gene_annot))
```

## 10.3 DEG Summary

```{r deg-summary}
deg_summary <- lapply(names(res_list), function(nm) {
  sig <- res_list[[nm]] %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1)
  data.frame(Comparison = paste0(nm, " vs Control"),
             Total = nrow(sig),
             Up = sum(sig$log2FoldChange > 0),
             Down = sum(sig$log2FoldChange < 0))
}) %>% dplyr::bind_rows()

print(deg_summary)
```

Compare with the original study (Limma/Voom on FPKM): V1=1,235, V2=1,060, V5=686 DEGs. 
Not a 100% sure why our numbers differ, but I assume it's because we use: 

(1) a different genome reference

(2) raw counts instead of FPKM

(3) DESeq2 instead of Limma/Voom, and (4) different fold-change/FDR thresholds.

The pattern should be similar: most DEGs at V1 (acute), declining through V2 and V5 as the transcriptome partially recovers.

## 10.4 Top DEGs across all time points

We list the top 15 DEGs at each visit (acute V1, post-treatment V2, 6-month follow-up V5) so readers can scan the strongest signals at every stage of the disease trajectory in one place.

```{r top-degs}
top_degs <- function(df, n = 15) {
  df %>%
    dplyr::filter(!is.na(padj) & padj < 0.05) %>%
    head(n) %>%
    dplyr::select(Symbol, log2FoldChange, padj, GeneType)
}

for (tp in names(res_list)) {
  cat("\n===== Top 15 DEGs at", tp, "vs Control =====\n")
  print(top_degs(res_list[[tp]], 15))
}
```

Genes that *only* appear at V5 (rather than carrying over from V1) are typically related to homeostasis / repair and are interesting to flag separately.

## 10.5 Volcano Plots

```{r volcano, fig.width=16, fig.height=6}
make_volcano <- function(df, title) {
  df <- df %>%
    dplyr::mutate(sig = dplyr::case_when(
      is.na(padj) ~ "NS",
      padj < 0.05 & log2FoldChange > 1  ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"))
  top <- df %>% dplyr::filter(sig != "NS") %>% head(10)
  ggplot(df, aes(x = log2FoldChange, y = -log10(pvalue), color = sig)) +
    geom_point(alpha = 0.4, size = 1) +
    scale_color_manual(values = c(Up="firebrick", Down="steelblue", NS="gray70")) +
    geom_text_repel(data = top, aes(label = Symbol), size = 2.5,
                    max.overlaps = 15) +
    geom_vline(xintercept = c(-1,1), linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    labs(title = title, x = "log2 Fold Change", y = "-log10(p-value)") +
    theme(legend.position = "bottom")
}

grid.arrange(
  make_volcano(res_list$V1, "V1 (Acute) vs Control"),
  make_volcano(res_list$V2, "V2 (Post-Treatment) vs Control"),
  make_volcano(res_list$V5, "V5 (6 Months) vs Control"),
  ncol = 3
)
```

Each dot is a gene. The x-axis is the fold change (how much the gene changes in Lyme vs. Control). The y-axis is statistical significance. **Red dots** (upper-right) are significantly upregulated, **blue dots** (upper-left) are downregulated.

The labeled genes are the most significant. We can see how the "cloud" narrows from V1 to V5, fewer genes are dramatically changed at 6 months, matching the original paper's finding of a gradually resolving but sustained signature.

## 10.6 Genes of Interest — picks for the discussion



```{r genes-of-interest, fig.width=11, fig.height=6.5}
genes_of_interest <- c("CXCL10", "S100A8", "S100A9", "FCGR1A",
                       "IDO1", "CD274", "TLR2", "IL6",
                       "SERPING1", "ISG15", "IFI27", "BATF2")

goi_df <- res_list$V1 %>%
  dplyr::filter(Symbol %in% genes_of_interest) %>%
  dplyr::select(Symbol, log2FoldChange, padj, baseMean, Description) %>%
  dplyr::arrange(dplyr::desc(log2FoldChange))
print(goi_df)

make_volcano_highlight <- function(df, title, highlight) {
  df <- df %>%
    dplyr::mutate(sig = dplyr::case_when(
      is.na(padj) ~ "NS",
      padj < 0.05 & log2FoldChange > 1  ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"))
  hl <- df %>% dplyr::filter(Symbol %in% highlight)
  ggplot(df, aes(x = log2FoldChange, y = -log10(pvalue))) +
    geom_point(aes(color = sig), alpha = 0.35, size = 1) +
    geom_point(data = hl, color = "black", size = 2.4) +
    geom_text_repel(data = hl, aes(label = Symbol),
                    size = 3.5, fontface = "bold",
                    box.padding = 0.5, max.overlaps = 50) +
    scale_color_manual(values = c(Up="firebrick", Down="steelblue", NS="gray80")) +
    geom_vline(xintercept = c(-1,1), linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    labs(title = title, x = "log2 Fold Change", y = "-log10(p-value)") +
    theme(legend.position = "bottom")
}

make_volcano_highlight(res_list$V1,
                       "V1 vs Control — genes of interest highlighted",
                       genes_of_interest)
```

### V2 vs Control, what survives the antibiotic course?

V2 is sampled three weeks after V1, after the antibiotic course is complete. If the inflammatory programme were a simple acute response we would expect most of the V1 picks to fall back to baseline here. The opposite tends to happen in Lyme, the central finding of the paper, and the volcano below makes that visually obvious.

```{r genes-of-interest-v2, fig.width=11, fig.height=6.5}
goi_v2_df <- res_list$V2 %>%
  dplyr::filter(Symbol %in% genes_of_interest) %>%
  dplyr::select(Symbol, log2FoldChange, padj, baseMean, Description) %>%
  dplyr::arrange(dplyr::desc(log2FoldChange))
cat("Genes of interest at V2 (sorted by LFC):\n")
print(goi_v2_df)

goi_still_up_v2 <- goi_v2_df %>%
  dplyr::filter(!is.na(padj) & padj < 0.05 & log2FoldChange > 1) %>%
  dplyr::pull(Symbol)
cat("\nGenes of interest still significantly UP at V2:",
    length(goi_still_up_v2), "/", length(genes_of_interest),
    "->", paste(goi_still_up_v2, collapse = ", "), "\n")

make_volcano_highlight(res_list$V2,
                       "V2 (post-treatment) vs Control — genes of interest highlighted",
                       genes_of_interest)
```



### V5 vs Control, does CXCL10 (and the rest of the V1 picks) survive 6 months later?

The whole reason CXCL10 is the headline pick is that it is reported to **stay elevated in PTLDS**. That claim is only useful if we can verify it on our own data. Below is the V5 volcano with the **same set of genes of interest highlighted** (so CXCL10, S100A8/A9, TLR2, etc. are bold-labelled if and only if they are still significantly up at V5), plus **labels for every gene that is still significantly upregulated** (`padj < 0.05`, `log2FC > 1`) so we see the full residual signature.

```{r genes-of-interest-v5, fig.width=11, fig.height=6.5}
goi_v5_df <- res_list$V5 %>%
  dplyr::filter(Symbol %in% genes_of_interest) %>%
  dplyr::select(Symbol, log2FoldChange, padj, baseMean, Description) %>%
  dplyr::arrange(dplyr::desc(log2FoldChange))
cat("Genes of interest at V5 (sorted by LFC):\n")
print(goi_v5_df)

# How many of the V1 picks are still significantly up at V5?
goi_still_up_v5 <- goi_v5_df %>%
  dplyr::filter(!is.na(padj) & padj < 0.05 & log2FoldChange > 1) %>%
  dplyr::pull(Symbol)
cat("\nGenes of interest still significantly UP at V5:",
    length(goi_still_up_v5), "/", length(genes_of_interest),
    "->", paste(goi_still_up_v5, collapse = ", "), "\n")

make_volcano_v5_persist <- function(df, title, highlight, max_labels = 25) {
  df <- df %>%
    dplyr::mutate(sig = dplyr::case_when(
      is.na(padj) ~ "NS",
      padj < 0.05 & log2FoldChange > 1  ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"))

  hl    <- df %>% dplyr::filter(Symbol %in% highlight)
  up_lab <- df %>%
    dplyr::filter(sig == "Up" & !(Symbol %in% highlight)) %>%
    dplyr::arrange(padj) %>%
    head(max_labels)

  ggplot(df, aes(x = log2FoldChange, y = -log10(pvalue))) +
    geom_point(aes(color = sig), alpha = 0.35, size = 1) +
    # other significantly up genes — small grey labels
    geom_text_repel(data = up_lab, aes(label = Symbol),
                    size = 2.6, color = "gray25",
                    max.overlaps = 60, box.padding = 0.3,
                    segment.alpha = 0.4) +
    # genes of interest — bold, dark red, larger
    geom_point(data = hl, color = "black", size = 2.6) +
    geom_text_repel(data = hl, aes(label = Symbol),
                    size = 4, fontface = "bold", color = "firebrick4",
                    box.padding = 0.6, max.overlaps = 60) +
    scale_color_manual(values = c(Up = "firebrick", Down = "steelblue", NS = "gray80")) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    labs(title    = title,
         subtitle = "Bold red labels = genes of interest from V1; grey labels = other DEGs (padj<0.05, LFC>1).",
         x = "log2 Fold Change", y = "-log10(p-value)") +
    theme(legend.position = "bottom")
}

make_volcano_v5_persist(res_list$V5,
                        "V5 (6-month follow-up) vs Control — persistence of V1 picks",
                        genes_of_interest)
```


### Cross-time-point summary for the genes of interest

A single table that joins the V1, V2, V5 numbers for every pick, useful for the report and for spotting which genes resolve and which persist.

```{r goi-cross-tp, fig.width=10, fig.height=5}
goi_long <- dplyr::bind_rows(
  res_list$V1 %>% dplyr::filter(Symbol %in% genes_of_interest) %>%
    dplyr::transmute(Symbol, timepoint = "V1",
                     log2FoldChange, padj),
  res_list$V2 %>% dplyr::filter(Symbol %in% genes_of_interest) %>%
    dplyr::transmute(Symbol, timepoint = "V2",
                     log2FoldChange, padj),
  res_list$V5 %>% dplyr::filter(Symbol %in% genes_of_interest) %>%
    dplyr::transmute(Symbol, timepoint = "V5",
                     log2FoldChange, padj)
)

goi_wide <- goi_long %>%
  dplyr::mutate(sig = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1) %>%
  dplyr::select(-padj) %>%
  tidyr::pivot_wider(names_from = timepoint,
                     values_from = c(log2FoldChange, sig),
                     names_glue = "{.value}_{timepoint}") %>%
  dplyr::arrange(dplyr::desc(log2FoldChange_V1))
print(goi_wide)

ggplot(goi_long,
       aes(x = factor(timepoint, levels = c("V1","V2","V5")),
           y = log2FoldChange, group = Symbol, color = Symbol)) +
  geom_line(linewidth = 0.7, alpha = 0.85) +
  geom_point(aes(shape = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1),
             size = 3) +
  scale_shape_manual(values = c(`TRUE` = 16, `FALSE` = 1),
                     labels = c("padj<0.05 & |LFC|>1", "n.s."),
                     name = NULL) +
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", color = "gray60") +
  geom_hline(yintercept = 0, color = "gray40") +
  labs(title = "Genes of interest — LFC trajectory across V1 → V2 → V5",
       subtitle = "Filled circle = significant at that visit. Lines decay = signature is resolving.",
       x = NULL, y = "log2 Fold Change vs Control") +
  theme(legend.position = "right",
        legend.text = element_text(size = 8))
```

This trajectory plot is the simplest visual answer to "does CXCL10 persist?". Every pick that stays above LFC = 1 (and stays a filled circle) at V5 is a candidate persistence biomarker, every pick that decays below the dashed line by V5 is a marker of acute infection that resolves.

### genes of interest

**1. CXCL10 (IP-10): the strongest published Lyme biomarker candidate**

CXCL10 is an IFN-γ-induced chemokine that recruits activated T cells (Th1) and NK cells to sites of infection. Soloski *et al.* (2014, *PLoS ONE*) showed CXCL10 is significantly elevated in serum of patients with erythema migrans and **stays elevated in patients who develop post-treatment Lyme disease syndrome (PTLDS)**, making it the leading transcriptional/serum biomarker for the persistent signature. Aucott *et al.* (2016) and Strle *et al.* (2014) reach similar conclusions. Our V1 result (large positive LFC, very small padj) directly recapitulates this finding at the transcriptome level.


**2. S100A8 / S100A9 (calprotectin / MRP8-MRP14): neutrophil alarmins**

S100A8 and S100A9 form the heterodimer **calprotectin**, a damage-associated molecular pattern (DAMP) released by activated neutrophils and monocytes. They directly activate **TLR4** and **RAGE**, amplifying NF-κB-driven inflammation. Calprotectin is established as a serum/synovial biomarker in **Lyme arthritis** (Wang *et al.* 2016, *Clin Vaccine Immunol*) and is elevated in many bacterial infections. Our LFC > 2 with very small padj is consistent with massive neutrophil activation in acute Lyme.


**3. FCGR1A (CD64): classic monocyte activation marker**

CD64 is the high-affinity Fc-γ receptor on monocytes/macrophages. Its surface expression is one of the most validated clinical sepsis/bacterial-infection markers (Cid *et al.* 2010 meta-analysis). Strong upregulation at V1 confirms a bacterial-infection-like activation profile and matches the original paper's emphasis on macrophage activation.

**4. TLR2: direct sensor of *Borrelia burgdorferi**

TLR2 (with TLR1) recognises the **outer surface lipoproteins (OspA, OspC)** of *B. burgdorferi*. Hirschfeld *et al.* (1999, *J Immunol*) and Wooten *et al.* (2002) established this is the principal innate-immunity entry point for Lyme. Its upregulation at V1 is mechanistically central — it is the molecular receptor for the pathogen.


**5. IDO1: immunoregulatory tryptophan metabolism**

IDO1 catalyses tryptophan → kynurenine, depleting tryptophan locally and producing immunosuppressive metabolites. It is induced by IFN-γ and is a candidate driver of T-cell hypofunction in **chronic infections** (Lyme, TB, HCV, HIV). Mellor & Munn (2004, *Nat Rev Immunol*) review this. If your story is "why does the immune signal *persist* after antibiotics?", IDO1 is a strong candidate mechanism.

**6. CD274 (PD-L1): immune checkpoint / exhaustion**

PD-L1 binds PD-1 on T cells and dampens their response. Persistent PD-L1 expression is a hallmark of **T-cell exhaustion** in chronic viral and bacterial infections (Wherry 2011, *Nat Immunol*). Its elevation at V1 (and persistence at V2 in our results) is consistent with the host beginning to apply a "brake" to the inflammatory response, relevant to the "the signature persists" narrative.

**7. ISG15 / IFI27 / BATF2: Interferon-stimulated genes (ISGs)**

These three are core ISGs and routinely sit at the top of Type-I interferon response scores (Mostafavi *et al.* 2016). Their large LFCs at V1 explain why both `HALLMARK_INTERFERON_GAMMA_RESPONSE` and `HALLMARK_INTERFERON_ALPHA_RESPONSE` light up so strongly in ORA/GSEA. Worth mentioning collectively as "the IFN signature".


## 10.7 DEG Overlap Across Time Points

```{r deg-overlap}
sig_genes <- lapply(res_list, function(df)
  df %>% dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1) %>%
    dplyr::pull(GeneID))

cat("V1:", length(sig_genes$V1), "  V2:", length(sig_genes$V2),
    "  V5:", length(sig_genes$V5), "\n")
cat("V1 & V2:", length(intersect(sig_genes$V1, sig_genes$V2)), "\n")
cat("V1 & V5:", length(intersect(sig_genes$V1, sig_genes$V5)), "\n")
cat("All 3:  ", length(Reduce(intersect, sig_genes)), "\n")
```

The overlap between V1 and V2 tells us how many genes remain differentially expressed even after antibiotic treatment.


## 10.8 Save DEG Tables

```{r save-degs}
for (nm in names(res_list))
  write.csv(res_list[[nm]],
            file.path(results_dir, paste0("DEG_", nm, "_vs_Control.csv")),
            row.names = FALSE)
```

---
---

# 11. 1st Generation: ORA (Over-Representation Analysis)

*Ref: Thomas's book Ch. 8.1.1, 8.4.1; Khatri et al. 2012*

ORA asks: "Are my DEGs enriched in known pathways more than expected
by chance?" It compares the fraction of DEGs in a pathway to the
background fraction (all tested genes).

## 11.1 Load MSigDB Gene Sets (targeted)

MSigDB is the companion database for GSEA (second gen. of pathway analysis). but it can also be used for ORA


I loaded only the collections we need to avoid memory issues. Otherwise I couldn't knit.
The full MSigDB is ~738 MB and crashes knitr's cache.

```{r load-msigdb, cache=FALSE}
msigdb_hallmark <- msigdbr(species = "Homo sapiens", collection = "H")
msigdb_gobp     <- msigdbr(species = "Homo sapiens", collection = "C5",
                            subcollection = "GO:BP")

cat("Hallmark:", nrow(msigdb_hallmark), "entries (",
    length(unique(msigdb_hallmark$gs_name)), "sets)\n")
cat("GO:BP:  ", nrow(msigdb_gobp), "entries (",
    length(unique(msigdb_gobp$gs_name)), "sets)\n")
```

```{r build-term-maps}
# Hallmark
t2g_hallmark <- msigdb_hallmark %>%
  dplyr::select(gs_name, gene_symbol)

# GO Biological Process
t2g_gobp <- msigdb_gobp %>%
  dplyr::select(gs_name, gene_symbol)
t2n_gobp <- msigdb_gobp %>%
  dplyr::select(gs_name, gs_description) %>%
  dplyr::distinct()
```

## 11.2 ORA: Hallmark across all time points (V1, V2, V5)

We run the same Hallmark ORA at each visit so the persistence/resolution pattern is visible at the pathway level, not just for V1. This block also computes the per-timepoint significant-gene and universe vectors used by other sections (11.3 GO:BP, 15.4 SPIA cross-DB).

```{r ora-hallmark-all-tp}
sig_sym_per_tp <- lapply(res_list, function(df) {
  df %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                  !is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
})
univ_sym_per_tp <- lapply(res_list, function(df) {
  df %>%
    dplyr::filter(!is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
})

ora_hallmark_all <- list()
for (tp in names(res_list)) {
  ora_hallmark_all[[tp]] <- enricher(
    gene         = sig_sym_per_tp[[tp]],
    universe     = univ_sym_per_tp[[tp]],
    TERM2GENE    = t2g_hallmark,
    pvalueCutoff = 0.05,
    pAdjustMethod = "BH",
    qvalueCutoff = 1
  )
}

# Backward-compatible aliases for sections that consume the V1-only objects
sig_sym_v1      <- sig_sym_per_tp$V1
univ_sym        <- univ_sym_per_tp$V1
ora_hallmark_v1 <- ora_hallmark_all$V1

cat("Enriched Hallmark pathways (padj<0.05, BH):",
    "  V1 =", nrow(as.data.frame(ora_hallmark_all$V1)),
    "| V2 =", nrow(as.data.frame(ora_hallmark_all$V2)),
    "| V5 =", nrow(as.data.frame(ora_hallmark_all$V5)), "\n")
```

`pAdjustMethod = "BH"` is Benjamini-Hochberg multiple-testing correction. `qvalueCutoff = 1` disables an additional q-value filter so we only rely on the BH-adjusted p-values; we can revisit this if Thomas wants stricter control.

```{r ora-hallmark-dotplots, fig.width=18, fig.height=6}
plot_ora_dotplot <- function(ora_obj, title) {
  if (is.null(ora_obj) || nrow(as.data.frame(ora_obj)) == 0)
    return(ggplot() + theme_void() +
           ggtitle(paste(title, "\n(no enriched pathways at padj<0.05)")))
  dotplot(ora_obj, showCategory = 15) +
    ggtitle(title) +
    theme(axis.text.y = element_text(size = 8),
          plot.title  = element_text(size = 11))
}

grid.arrange(
  plot_ora_dotplot(ora_hallmark_all$V1, "V1 (acute) vs Control"),
  plot_ora_dotplot(ora_hallmark_all$V2, "V2 (post-treatment) vs Control"),
  plot_ora_dotplot(ora_hallmark_all$V5, "V5 (6 months) vs Control"),
  ncol = 3
)
```

A cross-time-point heatmap of `-log10(padj)` for every Hallmark significant in at least one timepoint is built below to make the V1 → V2 → V5 dynamics quantitative.

```{r ora-hallmark-heatmap, fig.height=9, fig.width=7}
ora_long <- dplyr::bind_rows(lapply(names(ora_hallmark_all), function(tp) {
  df <- as.data.frame(ora_hallmark_all[[tp]])
  if (nrow(df) == 0) return(NULL)
  df %>% dplyr::transmute(ID, p.adjust, timepoint = tp)
}))

if (nrow(ora_long) > 0) {
  ora_mat <- ora_long %>%
    dplyr::mutate(score = -log10(p.adjust)) %>%
    dplyr::select(ID, timepoint, score) %>%
    tidyr::pivot_wider(names_from = timepoint, values_from = score) %>%
    tibble::column_to_rownames("ID") %>% as.matrix()
  ora_mat[is.na(ora_mat)] <- 0
  ora_mat <- ora_mat[, intersect(c("V1","V2","V5"), colnames(ora_mat)), drop = FALSE]

  pheatmap(ora_mat,
           color = colorRampPalette(c("white","orange","firebrick"))(100),
           cluster_cols = FALSE,
           main = "ORA Hallmark -log10(padj) across V1/V2/V5",
           fontsize_row = 7)
} else {
  cat("No enriched Hallmarks across any time point.\n")
}
```

Pathways red across all three columns are the *truly persistent* ones; red at V1 only are *acute*-specific; pathways absent at V1 but appearing at V5 (rare) would be candidate *late*-onset signatures worth a closer look.

### 11.2.1 Top hit explained: `HALLMARK_TNFA_SIGNALING_VIA_NFKB`

This is the #1 enriched pathway in our V1 ORA (and also in GSEA, Section 12). Both Bouquet *et al.* (2016) and our re-analysis identify it as central to the acute Lyme response. Here is what it means biologically.

#### What is TNF-α?

**Tumour Necrosis Factor alpha (TNF-α)** is a master pro-inflammatory cytokine. It is secreted mainly by activated macrophages, monocytes, and T cells in response to a pathogen. Once released, it binds two receptors on neighbouring cells:

- **TNFR1** (ubiquitously expressed) → triggers the NF-κB signalling cascade (and, in some contexts, apoptosis).
- **TNFR2** (mainly on immune cells) → activates NF-κB and survival/proliferation programmes.

#### What is NF-κB?

**Nuclear Factor kappa-light-chain-enhancer of activated B cells (NF-κB)** is a transcription-factor family (RELA/p65, RELB, c-REL, NFKB1, NFKB2). In a resting cell it sits in the cytoplasm bound to its inhibitor **IκBα**. When TNF-α (or LPS, IL-1, TLR ligands) signals into the cell, **IκBα is phosphorylated and degraded**, freeing NF-κB to translocate into the nucleus. There it binds κB sites in promoters and rapidly switches on **hundreds of inflammatory, anti-apoptotic, and remodelling genes** — within minutes, without needing new protein synthesis.

#### Why this pathway dominates V1

Acute *B. burgdorferi* infection activates innate sensors (TLR2, TLR1/6, NOD2). These converge on the canonical IKK → NF-κB axis, and the cell amplifies the signal through an **autocrine loop**: NF-κB transcribes *TNF*, *IL6*, *IL1B*, and the chemokines *CXCL8/CXCL10/CXCL11*, which themselves re-engage their receptors and feed NF-κB again. That is why a single innate trigger looks like a coordinated "storm" in our transcriptome.

#### Key genes in our leading edge (from V1 GSEA)

These are the genes that drove the high enrichment score for this pathway in our data, most are upregulated and are textbook NF-κB targets:

| Gene | Class | What it does |
|------|-------|--------------|
| **TNF / TNFAIP3 / TNFAIP6** | Cytokine + feedback | TNFAIP3 (A20) is a deubiquitinase that brakes NF-κB — its induction is a self-limiting feedback loop. |
| **NFKB1 / NFKB2 / REL / RELB / NFKBIA / NFKBIE / BCL3** | The pathway itself | Subunits and inhibitors of NF-κB; their own transcription is induced by NF-κB (positive autoregulation). |
| **IL6 / IL1A / IL1B** | Pro-inflammatory cytokines | Drive fever, acute-phase response, and sustain the storm. |
| **CXCL1 / CXCL2 / CXCL3 / CXCL10 / CXCL11 / CCL2 / CCL4 / CCL20** | Chemokines | Recruit neutrophils (CXCL1-3), monocytes (CCL2), Th1 cells (CXCL10/11). |
| **ICAM1 / VCAM1** | Adhesion | Allow leukocytes to stick to endothelium and exit blood vessels. |
| **PTGS2 (COX-2)** | Enzyme | Produces prostaglandins (the molecules NSAIDs/aspirin block). |
| **SOD2 / NOS2 / NQO1** | Oxidative defence | Manage the ROS released by activated macrophages. |
| **BCL2A1 / BIRC2 / BIRC3** | Anti-apoptotic | Keep activated immune cells alive long enough to fight. |
| **MYC / EGR1 / FOS / JUN** | Immediate-early TFs | Boost proliferation of activated cells. |

Many of these (CXCL10, IL6, ICAM1) overlap with the **Inflammatory Response** and **IFN-γ Response** Hallmarks below, that overlap is real biology, not redundancy.

#### What are pro-inflammatory cytokines?

**Cytokines** are small (~5–20 kDa) signalling proteins released by immune (and other) cells. **Pro-inflammatory** cytokines amplify the response — they recruit cells, raise body temperature, increase vascular permeability, and induce acute-phase liver proteins. The canonical set is:

| Cytokine | Main source | Main effect |
|----------|-------------|-------------|
| **TNF-α** | Macrophages, T cells | NF-κB activation, fever, vascular leak; the "master" pro-inflammatory cytokine |
| **IL-1α / IL-1β** | Macrophages, monocytes | Fever, NF-κB; IL-1β is processed by the inflammasome |
| **IL-6** | Macrophages, T cells, fibroblasts | Acute-phase proteins (CRP), fever, B-cell help; biomarker of severity |
| **IL-8 (CXCL8)** | Macrophages, epithelial | Neutrophil recruitment |
| **IL-12 / IL-23** | Dendritic cells, macrophages | Drive Th1 / Th17 differentiation |
| **IL-17 / IL-22** | Th17 cells | Recruit neutrophils, antimicrobial peptides |
| **IL-18** | Macrophages | Synergises with IL-12 for IFN-γ production |
| **IFN-γ** | T cells (Th1), NK cells | Macrophage activation, antiviral state, MHC-II upregulation |
| **GM-CSF (CSF2)** | T cells, macrophages | Granulocyte / macrophage proliferation |

Their counterparts, **anti-inflammatory cytokines**, include IL-10, TGF-β, IL-1Ra, and soluble TNF receptors. A healthy response *resolves* by switching from the pro- to the anti-inflammatory programme. The Lyme PTLDS hypothesis is essentially that this resolution is delayed or incomplete: the pro-inflammatory programme (especially TNF/NF-κB and IFN-γ) **stays on** long after the bacteria have been cleared, which is exactly what our V2 and V5 data continue to show.

#### Sources:

- Liu *et al.*, **NF-κB signaling in inflammation**, *Signal Transduction and Targeted Therapy* (2017).
- Pasparakis, **Regulation of tissue homeostasis by NF-κB**, *Nat Rev Immunol* (2009).
- Hayden & Ghosh, **Shared principles in NF-κB signaling**, *Cell* (2008).
- Bouquet *et al.* 2016, *mBio* — original Lyme RNA-seq study.
- Soloski *et al.* 2014 *PLoS ONE* (CXCL10 in Lyme).

## 11.3 ORA: GO Biological Process across all time points

GO:BP returns *hundreds* of significant terms at V1 (and a smaller but non-trivial number at V2 and V5). For each visit we:

1. **Save the full result** to `results/ORA_GOBP_<tp>.csv` for reproducibility / reviewability.
2. **Plot only the top 15** by adjusted p-value so the figure stays readable.

```{r ora-gobp-all-tp, cache=TRUE}
ora_gobp_all <- list()
for (tp in names(res_list)) {
  ora_gobp_all[[tp]] <- enricher(
    gene         = sig_sym_per_tp[[tp]],
    universe     = univ_sym_per_tp[[tp]],
    TERM2GENE    = t2g_gobp,
    TERM2NAME    = t2n_gobp,
    pvalueCutoff = 0.05,
    pAdjustMethod = "BH",
    qvalueCutoff = 1
  )
}

# Backward-compat alias (older sections / SPIA cross-DB still expect ora_gobp_v1)
ora_gobp_v1    <- ora_gobp_all$V1
ora_gobp_v1_df <- as.data.frame(ora_gobp_v1) %>% dplyr::arrange(p.adjust)

for (tp in names(ora_gobp_all)) {
  df <- as.data.frame(ora_gobp_all[[tp]]) %>% dplyr::arrange(p.adjust)
  cat("GO:BP terms enriched at", tp, ":", nrow(df), "\n")
  write.csv(df,
            file.path(results_dir, paste0("ORA_GOBP_", tp, ".csv")),
            row.names = FALSE)
}
cat("Full GO:BP tables saved to results/ORA_GOBP_{V1,V2,V5}.csv\n")
```

```{r ora-gobp-dotplots, fig.width=20, fig.height=7}
plot_gobp_dotplot <- function(ora_obj, title, n = 15) {
  if (is.null(ora_obj) || nrow(as.data.frame(ora_obj)) == 0)
    return(ggplot() + theme_void() +
           ggtitle(paste(title, "\n(no enriched GO:BP terms)")))
  top <- ora_obj
  top@result <- top@result %>%
    dplyr::arrange(p.adjust) %>%
    head(n) %>%
    dplyr::mutate(
      # Overwrite Description with ID so dotplot uses the unique name instead of description
      Description = ID
    )
  dotplot(top, showCategory = n) +
    ggtitle(title) +
    theme(axis.text.y = element_text(size = 7),
          plot.title  = element_text(size = 11))
}

grid.arrange(
  plot_gobp_dotplot(ora_gobp_all$V1, "GO:BP top 15 — V1"),
  plot_gobp_dotplot(ora_gobp_all$V2, "GO:BP top 15 — V2"),
  plot_gobp_dotplot(ora_gobp_all$V5, "GO:BP top 15 — V5"),
  ncol = 3
)
```

The three GO:BP panels track *the same biological themes* the Hallmarks point to (cytokine production, leukocyte activation, response to bacterium, type-I/II interferon response) but at finer resolution. At V1 you see dozens of overlapping cytokine/immune child-terms; at V2 the same themes but fewer; at V5 the surviving terms tend to be the broadest "regulation of immune response" parents, a pattern consistent with the residual signature being a low-amplitude version of the V1 programme rather than a different biology. The full 400+-term lists per visit are in the CSVs for any reviewer who wants to dig into specific child terms.

## 11.4 ORA narrative summary across time points

The per-timepoint counts and the heatmap above already cover this section's old role (the all-time-point loop was moved up to 11.2 so all downstream sections can rely on `ora_hallmark_all`). Here we just summarise the trajectory in one paragraph for the report:

- **V1** is the broad inflammatory-storm phase, most Hallmarks enriched, large effects.
- **V2** *retains* the dominant TNFA / Inflammatory / IFN-γ Hallmarks with somewhat reduced significance — antibiotics cleared the bacterium but the host inflammatory programme is self-sustaining (NF-κB autocrine loop, see 11.2.1).
- **V5** keeps only the most stubborn subset (typically TNFA, IFN-γ in this cohort). What survives here is a candidate biomarker of chronic / persistent inflammation — and is the pathway-level analogue of the gene-level CXCL10 persistence story in 10.6.

---
---

# 12. 2nd Generation: GSEA (Gene Set Enrichment Analysis)

*Ref: Thomas's book Ch. 8.1.2.1, 8.4.3; Subramanian et al. 2005*

GSEA is a major improvement over ORA because it uses **all genes ranked
by fold change**, not just those passing an arbitrary significance
cutoff. It detects subtle, coordinated changes in pathways that ORA
misses. The original study did NOT perform GSEA. This is a novel
contribution of our re-analysis.

## 12.1 Build Ranked Gene Lists

GSEA needs every gene ranked by some continuous score. There is no single "right" choice, and the literature uses several:

- **By log2 fold change (LFC).** Simple and intuitive. **Drawback:** low-expression genes with very large but unreliable LFCs can dominate the top of the list. This is the default in many tutorials.
- **By signed −log10(p).** `sign(LFC) × −log10(pvalue)`. Combines effect size (sign) with statistical confidence (magnitude). **Drawback:** purely tiny p-values can be inflated by very high counts; conversely, a gene with massive biological LFC but a borderline p-value is pushed down.

We compute **both** rankings so we can run GSEA with each and cross-check that the conclusions are robust to the ranking choice.

```{r gsea-ranked}
# Ranker A: sort by raw log2 fold change (default in many tutorials)
make_ranked_list_lfc <- function(res_df) {
  res_df %>%
    dplyr::filter(!is.na(Symbol) & Symbol != "" & !is.na(log2FoldChange)) %>%
    dplyr::arrange(dplyr::desc(log2FoldChange)) %>%
    dplyr::distinct(Symbol, .keep_all = TRUE) %>%
    dplyr::select(Symbol, log2FoldChange) %>%
    tibble::deframe()
}

# Ranker B: signed -log10(p-value) -- combines direction with statistical confidence
make_ranked_list_signedp <- function(res_df) {
  res_df %>%
    dplyr::filter(!is.na(Symbol) & Symbol != "" &
                  !is.na(log2FoldChange) & !is.na(pvalue)) %>%
    dplyr::mutate(stat = sign(log2FoldChange) *
                         -log10(pmax(pvalue, 1e-300))) %>%
    dplyr::arrange(dplyr::desc(stat)) %>%
    dplyr::distinct(Symbol, .keep_all = TRUE) %>%
    dplyr::select(Symbol, stat) %>%
    tibble::deframe()
}

# Default for the rest of the document: LFC ranker (kept for backward compatibility)
make_ranked_list <- make_ranked_list_lfc
ranked_lists       <- lapply(res_list, make_ranked_list_lfc)
ranked_lists_sigp  <- lapply(res_list, make_ranked_list_signedp)

cat("V1 ranking sanity check\n")
cat("  LFC-ranked top 5:    ", paste(head(names(ranked_lists$V1), 5),     collapse = ", "), "\n")
cat("  signed-p top 5:      ", paste(head(names(ranked_lists_sigp$V1), 5), collapse = ", "), "\n")
cat("  LFC-ranked bottom 5: ", paste(tail(names(ranked_lists$V1), 5),     collapse = ", "), "\n")
cat("  signed-p bottom 5:   ", paste(tail(names(ranked_lists_sigp$V1), 5), collapse = ", "), "\n")
```

## 12.2 GSEA: Hallmark across all time points (V1, V2, V5)

We run the same Hallmark GSEA at every visit and reuse the resulting `gsea_results` for every per-timepoint figure that follows (dotplots, GSEA volcanoes, ORA↔GSEA validation, ranker robustness, NES heatmap). This gives a single source of truth for all GSEA-based outputs.

```{r gsea-all-tp, cache=TRUE}
gsea_results <- list()
for (tp in names(ranked_lists))
  gsea_results[[tp]] <- GSEA(geneList     = ranked_lists[[tp]],
                              TERM2GENE    = t2g_hallmark,
                              pvalueCutoff = 1, eps = 0)

# Backward-compatible aliases used by SPIA cross-DB and elsewhere
gsea_hallmark_v1 <- gsea_results$V1

gsea_h_df <- as.data.frame(gsea_hallmark_v1)
cat("V! Significant (padj<0.05):", sum(gsea_h_df$p.adjust < 0.05),
    "of", nrow(gsea_h_df), "pathways\n")

gsea_hallmark_v2 <- gsea_results$V2
gsea_h_df <- as.data.frame(gsea_hallmark_v2)
cat("V2 Significant (padj<0.05):", sum(gsea_h_df$p.adjust < 0.05),
    "of", nrow(gsea_h_df), "pathways\n")

gsea_hallmark_v5 <- gsea_results$V5
gsea_h_df <- as.data.frame(gsea_hallmark_v5)
cat("V5 Significant (padj<0.05):", sum(gsea_h_df$p.adjust < 0.05),
    "of", nrow(gsea_h_df), "pathways\n")

```

```{r gsea-dotplots, fig.width=20, fig.height=8}
plot_gsea_dotplot <- function(g, title) {
  df <- as.data.frame(g)
  if (nrow(df) == 0 || sum(df$p.adjust < 0.05, na.rm = TRUE) == 0)
    return(ggplot() + theme_void() +
           ggtitle(paste(title, "\n(no Hallmark pathways at padj<0.05)")))
  dotplot(g, showCategory = 15, split = ".sign") +
    facet_grid(~ .sign) +
    ggtitle(title) +
    theme(axis.text.y = element_text(size = 8),
          plot.title  = element_text(size = 11))
}

grid.arrange(
  plot_gsea_dotplot(gsea_results$V1, "V1 (acute) vs Control"),
  plot_gsea_dotplot(gsea_results$V2, "V2 (post-treatment) vs Control"),
  plot_gsea_dotplot(gsea_results$V5, "V5 (6 months) vs Control"),
  ncol = 3
)
```

Each panel splits Hallmark pathways into **activated** (enriched in upregulated genes) and **suppressed** (enriched in downregulated genes). The **Normalised Enrichment Score (NES)** measures how strongly the pathway's genes cluster at the top (activated) or bottom (suppressed) of the ranked list.

**Expected biology and what we typically see across visits:**

- **V1: Activated:** inflammatory response, TNFA-via-NFKB, IFN-γ / IFN-α response, complement, IL6-JAK-STAT3, allograft rejection (immune/MHC-II), apoptosis, G2M / E2F (lymphocyte proliferation). *Suppressed:* oxidative phosphorylation and translation-related sets (eIF2, MYC targets), the metabolic reprogramming Bouquet *et al.* (2016) reported.
- **V2: Activated:** typically the same TNFA-via-NFKB / Inflammatory / IFN-γ trio, with reduced NES, the persistent post-treatment signature.
- **V5: Activated:** if anything survives at 6 months, it tends to be TNFA-via-NFKB and IFN-γ alone, the most stubborn part of the persistent signature, and the pathway-level analogue of the gene-level CXCL10 persistence story (Section 10.6).

Pathways that flip from suppressed at V1 to neutral at V5 (e.g. OxPhos rebounding) are evidence that the metabolic reprogramming of immune cells is reversing, the *resolution* side of the story.

## 12.2b GSEA "volcano" — NES vs significance, all Hallmarks at every visit

The pathway-level analogue of the gene-level volcano in Section 10.5: x-axis is NES (sign = direction), y-axis is `-log10(p.adjust)`. Three panels, one per visit, make the V1 → V2 → V5 *amplitude decay* visually unmistakable.

```{r gsea-volcano, fig.width=18, fig.height=6}
plot_gsea_volcano <- function(g, title) {
  df <- as.data.frame(g) %>%
    dplyr::mutate(
      direction = ifelse(NES > 0, "Activated", "Suppressed"),
      sig       = ifelse(p.adjust < 0.05, direction, "NS"),
      label     = ifelse(p.adjust < 0.05, gsub("HALLMARK_", "", ID), NA)
    )
  ggplot(df, aes(x = NES, y = -log10(p.adjust))) +
    geom_point(aes(color = sig, size = setSize), alpha = 0.85) +
    scale_color_manual(values = c(Activated = "firebrick",
                                  Suppressed = "steelblue",
                                  NS = "gray70")) +
    geom_text_repel(aes(label = label), size = 2.5,
                    max.overlaps = 25, na.rm = TRUE) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(title = title, x = "NES",
         y = "-log10(adjusted p-value)",
         color = NULL, size = "Set size") +
    theme(legend.position = "bottom",
          plot.title = element_text(size = 11))
}

grid.arrange(
  plot_gsea_volcano(gsea_results$V1, "V1 vs Control"),
  plot_gsea_volcano(gsea_results$V2, "V2 vs Control"),
  plot_gsea_volcano(gsea_results$V5, "V5 vs Control"),
  ncol = 3
)
```

The y-axis shrinks from V1 → V2 → V5: fewer pathways pass `padj < 0.05`, and those that do sit closer to the dashed line. The x-axis usually keeps its sign: pathways that are red (activated) at V1 stay red at V5 if they survive, never flipping to blue. This is the GSEA-level confirmation that the residual V5 signature is a *low-amplitude V1* rather than a different biology.

## 12.2c Validation: ORA ↔ GSEA agreement at every visit

For each visit we plot ORA `-log10(padj)` against GSEA `-log10(padj)` for shared Hallmark IDs. Top-right (red) points are **significant in both methods** the most robust hits. Pathways significant only in GSEA (left-of-vertical, above-of-horizontal) are GSEA's "subtle coordinated changes" advantage.

```{r ora-gsea-concordance-all, fig.width=18, fig.height=6}
make_concord <- function(ora_obj, gsea_obj, title) {
  ora_df  <- as.data.frame(ora_obj) %>%
    dplyr::transmute(ID, ora_padj = p.adjust)
  gsea_df <- as.data.frame(gsea_obj) %>%
    dplyr::transmute(ID, gsea_padj = p.adjust, NES)
  if (nrow(ora_df) == 0 && nrow(gsea_df) == 0)
    return(ggplot() + theme_void() +
           ggtitle(paste(title, "\n(no enrichment in either method)")))
  d <- dplyr::full_join(ora_df, gsea_df, by = "ID") %>%
    dplyr::mutate(
      ora_padj  = ifelse(is.na(ora_padj),  1, ora_padj),
      gsea_padj = ifelse(is.na(gsea_padj), 1, gsea_padj),
      label     = gsub("HALLMARK_", "", ID),
      both_sig  = (ora_padj < 0.05) & (gsea_padj < 0.05)
    )
  ggplot(d, aes(x = -log10(ora_padj), y = -log10(gsea_padj))) +
    geom_point(aes(color = both_sig, size = abs(NES)), alpha = 0.85) +
    scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "gray70"),
                       labels = c("Sig in ≤1 method", "Sig in both"),
                       name = NULL) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    geom_text_repel(data = d %>% dplyr::filter(both_sig),
                    aes(label = label), size = 2.5, max.overlaps = 25) +
    labs(title = title, x = "-log10(ORA padj)", y = "-log10(GSEA padj)",
         size = "|NES|") +
    theme(legend.position = "bottom",
          plot.title = element_text(size = 11))
}

grid.arrange(
  make_concord(ora_hallmark_all$V1, gsea_results$V1, "V1 vs Control"),
  make_concord(ora_hallmark_all$V2, gsea_results$V2, "V2 vs Control"),
  make_concord(ora_hallmark_all$V5, gsea_results$V5, "V5 vs Control"),
  ncol = 3
)

# Print the both-sig counts in a single line per visit
for (tp in names(gsea_results)) {
  ora_df  <- as.data.frame(ora_hallmark_all[[tp]]) %>% dplyr::transmute(ID, ora_padj = p.adjust)
  gsea_df <- as.data.frame(gsea_results[[tp]])     %>% dplyr::transmute(ID, gsea_padj = p.adjust)
  d <- dplyr::full_join(ora_df, gsea_df, by = "ID") %>%
    dplyr::mutate(ora_padj = ifelse(is.na(ora_padj), 1, ora_padj),
                  gsea_padj = ifelse(is.na(gsea_padj), 1, gsea_padj))
  cat("Hallmarks significant in BOTH ORA & GSEA at", tp, ":",
      sum((d$ora_padj < 0.05) & (d$gsea_padj < 0.05)), "\n")
}
```

At V1 you should see a populous top-right cluster — the V1 acute signal is robust to method. At V2 the cluster shrinks but the same canonical pathways survive (TNFA, Inflammatory, IFN-γ). At V5 the cluster is smallest and may even be empty for ORA (because few V5 genes pass `|LFC|>1, padj<0.05`); GSEA-only points above the horizontal but left-of-vertical are then the *subtle persistent* pathways, exactly the case GSEA was designed for.

## 12.2d Robustness check: GSEA with LFC ranker vs signed −log10(p) ranker, at every visit

Same robustness test as before, run for each visit. Each panel shows NES from the LFC ranker vs NES from the signed-p ranker; tight diagonal clustering of the "Sig in both" points means the result is not an artefact of how we rank the genes.

```{r gsea-ranker-compare-run, cache=TRUE}
gsea_results_signedp <- list()
for (tp in names(ranked_lists_sigp))
  gsea_results_signedp[[tp]] <- GSEA(geneList     = ranked_lists_sigp[[tp]],
                                      TERM2GENE    = t2g_hallmark,
                                      pvalueCutoff = 1, eps = 0)
```

```{r gsea-ranker-compare-plot, fig.width=18, fig.height=6}
plot_ranker_compare <- function(g_lfc, g_sgp, title) {
  df_lfc <- as.data.frame(g_lfc) %>%
    dplyr::transmute(ID, NES_lfc = NES, padj_lfc = p.adjust)
  df_sgp <- as.data.frame(g_sgp) %>%
    dplyr::transmute(ID, NES_sgp = NES, padj_sgp = p.adjust)
  d <- dplyr::full_join(df_lfc, df_sgp, by = "ID") %>%
    dplyr::mutate(
      padj_lfc = ifelse(is.na(padj_lfc), 1, padj_lfc),
      padj_sgp = ifelse(is.na(padj_sgp), 1, padj_sgp),
      label    = gsub("HALLMARK_", "", ID),
      both_sig = padj_lfc < 0.05 & padj_sgp < 0.05
    )
  rho <- suppressWarnings(cor(d$NES_lfc, d$NES_sgp, use = "complete.obs"))
  ggplot(d, aes(x = NES_lfc, y = NES_sgp)) +
    geom_point(aes(color = both_sig), alpha = 0.85, size = 2.4) +
    scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "gray70"),
                       labels = c("Sig in ≤1 ranker", "Sig in both"),
                       name = NULL) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = 0, linetype = "dotted", color = "gray60") +
    geom_vline(xintercept = 0, linetype = "dotted", color = "gray60") +
    geom_text_repel(data = d %>% dplyr::filter(both_sig),
                    aes(label = label), size = 2.5, max.overlaps = 25) +
    labs(title = paste0(title, "  (Pearson r = ", round(rho, 2), ")"),
         x = "NES (LFC ranker)", y = "NES (signed −log10p ranker)") +
    theme(legend.position = "bottom",
          plot.title = element_text(size = 11))
}

grid.arrange(
  plot_ranker_compare(gsea_results$V1, gsea_results_signedp$V1, "V1 vs Control"),
  plot_ranker_compare(gsea_results$V2, gsea_results_signedp$V2, "V2 vs Control"),
  plot_ranker_compare(gsea_results$V5, gsea_results_signedp$V5, "V5 vs Control"),
  ncol = 3
)
```

Pearson correlations near 1 across all three visits mean the GSEA result is genuinely robust to ranker choice at every disease stage, *not* just at V1. If V5 has a noticeably lower correlation it is usually because the V5 signal is small enough that ranker noise matters more (small denominators).

## 12.3 GSEA Enrichment ("mountain") Plots for Key Pathways — across all visits

For each of the four signature pathways we draw the mountain plot at every visit. The peak height (enrichment score, ES) tracks how strongly the pathway's genes cluster at the top of the ranked list, visually showing the V1 → V2 → V5 *decay* of each pathway in one figure.

```{r gsea-mountains, fig.show='hold', out.width='32%', fig.height=6}
key_pathways <- c("HALLMARK_INFLAMMATORY_RESPONSE",
                  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                  "HALLMARK_COMPLEMENT")

mountain_for <- function(g, pw, title) {
  df <- as.data.frame(g)
  if (!(pw %in% df$ID))
    return(ggplot() + theme_void() +
           ggtitle(paste0(title, "\n(", gsub("HALLMARK_", "", pw),
                          " not enriched here)")))
  gseaplot2(g, geneSetID = pw, title = paste0(title, " — ",
                                              gsub("HALLMARK_", "", pw)))
}

#for (pw in key_pathways) {
 # cat("\n\n#### ", gsub("HALLMARK_", "", pw), "\n\n", sep = "")
  #grid.arrange(
   # mountain_for(gsea_results$V1, pw, "V1"),
    #mountain_for(gsea_results$V2, pw, "V2"),
    #mountain_for(gsea_results$V5, pw, "V5"),
  #  ncol = 3
  #)
#}
for (pw in key_pathways) {
  cat("\n\n#### ", gsub("HALLMARK_", "", pw), "\n\n", sep = "")
  
  # Just print them. The chunk options out.width='32%' and fig.show='hold' 
  # will automatically place them in a 3-column row!
  print(mountain_for(gsea_results$V1, pw, "V1"))
  print(mountain_for(gsea_results$V2, pw, "V2"))
  print(mountain_for(gsea_results$V5, pw, "V5"))
}
```

 A high peak on the left = pathway genes pile up at the top of the ranked list (activated, in our LFC ranker that means upregulated). The peak typically gets lower and broader from V1 → V2 → V5; if the peak goes flat by V5, the pathway has fully resolved; if it stays sharp, the pathway is persistent, which is the canonical chronic / PTLDS-relevant biology.

## 12.4 GSEA NES heatmap — all Hallmarks across all visits

```{r nes-heatmap, fig.height=9, fig.width=7}
nes_all <- lapply(names(gsea_results), function(tp) {
  as.data.frame(gsea_results[[tp]]) %>%
    dplyr::select(ID, NES, p.adjust) %>%
    dplyr::mutate(timepoint = tp)
}) %>% dplyr::bind_rows()

# Keep pathways significant in at least one time point
sig_any <- nes_all %>% dplyr::filter(p.adjust < 0.05) %>%
  dplyr::pull(ID) %>% unique()

if (length(sig_any) > 0) {
  nes_mat <- nes_all %>%
    dplyr::filter(ID %in% sig_any) %>%
    dplyr::select(ID, NES, timepoint) %>%
    tidyr::pivot_wider(names_from = timepoint, values_from = NES) %>%
    tibble::column_to_rownames("ID") %>% as.matrix()
  nes_mat[is.na(nes_mat)] <- 0
  nes_mat <- nes_mat[, intersect(c("V1","V2","V5"), colnames(nes_mat)), drop = FALSE]

  pheatmap(nes_mat,
           color = colorRampPalette(c("steelblue","white","firebrick"))(100),
           cluster_cols = FALSE,
           main = "GSEA NES: Hallmark Pathways Across Time Points",
           fontsize_row = 7)
}
```

This heatmap is the compact "everything at once" view. **Red** = activated, **blue** = suppressed. Pathways that remain red from V1 through V5 represent the **sustained signature** described in the paper. Pathways that fade to white by V5 indicate resolution. Compared to the per-visit dotplots above this heatmap is best for spotting *patterns of decay* (gradient from V1 to V5) rather than absolute amplitudes.

---
---

# 13. Biologic Theme Comparison

*Ref: Thomas's book Ch. 8.4.2*

`compareCluster` from clusterProfiler directly compares enrichment results across multiple gene lists (our three time points), making the temporal dynamics visually clear.

### Why V5 was missing in the previous version of this plot

`compareCluster()` with `pvalueCutoff = 0.05` *drops any cluster (visit) that has zero significant terms* from the dotplot, which is why earlier renders showed only V1 and V2, V5 has fewer DEGs and frequently no Hallmark passes the strict 0.05 cutoff. We diagnose this explicitly below and use two complementary remedies so V5 is always represented:

1. Print the per-visit input gene-list size and the per-visit number of significant Hallmarks so we can *see* the dropout numerically.
2. Use a more permissive `pvalueCutoff` (0.2) and rely on **dot transparency / size** (driven by `p.adjust`) to convey significance — this keeps V5 in the figure even when nothing passes 0.05, and makes the V1 → V2 → V5 *amplitude decay* the actual visual story.

```{r compare-cluster-diagnose}
gene_lists <- lapply(res_list, function(df) {
  df %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                  !is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
})

cat("DEG-list sizes feeding compareCluster:\n")
for (tp in names(gene_lists))
  cat("  ", tp, ":", length(gene_lists[[tp]]), "genes\n")

# Count Hallmarks significant per visit at the strict cutoff
cat("\nHallmarks at padj<0.05 per visit (from Section 11.2 ora_hallmark_all):\n")
for (tp in names(ora_hallmark_all))
  cat("  ", tp, ":", sum(as.data.frame(ora_hallmark_all[[tp]])$p.adjust < 0.05),
      "Hallmarks\n")
```

```{r compare-cluster, fig.height=10, fig.width=10}
cc_result <- compareCluster(
  geneClusters  = gene_lists,
  fun           = "enricher",
  TERM2GENE     = t2g_hallmark,
  pvalueCutoff  = 0.20,
  pAdjustMethod = "BH"
)

cc_df <- as.data.frame(cc_result)
cat("Visits represented in compareCluster output:\n")
print(table(cc_df$Cluster))

if (nrow(cc_df) > 0)
  dotplot(cc_result, showCategory = 20) +
    ggtitle("Biologic Theme Comparison: Hallmark across V1, V2, V5") +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          axis.text.y = element_text(size = 8))
```

**How to read the figure (now with V5 included):**

- Each column is a visit; each row is a Hallmark pathway. **Dot size** = number of V-specific DEGs that hit the pathway. **Dot colour** = `p.adjust` (darker = more significant).
- Pathways with dots in **all three columns** are the **persistent** signature, TNFA-via-NFKB, Inflammatory Response, IFN-γ Response are typical members.
- Pathways with dots in V1 (and maybe V2) but **missing at V5** are the *acute-only* signature, typical examples are Complement and IL6-JAK-STAT3 in this cohort.
- A column that is *very sparse* at V5 (few or no dots) is itself a finding: it directly visualises the cohort-mean resolution of the acute response by 6 months. Combined with the per-patient GSVA boxplots in Section 14.7, this distinguishes the cohort-level resolution from the subset of PTLDS-candidate patients.
- If you would like a stricter view (only `padj < 0.05` dots), uncomment `pvalueCutoff = 0.05` above, V5 will then drop out as in the earlier version, but the V1/V2 dots that remain will all be unambiguously significant.


---
---

# 14. 2nd Generation (cont.): GSVA (Gene Set Variation Analysis)

*Ref: Thomas's book Ch. 8.1.2.2, 8.4.5; Hänzelmann et al. 2013*

GSVA completes the 2nd generation of pathway analysis. Unlike GSEA
(which compares two groups), GSVA computes a **per-sample pathway
score**. This preserves the full sample structure, so we can test
pathway differences with standard linear models -- exactly as Thomas
describes in Ch. 8.4.5.

Think of it this way:

- **ORA** (1st gen): "Which pathways have too many DEGs?" (binary: gene is DE or not)
- **GSEA** (2nd gen-a): "Do pathway genes cluster at the top/bottom of the ranked list?" (uses fold-change magnitude, but still compares two groups)
- **GSVA** (2nd gen-b): "For each sample individually, how active is this pathway?" (produces a score per sample per pathway -- then you can run any statistical model on those scores)

## 14.1 Prepare Gene Sets as Named Lists

GSVA needs gene sets as a named list of character vectors (gene symbols).

```{r gsva-genesets}
hallmark_sets <- msigdb_hallmark %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(unique(gene_symbol)), .groups = "drop") %>%
  tibble::deframe()

cat("Hallmark gene sets:", length(hallmark_sets), "\n")
cat("Example set sizes:", head(sapply(hallmark_sets, length)), "\n")
```

## 14.2 Prepare Expression Matrix with Gene Symbols

GSVA needs the matrix rows to be gene symbols (matching the gene sets),
not Entrez IDs.

```{r gsva-matrix}
# Use VST-normalised data for GSVA (Gaussian kernel, Ch. 8.4.5)
vst_mat <- assay(vsd)

# Map Entrez IDs to Symbols
symbol_map <- gene_annot$Symbol[match(as.numeric(rownames(vst_mat)),
                                       gene_annot$GeneID)]

# Keep only rows with valid symbols, handle duplicates by keeping highest mean
keep_sym <- !is.na(symbol_map) & symbol_map != ""
vst_sym  <- vst_mat[keep_sym, ]
rownames(vst_sym) <- symbol_map[keep_sym]

# Remove duplicate gene symbols (keep the one with highest mean expression)
dup_idx <- duplicated(rownames(vst_sym))
vst_sym <- vst_sym[!dup_idx, ]

cat("GSVA input matrix:", nrow(vst_sym), "genes x", ncol(vst_sym), "samples\n")
```

## 14.3 Run GSVA

```{r run-gsva, cache=TRUE}
gsva_param <- gsvaParam(
  exprData = vst_sym,
  geneSets = hallmark_sets,
  kcdf     = "Gaussian",   # Gaussian for continuous (VST) data; Poisson for counts
  maxDiff  = TRUE
)

gsva_scores <- gsva(gsva_param)

cat("GSVA output:", nrow(gsva_scores), "pathways x", ncol(gsva_scores), "samples\n")
```

## 14.4 GSVA Heatmap

Each cell shows how active a pathway is in a given sample (relative
to other samples). This reveals the per-patient, per-time-point
pathway landscape.

```{r gsva-heatmap, fig.height=12, fig.width=10}
pheatmap(gsva_scores,
         annotation_col = row_annot,
         annotation_colors = ann_colors,
         show_colnames = FALSE,
         fontsize_row = 6,
         color = colorRampPalette(c("steelblue", "white", "firebrick"))(100),
         main = "GSVA Scores: Hallmark Pathways (per sample)",
         clustering_distance_rows = "correlation",
         cutree_cols = 4)
```

The key advantage of GSVA over GSEA: you can see individual patient
variation. Some V5 patients may still have highly activated
inflammatory pathways (potential PTLDS candidates), while others
return to control-like levels.

## 14.5 Test GSVA Scores Between Groups

Following Thomas's Ch. 8.4.5: since GSVA scores are continuous values
per sample, we can use a standard linear model (limma) to test which
pathways differ significantly between groups. This mirrors the workflow
from Ch. 7.5.1.

```{r gsva-limma}
library(limma)

design_gsva <- model.matrix(~ 0 + group, data = annot)
colnames(design_gsva) <- levels(annot$group)

# Fit linear model to GSVA scores (instead of gene expression)
fit_gsva <- lmFit(gsva_scores, design_gsva)

# Define contrasts (same comparisons as DESeq2)
contrast_mat <- makeContrasts(
  V1_vs_Ctrl = Lyme_V1 - Control,
  V2_vs_Ctrl = Lyme_V2 - Control,
  V5_vs_Ctrl = Lyme_V5 - Control,
  levels = design_gsva
)

fit_gsva2 <- contrasts.fit(fit_gsva, contrast_mat) %>% eBayes()
```

```{r gsva-results}
gsva_results_list <- list()
for (coef_name in colnames(contrast_mat)) {
  tt <- topTable(fit_gsva2, coef = coef_name, number = Inf) %>%
    tibble::rownames_to_column("pathway")
  gsva_results_list[[coef_name]] <- tt
  
  n_sig <- sum(tt$adj.P.Val < 0.05)
  cat(coef_name, ":", n_sig, "pathways with adj.P.Val < 0.05\n")
}
```

## 14.6 GSVA Volcano-Style Plot

```{r gsva-volcano, fig.width=16, fig.height=6}
gsva_volcano_plots <- lapply(names(gsva_results_list), function(nm) {
  df <- gsva_results_list[[nm]] %>%
    dplyr::mutate(
      sig = ifelse(adj.P.Val < 0.05, "Significant", "NS"),
      label = ifelse(adj.P.Val < 0.05,
                     gsub("HALLMARK_", "", pathway), NA)
    )
  ggplot(df, aes(x = logFC, y = -log10(P.Value), color = sig)) +
    geom_point(size = 2.5) +
    scale_color_manual(values = c(Significant = "firebrick", NS = "gray60")) +
    geom_text_repel(aes(label = label), size = 2.2, max.overlaps = 20,
                    na.rm = TRUE) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    labs(title = gsub("_", " ", nm),
         x = "GSVA score difference", y = "-log10(p-value)") +
    theme(legend.position = "none")
})

grid.arrange(grobs = gsva_volcano_plots, ncol = 3)
```

Each dot is a Hallmark pathway. The x-axis is the difference in GSVA
score between Lyme and Control (positive = more active in Lyme).
Labeled pathways are significant (adj.P.Val < 0.05). Compare the
three panels: V1 should have the most extreme differences, V5 the
least -- mirroring the gene-level DEG pattern.

## 14.7 GSVA Score Boxplots for Key Pathways

```{r gsva-boxplots, fig.height=10, fig.width=12}
key_pw <- c("HALLMARK_INFLAMMATORY_RESPONSE",
            "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
            "HALLMARK_INTERFERON_GAMMA_RESPONSE",
            "HALLMARK_COMPLEMENT",
            "HALLMARK_IL6_JAK_STAT3_SIGNALING",
            "HALLMARK_APOPTOSIS")

gsva_long <- gsva_scores[rownames(gsva_scores) %in% key_pw, ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("pathway") %>%
  tidyr::pivot_longer(-pathway, names_to = "GSM", values_to = "score") %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group, visit), by = "GSM") %>%
  dplyr::mutate(pathway = gsub("HALLMARK_", "", pathway))

ggplot(gsva_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.6) +
  facet_wrap(~ pathway, scales = "free_y", ncol = 3) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "GSVA Scores for Key Immune Pathways",
       subtitle = "Each dot is a patient. Higher = more pathway activity in that sample.",
       x = NULL, y = "GSVA Score") +
  theme(legend.position = "none",
        strip.text = element_text(size = 8))
```

These boxplots are the unique output of GSVA that neither ORA nor GSEA
can produce. Each dot is an individual patient's pathway activity.
Look for:

- **V1 >> Control**: Pathway is activated during acute infection (expected for immune/inflammatory pathways).
- **V2 still elevated**: Pathway stays active even after antibiotic treatment, evidence for the sustained signature.
- **V5 approaching Control**: Pathway is resolving (good sign for most patients).
- **V5 patients still elevated**: These outlier dots may correspond to patients with persistent symptoms (PTLDS).

This connects directly to the paper's central finding: the gene expression
signature of Lyme disease persists after treatment. GSVA lets us see
this at the **individual patient level**, which the original study's
bulk comparisons could not show.

---
---

# 15. Third Generation: SPIA (topology-aware KEGG perturbation)

*Ref: Khatri et al. 2012 (three generations); Tarca et al. 2009; Thomas's book Ch. 8 — pathway topology*

First-generation ORA asks whether a pathway contains more DEG labels than expected. Second-generation GSEA/GSVA uses ranked scores or per-sample enrichment without explicit *wiring* between genes. **Signalling Pathway Impact Analysis (SPIA)** adds a *perturbation* term: it walks curated **KEGG** signalling topology so that genes upstream/downstream and activation vs inhibition relations contribute differently from a flat hypergeometric count. The combined evidence (over-representation **pNDE** + topology **pPERT**) yields a global **pG** (and FDR-style **pGFdr**).

### Should we trust SPIA for this course project?

SPIA is pedagogically valuable: it makes the “topology” idea concrete and produces interpretable **Activated / Inhibited** status per pathway. For a *scientific* discussion with your professor, consider these limitations:

- **Database scope:** SPIA is tied to **KEGG** pathway maps (not MSigDB Hallmark). Immune–infection biology is only partially captured; many Lyme-relevant processes appear in Hallmark/GO instead.
- **Model assumptions:** Edge signs and “perturbation accumulation” are approximations; real immune networks are context- and cell-type–specific.
- **Redundancy:** Top SPIA hits may overlap messages already seen in ORA/GSEA/GSVA — agreement strengthens evidence; disagreement invites caution, not automatic dismissal.
- **Licensing / maintenance:** KEGG data are periodically updated; results can shift slightly between SPIA/database versions.

We run SPIA **anyway** so you can compare outputs side-by-side with first- and second-generation results and document the workflow.

## 15.1 Prepare DE vectors (Entrez) and run SPIA

SPIA expects a **named numeric vector of log2 fold changes for differentially expressed genes** (Entrez IDs as names) and the **universe** of all Entrez IDs tested (`rownames(dds)`).

```{r spia-run, cache=TRUE}
spia_from_res <- function(res_df, dds_obj) {
  sig <- res_df %>%
    dplyr::filter(!is.na(padj), padj < 0.05, !is.na(log2FoldChange)) %>%
    dplyr::mutate(GeneID = as.character(GeneID))
  if (nrow(sig) < 5)
    return(NULL)
  # one value per Entrez: keep strongest |LFC| if duplicates exist
  sig <- sig %>%
    dplyr::group_by(GeneID) %>%
    dplyr::slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
    dplyr::ungroup()
  de <- sig$log2FoldChange
  names(de) <- sig$GeneID
  all_ids <- as.character(rownames(dds_obj))
  spia(de, all_ids, organism = "hsa", nB = 2000, verbose = FALSE)
}

spia_v1 <- spia_from_res(res_list$V1, dds)
spia_v2 <- spia_from_res(res_list$V2, dds)
spia_v5 <- spia_from_res(res_list$V5, dds)
```

## 15.2 Top KEGG pathways (global pG)

```{r spia-tables}
fmt_spia <- function(x, n = 12) {
  if (is.null(x) || nrow(x) == 0)
    return(invisible(NULL))
  tibble::rownames_to_column(as.data.frame(x), "pathway_id") %>%
    dplyr::arrange(pG) %>%
    dplyr::select(Name, ID, Status, pNDE, pPERT, pG, pGFdr) %>%
    head(n)
}

fmt_spia(spia_v1)
fmt_spia(spia_v2)
fmt_spia(spia_v5)
```

## 15.3 Visualise top SPIA hits across V1, V2, V5

For every visit we show the top-15 KEGG pathways (by global pG) as a horizontal bar plot, coloured by SPIA's `Status` (Activated / Inhibited / Mixed). Comparing the three panels gives the SPIA-equivalent of the GSEA NES heatmap: which KEGG signalling maps stay activated through V2 / V5, which fade.

```{r spia-plots-all, fig.height=7, fig.width=20}
plot_spia_top <- function(s, title, n = 15) {
  if (is.null(s) || nrow(s) == 0)
    return(ggplot() + theme_void() +
           ggtitle(paste(title, "\n(no SPIA results — too few DEGs at padj<0.05)")))
  top <- tibble::rownames_to_column(as.data.frame(s), "pathway_id") %>%
    dplyr::arrange(pG) %>%
    head(n)
  ggplot(top, aes(x = reorder(Name, -log10(pG)),
                  y = -log10(pG), fill = Status)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = c(Activated = "firebrick",
                                 Inhibited = "steelblue",
                                 Mixed = "gray50"),
                      na.value = "gray70") +
    labs(title = title,
         subtitle = "Bar height = -log10(global pG). Status from SPIA.",
         x = NULL, y = "-log10(pG)") +
    theme(legend.position = "bottom",
          axis.text.y = element_text(size = 8),
          plot.title  = element_text(size = 11))
}

grid.arrange(
  plot_spia_top(spia_v1, "V1 (acute) vs Control"),
  plot_spia_top(spia_v2, "V2 (post-treatment) vs Control"),
  plot_spia_top(spia_v5, "V5 (6 months) vs Control"),
  ncol = 3
)
```

**Reading across panels.** Pathways with large **|tA|** (total net accumulation) and small **pPERT** are those where **topology** contributes strongly, not just list overlap. KEGG pathways that stay coloured firebrick at V2/V5 are SPIA's version of the persistent inflammatory signature, typically *Cytokine-cytokine receptor interaction*, *TNF signalling*, *JAK-STAT*, *Toll-like receptor signalling* are the ones that survive longest. KEGG pathways that drop out from V1 to V5 (often *Cell cycle*, *DNA replication*) are the proliferation/expansion programmes that resolve once the acute response ends.

## 15.4 Does SPIA agree with the Hallmark / GO results? (cross-database validation)

**Short answer:** SPIA uses a *different reference database* (KEGG signalling maps) than ORA/GSEA/GSVA in Sections 11–14 (which use **MSigDB Hallmark** + **GO:BP**). Pathway *names* therefore do not match 1-to-1 — but we can still check whether the same biological signal is captured by all three databases.

### Why a direct name-match doesn't work

| Database | Source | Granularity | Used by |
|----------|--------|-------------|---------|
| **MSigDB Hallmark** | Liberzon *et al.* 2015, 50 curated, low-redundancy gene sets | broad biological themes (~200 genes / set) | ORA, GSEA, GSVA in Sections 11–14 |
| **GO:BP** | Gene Ontology Consortium | thousands of overlapping terms | ORA in Section 11.3 |
| **KEGG signalling maps** | Kanehisa Lab | ~330 curated pathways with **edges** (activation/inhibition) | SPIA in this section |

There is overlap *in the underlying biology* (e.g. "Cytokine-cytokine receptor interaction" in KEGG ↔ `HALLMARK_INFLAMMATORY_RESPONSE`) but the gene sets are not identical, KEGG is more mechanistic and signalling-focused, Hallmark is more thematic.

### Strategy

The cleanest validation is **same database, multiple methods**: run **ORA-KEGG** and **GSEA-KEGG** on V1 using KEGG gene sets pulled from MSigDB, then compare to **SPIA**. Any disagreement is then *methodological* (topology vs. flat enrichment), not *database-driven*. We then add a **gene-level** overlap with the Hallmark leading edge to bridge to Sections 11–12.

### 15.4.1 Pull KEGG gene sets from MSigDB

```{r kegg-genesets, cache=FALSE}
msigdb_kegg <- msigdbr(species = "Homo sapiens", collection = "C2") %>%
  dplyr::filter(grepl("^KEGG_", gs_name))

cat("KEGG gene sets pulled from MSigDB:",
    length(unique(msigdb_kegg$gs_name)), "pathways\n")

t2g_kegg <- msigdb_kegg %>% dplyr::select(gs_name, gene_symbol)
```

### 15.4.2 KEGG ORA + KEGG GSEA at every visit

```{r kegg-ora-gsea-all, cache=TRUE}
ora_kegg_all  <- list()
gsea_kegg_all <- list()
for (tp in names(res_list)) {
  ora_kegg_all[[tp]] <- enricher(
    gene         = sig_sym_per_tp[[tp]],
    universe     = univ_sym_per_tp[[tp]],
    TERM2GENE    = t2g_kegg,
    pvalueCutoff = 1, pAdjustMethod = "BH", qvalueCutoff = 1
  )
  gsea_kegg_all[[tp]] <- GSEA(
    geneList     = ranked_lists[[tp]],
    TERM2GENE    = t2g_kegg,
    pvalueCutoff = 1, eps = 0
  )
}

# Backward-compat aliases
ora_kegg_v1  <- ora_kegg_all$V1
gsea_kegg_v1 <- gsea_kegg_all$V1

for (tp in names(res_list)) {
  cat(tp,
      "  KEGG-ORA  sig (padj<0.05):",
      sum(as.data.frame(ora_kegg_all[[tp]])$p.adjust  < 0.05),
      "  KEGG-GSEA sig (padj<0.05):",
      sum(as.data.frame(gsea_kegg_all[[tp]])$p.adjust < 0.05), "\n")
}
```

### 15.4.3 Map SPIA → MSigDB-KEGG and merge — for each visit

SPIA exposes the human-readable pathway name (e.g. `"Cytokine-cytokine receptor interaction"`), MSigDB exposes the upper-snake-case `gs_name` (e.g. `"KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION"`). We normalise both, join, and produce a per-visit `cross_kegg_<tp>` table.

```{r spia-kegg-mapping-all}
norm_kegg_name <- function(s) {
  s <- toupper(s); s <- gsub("[^A-Z0-9]+", "_", s)
  gsub("^_+|_+$", "", s)
}

build_cross <- function(spia_obj, ora_obj, gsea_obj) {
  if (is.null(spia_obj) || nrow(spia_obj) == 0) return(NULL)
  spia_df <- tibble::rownames_to_column(as.data.frame(spia_obj), "spia_pwid") %>%
    dplyr::transmute(spia_pwid,
                     pathway_name = Name,
                     norm         = norm_kegg_name(Name),
                     spia_status  = Status,
                     spia_pG      = pG,
                     spia_pGFdr   = pGFdr)
  ora_df  <- as.data.frame(ora_obj) %>%
    dplyr::transmute(norm = sub("^KEGG_", "", ID),
                     ora_padj  = p.adjust,
                     ora_count = Count)
  gsea_df <- as.data.frame(gsea_obj) %>%
    dplyr::transmute(norm = sub("^KEGG_", "", ID),
                     gsea_padj = p.adjust,
                     gsea_NES  = NES)
  spia_df %>%
    dplyr::left_join(ora_df,  by = "norm") %>%
    dplyr::left_join(gsea_df, by = "norm")
}

cross_kegg_list <- list(
  V1 = build_cross(spia_v1, ora_kegg_all$V1, gsea_kegg_all$V1),
  V2 = build_cross(spia_v2, ora_kegg_all$V2, gsea_kegg_all$V2),
  V5 = build_cross(spia_v5, ora_kegg_all$V5, gsea_kegg_all$V5)
)

# Backward-compat alias
cross_kegg <- cross_kegg_list$V1

for (tp in names(cross_kegg_list)) {
  ck <- cross_kegg_list[[tp]]
  if (is.null(ck)) {
    cat(tp, ": no SPIA results to map.\n"); next
  }
  cat(tp, ": SPIA pathways successfully mapped to MSigDB KEGG:",
      sum(!is.na(ck$gsea_padj) | !is.na(ck$ora_padj)),
      "/", nrow(ck), "\n")
}
```

```{r spia-cross-tables-all, results='asis'}
fmt_cross <- function(ck, n = 10) {
  if (is.null(ck) || nrow(ck) == 0) return(NULL)
  ck %>%
    dplyr::arrange(spia_pG) %>%
    dplyr::mutate(
      spia_pG    = signif(spia_pG, 2),
      spia_pGFdr = signif(spia_pGFdr, 2),
      ora_padj   = signif(ora_padj, 2),
      gsea_padj  = signif(gsea_padj, 2),
      gsea_NES   = round(gsea_NES, 2)
    ) %>%
    dplyr::select(pathway_name, spia_status,
                  spia_pG, spia_pGFdr,
                  ora_padj, ora_count,
                  gsea_NES, gsea_padj) %>%
    head(n)
}

for (tp in names(cross_kegg_list)) {
  cat("\n\n#### Top 10 SPIA pathways at ", tp, " — with KEGG-ORA & KEGG-GSEA evidence\n\n", sep = "")
  tab <- fmt_cross(cross_kegg_list[[tp]], 10)
  if (is.null(tab)) {
    cat("_No SPIA results at this visit._\n")
  } else {
    print(knitr::kable(tab, row.names = FALSE))
  }
}
```

### 15.4.4 Visualise SPIA ↔ KEGG-GSEA agreement at every visit

```{r spia-vs-gsea-kegg-all, fig.width=18, fig.height=6}
plot_spia_gsea <- function(ck, title) {
  if (is.null(ck))
    return(ggplot() + theme_void() +
           ggtitle(paste(title, "\n(no SPIA results)")))
  d <- ck %>%
    dplyr::filter(!is.na(spia_pG) & !is.na(gsea_padj)) %>%
    dplyr::mutate(
      sig_spia = spia_pGFdr < 0.05,
      sig_gsea = gsea_padj  < 0.05,
      both     = sig_spia & sig_gsea,
      label    = ifelse(both, pathway_name, NA)
    )
  ggplot(d, aes(x = -log10(spia_pG), y = -log10(gsea_padj))) +
    geom_point(aes(color = both, size = abs(gsea_NES)), alpha = 0.85) +
    scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "gray70"),
                       labels = c("Sig in ≤1 method", "Sig in both"),
                       name = NULL) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    geom_text_repel(aes(label = label), size = 2.5, max.overlaps = 25, na.rm = TRUE) +
    labs(title = title,
         x = "-log10(SPIA pG)", y = "-log10(KEGG-GSEA padj)",
         size = "|GSEA NES|") +
    theme(legend.position = "bottom",
          plot.title = element_text(size = 11))
}

grid.arrange(
  plot_spia_gsea(cross_kegg_list$V1, "V1 vs Control"),
  plot_spia_gsea(cross_kegg_list$V2, "V2 vs Control"),
  plot_spia_gsea(cross_kegg_list$V5, "V5 vs Control"),
  ncol = 3
)
```

A populated top-right corner at V1 means SPIA and KEGG-GSEA agree on the same database the V1 inflammatory signal is robust to method, not just topology-induced. 

### 15.4.5 Bridge back to Hallmark, *gene-level* agreement at every visit

For each visit intersect the **leading-edge genes** of the top-5 Hallmark GSEA hits with the **member genes** of the top-5 SPIA KEGG pathways. High counts in red cells mean Hallmark and KEGG are describing the same gene populations.

```{r spia-hallmark-gene-overlap-all, fig.width=20, fig.height=5}
hallmark_members_long <- msigdb_hallmark %>%
  dplyr::select(gs_name, gene_symbol)

build_overlap <- function(g_hallmark, ck) {
  df_h <- as.data.frame(g_hallmark)
  if (nrow(df_h) == 0 || is.null(ck)) return(NULL)
  top5_h <- df_h %>% dplyr::arrange(p.adjust) %>% head(5)
  if (nrow(top5_h) == 0) return(NULL)
  h_genes <- setNames(strsplit(top5_h$core_enrichment, "/"),
                      gsub("HALLMARK_", "", top5_h$ID))
  top5_s <- ck %>% dplyr::arrange(spia_pG) %>% head(5)
  if (nrow(top5_s) == 0) return(NULL)
  kegg_members <- split(msigdb_kegg$gene_symbol, msigdb_kegg$gs_name)
  s_genes <- lapply(top5_s$norm, function(n) {
    key <- paste0("KEGG_", n)
    if (key %in% names(kegg_members)) unique(kegg_members[[key]]) else character(0)
  })
  names(s_genes) <- top5_s$pathway_name
  m <- outer(seq_along(h_genes), seq_along(s_genes),
             Vectorize(function(i, j) {
               a <- h_genes[[i]]; b <- s_genes[[j]]
               if (length(a) == 0 || length(b) == 0) return(0)
               length(intersect(a, b))
             }))
  rownames(m) <- names(h_genes); colnames(m) <- names(s_genes)
  m
}

plot_overlap <- function(m, title) {
  if (is.null(m))
    return(ggplot() + theme_void() +
           ggtitle(paste(title, "\n(no top hits to compare)")))
  d <- as.data.frame(as.table(m))
  colnames(d) <- c("Hallmark", "SPIA_KEGG", "shared_genes")
  ggplot(d, aes(x = SPIA_KEGG, y = Hallmark, fill = shared_genes)) +
    geom_tile(color = "white") +
    geom_text(aes(label = shared_genes), size = 3.2) +
    scale_fill_gradient(low = "white", high = "firebrick") +
    labs(title = title,
         x = "SPIA top KEGG", y = "Hallmark top GSEA",
         fill = "Shared\ngenes") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 7),
          axis.text.y = element_text(size = 7),
          plot.title  = element_text(size = 11),
          legend.position = "bottom")
}

m_v1 <- build_overlap(gsea_results$V1, cross_kegg_list$V1)
m_v2 <- build_overlap(gsea_results$V2, cross_kegg_list$V2)
m_v5 <- build_overlap(gsea_results$V5, cross_kegg_list$V5)

grid.arrange(
  plot_overlap(m_v1, "V1 — Hallmark ↔ SPIA KEGG"),
  plot_overlap(m_v2, "V2 — Hallmark ↔ SPIA KEGG"),
  plot_overlap(m_v5, "V5 — Hallmark ↔ SPIA KEGG"),
  ncol = 3
)
```

Strong red diagonals at V1 and V2 confirm that Hallmark and KEGG are describing the same biology at those time points (just in different vocabulary). At V5 the heatmap may be blank or pale, if either Hallmark GSEA or SPIA produces no top-5 hits at V5, that absence itself answers the question: there is no robust pathway-level signal at 6 months in this cohort, only a residual subset of patients (visible in 14.7).

### 15.4.6 Save the cross-database tables — for every visit

```{r save-spia-cross-all}
for (tp in names(cross_kegg_list)) {
  ck <- cross_kegg_list[[tp]]
  if (is.null(ck)) next
  write.csv(ck %>% dplyr::arrange(spia_pG),
            file.path(results_dir,
                      paste0("Cross_DB_SPIA_vs_KEGG_ORA_GSEA_", tp, ".csv")),
            row.names = FALSE)
  cat("Saved:",
      file.path(results_dir,
                paste0("Cross_DB_SPIA_vs_KEGG_ORA_GSEA_", tp, ".csv")), "\n")
}
```

---
---

Save All Results

```{r save-all-results}
for (tp in names(ora_hallmark_all)) {
  df <- as.data.frame(ora_hallmark_all[[tp]])
  if (nrow(df) > 0) write.csv(df, file.path(results_dir,
    paste0("ORA_Hallmark_", tp, ".csv")), row.names = FALSE)
}
for (tp in names(gsea_results)) {
  df <- as.data.frame(gsea_results[[tp]])
  if (nrow(df) > 0) write.csv(df, file.path(results_dir,
    paste0("GSEA_Hallmark_", tp, ".csv")), row.names = FALSE)
}
for (nm in names(gsva_results_list)) {
  write.csv(gsva_results_list[[nm]], file.path(results_dir,
    paste0("GSVA_Hallmark_", nm, ".csv")), row.names = FALSE)
}
write.csv(gsva_scores, file.path(results_dir, "GSVA_scores_per_sample.csv"))

if (!is.null(spia_v1) && nrow(spia_v1) > 0)
  write.csv(as.data.frame(spia_v1), file.path(results_dir, "SPIA_KEGG_V1_vs_Control.csv"),
            row.names = TRUE)
if (!is.null(spia_v2) && nrow(spia_v2) > 0)
  write.csv(as.data.frame(spia_v2), file.path(results_dir, "SPIA_KEGG_V2_vs_Control.csv"),
            row.names = TRUE)
if (!is.null(spia_v5) && nrow(spia_v5) > 0)
  write.csv(as.data.frame(spia_v5), file.path(results_dir, "SPIA_KEGG_V5_vs_Control.csv"),
            row.names = TRUE)

cat("All results saved to:", results_dir, "\n")
```

---
---

# 16. Results Interpretation & Discussion

This section brings together all findings into a coherent narrative.

## 16.1 What did we find? (The Big Picture)

```{r interpretation-summary}
cat("DATA\n","  Genome reference: GRCh38.p13 (upgraded from hg19)\n"
,"  Aligner: Kallisto (upgraded from TopHat2/Cufflinks2)\n",
"  Samples analysed:", ncol(counts_filtered), "\n",
"  Genes after filtering:", nrow(counts_filtered), "\n\n",
"DIFFERENTIAL EXPRESSION (DESeq2, |lfc|>1, padj<0.05)\n")
print(deg_summary)

cat("\nOriginal study (Limma/Voom on FPKM, |FC|>1.5, FDR<0.1%):\n",
"  V1: 1,235 DEGs | V2: 1,060 DEGs | V5: 686 DEGs\n\n",
"PATHWAY ANALYSIS (ORA + GSEA/GSVA + SPIA)\n",
"  1st gen ORA  (Hallmark, V1):", nrow(as.data.frame(ora_hallmark_v1)),
    "enriched pathways\n",
"  1st gen ORA  (GO:BP, V1):  ",
    sum(as.data.frame(ora_gobp_v1)$p.adjust < 0.05), "enriched terms\n")
cat("  2nd gen GSEA (Hallmark, V1):", sum(gsea_h_df$p.adjust < 0.05),
    "significant pathways\n",
"  2nd gen GSVA (Hallmark, V1):",
    sum(gsva_results_list$V1_vs_Ctrl$adj.P.Val < 0.05),
    "differentially active pathways\n")
if (!is.null(spia_v1) && nrow(spia_v1) > 0) {
  cat("  3rd gen SPIA (KEGG, V1):    ",
      sum(spia_v1$pGFdr < 0.05, na.rm = TRUE),
      "pathways with pGFdr < 0.05\n")
} else {
  cat("  3rd gen SPIA (KEGG, V1):    not computed or empty\n")
}
```

## 16.2 Biological Interpretation

### The Acute Phase (V1): A Strong Inflammatory Storm

At V1 (acute Lyme diagnosis, before antibiotics) we detect the largest number of DEGs and the most enriched pathways. ORA, GSEA and GSVA all converge on the same five Hallmark pathways at the top , and the cross-method validation in **Section 12.2c** confirms they are not method-specific artefacts. The dominant biological themes are:

- **TNFA signaling via NF-κB** , *master pro-inflammatory programme*. The full mechanism (TNF-α → TNFR1 → IKK → NF-κB → autocrine cytokine loop), the leading-edge gene table, and the cytokine glossary are detailed in **Section 11.2.1**. Original paper also called out TNF-α as a top upstream regulator.
- **Interferon-γ response** , IFN-γ from T cells and NK cells activates macrophages and drives Th1 immunity (the lineage required against spirochetal pathogens). Matches the original study's TREM1-mediated Th1 activation finding. The core ISGs *ISG15 / IFI27 / BATF2* (highlighted in **Section 10.6**) explain why this Hallmark scores so high.
- **Inflammatory response** , broad cytokine, chemokine, and leukocyte-recruitment programme; expected hallmark of an acute bacterial infection.
- **Complement** , ancient innate defence that lyses bacteria and tags them for phagocytosis. *SERPING1* (a complement regulator highlighted in 10.6) sits in the leading edge.
- **IL6–JAK–STAT3 signaling** , IL-6 drives the acute-phase response (CRP, fever); identified as a top upstream regulator at V1 and V2 in the original paper.

The gene-level story (**Section 10.6**) maps cleanly onto these pathway-level themes:

- **Pathogen sensing.** *TLR2* (Section 10.6, pick #4) is the direct *B. burgdorferi* lipoprotein receptor , the molecular entry point that lights up the entire downstream NF-κB / IFN-γ programme described above.
- **Innate effector activation.** *S100A8 / S100A9* (calprotectin, pick #2) and *FCGR1A* (CD64, pick #3) reflect massive neutrophil + monocyte activation, which is what the *Inflammatory Response* and *Complement* Hallmarks summarise at the gene-set level.
- **Persistent biomarker.** *CXCL10* (pick #1) is the single most cited Lyme/PTLDS biomarker; its strong V1 induction is the gene-level driver of *Interferon-γ Response* and *IL6-JAK-STAT3*, and it carries forward into V2/V5 (see below).
- **Resolution / exhaustion machinery already engaged.** *IDO1* (pick #5) and *CD274 / PD-L1* (pick #6) are induced even at V1, hinting that the host is already activating immunoregulatory brakes , relevant for why the signature later fails to fully resolve.

**SPIA (Section 15)** adds the topology-aware view: pathways where both list enrichment *and* signed graph perturbation suggest activation/inhibition (e.g. chemokine signalling, lysosome, cell-cycle KEGG maps). Treat as complementary to Hallmark/GO, not a one-to-one match.

### Post-Treatment (V2): The Signature Persists

At V2 (3 weeks later, after completing antibiotics) a substantial number of DEGs and Hallmark pathways remain active , the paper's central finding: **the gene-expression signature does not promptly return to baseline after treatment**, unlike most other acute infections.

The NES heatmap (**Section 12.4**) and GSVA boxplots (**Section 14.7**) show that most inflammatory Hallmarks (TNFA-via-NFKB, Inflammatory Response, IFN-γ Response , see 11.2.1 and 10.6 for what those *mean*) stay elevated at V2 with reduced but non-zero intensity. The mechanism implied by the genes in 10.6 is consistent with this: even after the bacterium is cleared, the *NF-κB feedback loop* documented in 11.2.1 has run for weeks, induced its own self-sustaining cytokine pool (IL-6, IL-1, TNF), and only relaxes slowly. The IDO1 / PD-L1 brakes seen at V1 are now actively trying to dampen the response, but incompletely.

### Six Months Later (V5): Partial but Incomplete Recovery

By V5 (6 months post-treatment) the number of DEGs drops substantially but does not reach zero, and a residual subset of inflammatory pathways still register as differentially active. The **per-sample GSVA boxplots** (Section 14.7) are the most informative view here: most V5 patients approach control levels, but **a minority of outlier samples remain elevated** , these are the candidate PTLDS cases.

The original paper found 686 DEGs at V5 and noted 31–60% pathway overlap with chronic autoimmune diseases (SLE, RA). Our re-analysis is consistent with that: a lingering transcriptomic perturbation at 6 months, driven by the same TNF-α/NF-κB and IFN-γ programmes from V1 (see 11.2.1) , just at lower intensity and now restricted to a subset of patients.

If we wanted *one gene to track* across V1 → V2 → V5 in a future serum-based clinical follow-up, **CXCL10 (pick #1 in Section 10.6)** is the strongest candidate, with prior independent validation as a PTLDS biomarker (Soloski et al. 2014, Aucott et al. 2016).

## 17.3 Answering Open Questions from the Analysis

**Section 5.2, "Is the mean set correctly for the mean-variance plot?"**
Yes. The plot correctly shows that for raw RNA-Seq counts, the variance
consistently exceeds the mean (overdispersion). The red Poisson line
(variance = mean) is a theoretical baseline. The fact that points sit
above it is not an error; it is the fundamental property of count data
that justifies using the negative binomial model in DESeq2.

**Section 7.4, "VST doesn't seem to flatten the mean-SD relationship?"**
The VST does reduce the dependency, even if visually the "before" and
"after" plots look similar. The key metric is the correlation between
mean and SD: even a modest reduction (e.g., from 0.8 to 0.5) meaningfully
improves downstream analyses like PCA. The fact that the correlation
doesn't drop to zero is normal -- VST is an approximation, not magic.
For small sample sizes, `rlog()` (regularised log) can perform better
than VST but is much slower. Both are acceptable for this project.

**MA plots (Sections 5.3, 7.3, 10.1b) , do they validate normalization?**  
Section 5.3 uses raw log2 counts: systematic offset is expected. Section 7.3 uses **size-factor–normalized** counts only; the loess trend should sit closer to **M = 0**. Section **10.1b** uses **DESeq2**'s `plotMA()` after dispersion shrinkage: the central mass of points at moderate-to-high expression should lie near **LFC = 0** for non-DEGs (blue = padj &lt; alpha). Together, these three plots answer "did we fix depth?" vs "what does the fitted model say about differential expression?"

**Section 12.2, "Which pathways should we expect activated/suppressed?"**
Based on what we know about Lyme disease biology:

*Expected activated:* Inflammatory response, TNF-alpha signaling,
interferon signaling (IFN-gamma, IFN-alpha), complement, IL6-JAK-STAT,
TLR signaling, NF-kB signaling, apoptosis.

*Potentially suppressed:* The original paper found eIF2 signaling
(translation initiation) downregulated at all time points. In our GSEA
results, look for pathways with negative NES (blue in the heatmap).
Suppressed MYC targets or mTORC1 signaling would indicate reduced
cell growth/proliferation, consistent with immune cells diverting
resources from growth to defence.

## 17.4 Book Chapter Reference Map

| Our Section | Thomas's Book Chapter | Method |
|-------------|-----------------------|--------|
| 4.1 Library sizes | Ch. 7.2 | Quality Control / depth imbalance check |
| 4.3 RLE plot | Ch. 7.2.4 | Relative Log Expression |
| 5.3 MA plot | Ch. 7.2.7 | Mean-difference plot |
| 7.2 Size factors | Ch. 7.6 | Median-of-ratios normalisation (handles depth imbalance from 4.1) |
| 7. DESeq2 | Ch. 7.6 | Differential expression |
| 8. PCA | Ch. 7.2.2 | Dimensionality reduction |
| 9. Distance heatmap | Ch. 7.2.1 | Sample distance |
| 10.1a Dispersion | Ch. 7.6 | Quantifies dataset variability |
| 10.1b / 7.3 / 5.3 MA | Ch. 7.2.7 | Mean–difference / DESeq2 MA |
| 10.6 Genes of interest | n/a | Literature-curated picks for discussion |
| 11. ORA | Ch. 8.1.1, 8.4.1 | 1st generation pathway analysis |
| 11.2.1 TNFA-via-NFKB | Ch. 8.2 | Detailed pathway explanation |
| 12. GSEA | Ch. 8.1.2.1, 8.4.3 | 2nd generation pathway analysis |
| 12.2b GSEA volcano | Ch. 8.4.3 | Pathway-level volcano (NES vs adj.p) |
| 12.2c ORA ↔ GSEA | Ch. 8 | Cross-method validation |
| 13. compareCluster | Ch. 8.4.2 | Biologic theme comparison |
| 14. GSVA | Ch. 8.1.2.2, 8.4.5 | 2nd generation (per-sample scores) |
| 15. SPIA | Ch. 8 (topology); Khatri et al. 2012 | 3rd generation (KEGG perturbation) |
| MSigDB usage | Ch. 8.2 | Molecular Signature Database |

---


```{r session-info, include=F}
sessionInfo()
```
