---
title: "Lyme Disease Differential Gene Expression Analysis"
subtitle: "Re-analysis of Bouquet et al. 2016 (GSE63085) using DESeq2 & clusterProfiler"
author: "Big Data Course Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 7
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", cache = TRUE)
```

# 1. Environment Setup

## 1.1 Install Required Packages (run once)

```{r install-packages, eval=FALSE, include=FALSE}
if (rstudioapi::isAvailable()) {setwd(dirname(rstudioapi::getActiveDocumentContext()$path))}
options(width = 200)


# use this function to install and load packages
using <- function(pkgs) {
  for (pkg in pkgs) {  # Loop through each package in the provided vector
    if (!requireNamespace(pkg, quietly = TRUE)) {    # If not installed, install it
      install.packages(pkg, repos = "https://cloud.r-project.org")}
    # Load the package:
    library(pkg, character.only = TRUE)}}

```

## 1.2 Load Libraries

```{r load-libraries}
using(c("tidyverse", "pheatmap", "RColorBrewer", "ggrepel", "corrplot", "gridExtra", "viridis", "BiocManager","DESeq2", "org.Hs.eg.db", "clusterProfiler", "EnhancedVolcano","msigdbr", "PCAtools", "GSVA", "enrichplot"))


theme_set(theme_bw(base_size = 13))
```

# 2. Data Loading

```{r directories}
project_dir <- getwd()
data_dir    <- file.path(project_dir, "Data")
results_dir <- file.path(project_dir, "results")
plots_dir   <- file.path(project_dir, "plots")

for (d in c(results_dir, plots_dir))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
```

---

# 2. Load Data

The raw FASTQ data from GEO accession GSE63085 was re-aligned by Thomas
Mohrkeg to the modern **GRCh38.p13** human reference genome (replacing
the outdated hg19 used in the original 2016 study), and quantified into
raw integer counts -- the proper input format for DESeq2.

**References:**
- Original study: Bouquet J et al. *mBio* 2016; 7(1):e00100-16.
- Downstream analysis methodology: Mohrkeg T. *Beginner's Handbook of
  -Omics Data Analysis*, Chapter 8.

## 2.1 Raw Counts Matrix

```{r load-counts}
counts_raw <- read.delim(
  gzfile(file.path(data_dir, "GSE63085_raw_counts_GRCh38.p13_NCBI.tsv.gz")),
  row.names = 1, check.names = FALSE
)

cat("Counts matrix:", nrow(counts_raw), "genes x", ncol(counts_raw), "samples\n")
cat("All integers:", all(counts_raw == round(counts_raw)), "\n")
```

## 2.2 Gene Annotation (featureData)

This maps NCBI Entrez Gene IDs to gene symbols, descriptions, biotypes,
chromosomal locations, and GO terms.

```{r load-gene-annot}
gene_annot <- read.delim(
  gzfile(file.path(data_dir, "Human.GRCh38.p13.annot.tsv.gz")),
  stringsAsFactors = FALSE
)

cat("Gene annotation:", nrow(gene_annot), "genes x", ncol(gene_annot), "columns\n")
cat("Columns:", paste(names(gene_annot), collapse = ", "), "\n")
```

## 2.3 Sample Annotation

```{r load-sample-annot}
sample_annot <- read.csv(
  file.path(project_dir, "sample_annotations.csv"),
  stringsAsFactors = FALSE
)
cat("Sample annotation:", nrow(sample_annot), "rows\n")
```

## 2.4 Align Samples

Thomas's pipeline excluded some samples during quality filtering. We keep
only those present in both the counts matrix and the annotation.

```{r align-data}
shared     <- intersect(colnames(counts_raw), sample_annot$GSM)
only_annot <- setdiff(sample_annot$GSM, colnames(counts_raw))

cat("Counts samples: ", ncol(counts_raw), "\n")
cat("Annotation rows:", nrow(sample_annot), "\n")
cat("Shared:         ", length(shared), "\n")
cat("Excluded by Thomas:", length(only_annot), "\n")


if (length(only_annot) > 0) {
  cat("\nExcluded sample details:\n")
    sample_annot %>%
    dplyr::filter(GSM %in% only_annot) %>%
    dplyr::select(GSM, Observation, time) %>%
    print()
}
```

```{r subset-and-order}
counts <- counts_raw[, shared]
annot  <- sample_annot %>%
  dplyr::filter(GSM %in% shared) %>%
  dplyr::arrange(match(GSM, shared))

stopifnot(all(colnames(counts) == annot$GSM))
```

## 2.5 Engineer Metadata Columns

We create clean factor variables for the experimental design.

```{r engineer-metadata}
annot <- annot %>%
  dplyr::mutate(
    patient_id = gsub("^.*-([0-9]+)_V.*$", "\\1",
                      gsub("^.*-([0-9]+)_V$", "\\1", Observation)),
    visit = dplyr::case_when(
      grepl("V1", Observation) ~ "V1",
      grepl("V2", Observation) ~ "V2",
      grepl("V5", Observation) ~ "V5",
      TRUE                     ~ "Control"),
    group = dplyr::case_when(
      disease.state == "Healthy controls" ~ "Control",
      visit == "V1" ~ "Lyme_V1",
      visit == "V2" ~ "Lyme_V2",
      visit == "V5" ~ "Lyme_V5"),
    condition = ifelse(disease.state == "Healthy controls", "Control", "Lyme"),
    group     = factor(group,     levels = c("Control","Lyme_V1","Lyme_V2","Lyme_V5")),
    visit     = factor(visit,     levels = c("Control","V1","V2","V5")),
    condition = factor(condition, levels = c("Control","Lyme"))
  )

rownames(annot) <- annot$GSM

cat("Samples per group:\n")
print(table(annot$group))
```

---

# 3. Gene Annotation Overview

## 3.1 Gene Biotype Distribution

Understanding which gene types dominate the dataset informs our filtering
strategy (for DE, protein-coding genes carry the most biological signal).


```{r gene-biotypes}
gene_annot_matched <- gene_annot %>%
  dplyr::filter(GeneID %in% as.numeric(rownames(counts)))
  
biotype_counts <- gene_annot_matched %>%
  dplyr::count(GeneType, sort = TRUE) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))

print(biotype_counts)

ggplot(biotype_counts %>% dplyr::filter(n > 50),
       aes(x = reorder(GeneType, n), y = n, fill = GeneType)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(n, " (", pct, "%)")), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Gene Biotypes in Counts Matrix (GRCh38.p13)",
       x = "Gene Type", y = "Number of Genes") +
  scale_fill_viridis_d()
```

---

# 4. Sample-Level Quality Control

## 4.1 Library Sizes

Library size = total mapped read counts per sample. Large differences
mean different sequencing depths, which DESeq2 normalizes via size
factors.

```{r library-sizes, fig.height=10}
lib_sizes <- data.frame(
  GSM   = colnames(counts),
  total = colSums(counts),
  stringsAsFactors = FALSE
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group, Observation), by = "GSM")

ggplot(lib_sizes, aes(x = reorder(Observation, total),
                       y = total / 1e6, fill = group)) +
  geom_col() + coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library Size per Sample (millions of reads)",
       x = "Sample", y = "Total Counts (millions)", fill = "Group") +
  theme(axis.text.y = element_text(size = 5))
```

```{r lib-size-boxplot}
ggplot(lib_sizes, aes(x = group, y = total / 1e6, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Library Size Distribution by Group",
       x = "Group", y = "Total Counts (millions)") +
  theme(legend.position = "none")
```

## 4.2 Detected Genes per Sample

```{r detected-genes}
detected <- data.frame(
  GSM = colnames(counts),
  n_detected = colSums(counts > 0),
  stringsAsFactors = FALSE
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(detected, aes(x = group, y = n_detected, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Number of Detected Genes per Sample (count > 0)",
       x = "Group", y = "Genes Detected") +
  theme(legend.position = "none")

cat("Detected genes summary:\n"); print(summary(detected$n_detected))
```

## 4.3 Gene Count Sparsity

```{r sparsity}
total_entries  <- prod(dim(counts))
cat("Sparsity (% zeros):",
    round(100 * sum(counts == 0) / prod(dim(counts)), 1), "%\n")
```

---

# 5. Expression Distributions

## 5.1 Per-Sample Density (log2)

```{r density-plots}
log2_counts <- log2(counts + 1)

# Per-sample density
log2_long <- log2_counts %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  tidyr::pivot_longer(-gene, names_to = "GSM", values_to = "log2_count") %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group), by = "GSM")

ggplot(log2_long, aes(x = log2_count, group = GSM, color = group)) +
  geom_density(linewidth = 0.3, alpha = 0.5) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Per-Sample Density of log2(count + 1)",
       x = "log2(count + 1)", y = "Density", color = "Group") +
  xlim(0, 20)
```

## 5.2 Boxplots by Sample

```{r boxplots, fig.height=9}
ggplot(log2_long, aes(x = GSM, y = log2_count, fill = group)) +
  geom_boxplot(outlier.size = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-Sample Expression Distribution",
       x = "Sample", y = "log2(count + 1)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) +
  coord_cartesian(ylim = c(0, 18))
```

## 5.3 Mean-Variance Relationship

RNA-Seq count data is overdispersed (variance > mean), which is why
DESeq2 uses a negative binomial model rather than simple tests.

```{r mean-variance}
gene_stats <- data.frame(
  mean_count = rowMeans(counts),
  var_count  = apply(counts, 1, var)
)

ggplot(gene_stats, aes(x = log10(mean_count + 1), y = log10(var_count + 1))) +
  geom_point(alpha = 0.05, size = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 0.8) +
  labs(title = "Mean-Variance Relationship (Raw Counts)",
       subtitle = "Red dashed = Poisson (var = mean). Points above = overdispersion.",
              x = "log10(Mean + 1)", y = "log10(Variance + 1)")
```

---

# 6. Gene Filtering

We remove genes with very low counts to reduce noise and lessen the
multiple-testing burden. We require at least N samples with count >= 10,
where N = the smallest group size.

```{r prefilter}
smallest_group <- min(table(annot$group))
cat("Smallest group size:", smallest_group, "\n")
keep <- rowSums(counts >= 10) >= smallest_group
counts_filtered <- counts[keep, ]
cat("Before:", nrow(counts), "genes\n")
cat("After: ", nrow(counts_filtered), "genes\n")
cat("Removed:", nrow(counts) - nrow(counts_filtered),
    "(", round(100 * (1 - nrow(counts_filtered)/nrow(counts)), 1), "%)\n")

```

```{r post-filter-biotypes}

biotype_after <- gene_annot %>%
  dplyr::filter(GeneID %in% as.numeric(rownames(counts_filtered))) %>%
  dplyr::count(GeneType, sort = TRUE) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1))

print(biotype_after)
```

---

# 7. DESeq2 Normalization

## 7.1 Create DESeqDataSet

The `design = ~ group` formula tells DESeq2 to model expression as a
function of our experimental groups (Control, Lyme_V1, Lyme_V2, Lyme_V5).

```{r create-dds}
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered,
  colData   = annot,
  design    = ~ group
)

cat("DESeqDataSet:", nrow(dds), "genes x", ncol(dds), "samples\n")
```

## 7.2 Size Factors

DESeq2's median-of-ratios normalization estimates a size factor per
sample, correcting for sequencing depth differences. This is superior to
total-count normalization and far better than the FPKM used in the
original study.

```{r size-factors}
dds <- estimateSizeFactors(dds)

sf_df <- data.frame(
  GSM = colnames(dds), size_factor = sizeFactors(dds),
  lib_size = colSums(counts_filtered)
) %>%
  dplyr::left_join(annot %>% dplyr::select(GSM, group, Observation), by = "GSM")


ggplot(sf_df, aes(x = lib_size / 1e6, y = size_factor, color = group)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "gray40", linetype = "dashed") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "DESeq2 Size Factor vs Library Size",
       x = "Library Size (millions)", y = "Size Factor")

```

## 7.3 Variance Stabilizing Transformation (VST)

Raw counts are unsuitable for visualization because of the mean-variance
relationship. VST produces values on a log2-like scale with approximately
constant variance, appropriate for PCA and heatmaps.
```{r vst}
vsd <- vst(dds, blind = FALSE)

cat("VST value range:", round(range(assay(vsd)), 2), "\n")
```

```{r vst-effect, fig.height=5}
par(mfrow = c(1, 2))
plot(rowMeans(log2(counts_filtered + 1)),
     apply(log2(counts_filtered + 1), 1, sd),
     pch = 16, cex = 0.3, col = rgb(0.2, 0.4, 0.8, 0.1),
     xlab = "Mean", ylab = "SD", main = "Before: log2(counts+1)")
plot(rowMeans(assay(vsd)), apply(assay(vsd), 1, sd),     pch = 16, cex = 0.3, col = rgb(0.8, 0.2, 0.2, 0.1),
     xlab = "Mean", ylab = "SD", main = "After: VST")
par(mfrow = c(1, 1))
```

---

# 8. Principal Component Analysis (PCA)

PCA reveals the dominant axes of variation. We want biological signal
(time point, disease state) to drive separation.

## 8.1 Compute PCA (top 2000 most variable genes)

```{r pca-compute}
# Use top 2000 most variable genes
rv <- rowVars(assay(vsd))
top_idx <- order(rv, decreasing = TRUE)[1:min(2000, length(rv))]

pca <- prcomp(t(assay(vsd)[top_idx, ]), center = TRUE, scale. = TRUE)
var_pct <- round(100 * summary(pca)$importance[2, ], 1)

pca_df <- data.frame(pca$x[, 1:5]) %>%
  tibble::rownames_to_column("GSM") %>%
  dplyr::left_join(annot, by = "GSM")
```

## 8.2 Scree Plot

```{r scree}
scree <- data.frame(PC = 1:20, Variance = var_pct[1:20])

ggplot(scree, aes(x = PC, y = Variance)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_line(color = "darkred", linewidth = 0.8) +
  geom_point(color = "darkred", size = 2) +
  labs(title = "Scree Plot", x = "Principal Component",
       y = "% Variance Explained") +
  scale_x_continuous(breaks = 1:20)

cat("Cumulative variance (PC1-5):", cumsum(var_pct[1:5]), "\n")
```

## 8.3 PCA by Group (Time Point)

This is the key plot -- do Lyme patients at different time points
separate from controls?


```{r pca-group, fig.width=15, fig.height=5}
make_pca <- function(df, pcx, pcy, vpct) {
  ggplot(df, aes(x = .data[[pcx]], y = .data[[pcy]], color = group)) +
    geom_point(size = 3, alpha = 0.8) +
    stat_ellipse(level = 0.68, linetype = "dashed") +
    scale_color_brewer(palette = "Set2") +
    labs(x = paste0(pcx, " (", vpct[as.numeric(gsub("PC","",pcx))], "%)"),
         y = paste0(pcy, " (", vpct[as.numeric(gsub("PC","",pcy))], "%)"))}

grid.arrange(
  make_pca(pca_df, "PC1", "PC2", var_pct) + ggtitle("PC1 vs PC2"),
  make_pca(pca_df, "PC1", "PC3", var_pct) + ggtitle("PC1 vs PC3"),
  make_pca(pca_df, "PC2", "PC3", var_pct) + ggtitle("PC2 vs PC3"),
  ncol = 3
)
```

## 8.4 PCA: Disease vs Control with Visit Shape

```{r pca-condition}
ggplot(pca_df, aes(x = PC1, y = PC2, color = condition, shape = visit)) +
  geom_point(size = 3.5, alpha = 0.8) +
  scale_color_manual(values = c(Control = "#2ca02c", Lyme = "#d62728")) +
  scale_shape_manual(values = c(Control = 16, V1 = 17, V2 = 15, V5 = 18)) +
  labs(title = "PCA: Lyme Disease vs Controls",
       x = paste0("PC1 (", var_pct[1], "%)"),
       y = paste0("PC2 (", var_pct[2], "%)"))
```

## 8.5 PCA: Patient Trajectories Over Time

Lines connect the same patient across V1 -> V2 -> V5, showing how their
transcriptome moves through PCA space during treatment and recovery.

```{r pca-trajectories, fig.height=7}
lyme_pca <- pca_df %>% dplyr::filter(condition == "Lyme") %>%
  dplyr::arrange(patient_id, visit)

ggplot() +
  geom_point(data = pca_df %>% dplyr::filter(condition == "Control"),
             aes(x = PC1, y = PC2), color = "gray60", size = 3, alpha = 0.5) +
  geom_path(data = lyme_pca,
            aes(x = PC1, y = PC2, group = patient_id),
            color = "gray40", alpha = 0.3, linewidth = 0.5) +
  geom_point(data = lyme_pca,
             aes(x = PC1, y = PC2, color = visit, shape = visit), size = 3) +
  scale_color_manual(values = c(V1 = "#d62728", V2 = "#ff7f0e", V5 = "#1f77b4")) +
  labs(title = "Patient Trajectories Through PCA Space",
       subtitle = "Gray = controls. Lines connect same patient across visits.",
       x = paste0("PC1 (", var_pct[1], "%)"),
       y = paste0("PC2 (", var_pct[2], "%)"))
```


---

# 9. Sample Relationships

## 9.1 Distance Heatmap

```{r distance-heatmap, fig.height=9, fig.width=10}
sample_dists <- dist(t(assay(vsd)))
dist_mat     <- as.matrix(sample_dists)

row_annot <- data.frame(Group = annot$group, row.names = annot$GSM)
ann_colors <- list(Group = c(Control = "#66C2A5", Lyme_V1 = "#FC8D62",
                              Lyme_V2 = "#8DA0CB", Lyme_V5 = "#E78AC3"))

ann_colors <- list(
  Group = c(Control = "#66C2A5", Lyme_V1 = "#FC8D62",
            Lyme_V2 = "#8DA0CB", Lyme_V5 = "#E78AC3")
)

pheatmap(dist_mat,
         clustering_distance_rows = sample_dists,
         clustering_distance_cols = sample_dists,
         annotation_row = row_annot, annotation_col = row_annot,
         annotation_colors = ann_colors,
         show_rownames = FALSE, show_colnames = FALSE,
         color = viridis(100),
         main = "Sample-to-Sample Euclidean Distance (VST)")
```

## 9.2 Correlation Heatmap

```{r corr-heatmap, fig.height=9, fig.width=10}
pheatmap(cor(assay(vsd)),
         annotation_row = row_annot, annotation_col = row_annot,
         annotation_colors = ann_colors,
         show_rownames = FALSE, show_colnames = FALSE,
         color = colorRampPalette(brewer.pal(9, "RdYlBu"))(100),
         main = "Sample Pearson Correlation (VST)")
```

## 9.3 Top 50 Most Variable Genes

```{r top50-heatmap, fig.height=12, fig.width=10}
top50_idx <- head(order(rv, decreasing = TRUE), 50)
top50_ids <- rownames(assay(vsd))[top50_idx]

top50_symbols <- gene_annot$Symbol[match(as.numeric(top50_ids), gene_annot$GeneID)]
top50_labels  <- ifelse(is.na(top50_symbols) | top50_symbols == "",
                         top50_ids, top50_symbols)
hm_mat <- assay(vsd)[top50_idx, ]
rownames(hm_mat) <- top50_labels
hm_scaled <- t(scale(t(hm_mat)))

pheatmap(hm_scaled,
         annotation_col = row_annot,
         annotation_colors = ann_colors,
         show_colnames = FALSE,
         fontsize_row = 7,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
         main = "Top 50 Most Variable Genes (Z-scored VST)",
         cutree_cols = 4,
         cluster_rows = TRUE, cluster_cols = TRUE)
```

---

# 10. Longitudinal Completeness

```{r longitudinal}
annot %>%
  dplyr::filter(condition == "Lyme") %>%
  dplyr::group_by(patient_id) %>%
  dplyr::summarise(n_visits = dplyr::n_distinct(visit),
                   visits = paste(sort(as.character(visit)), collapse=", "),
                   .groups = "drop") %>%
  dplyr::count(n_visits, visits) %>%
  print()
```

---

# 11. Differential Expression with DESeq2

## 11.1 Run DESeq2

This performs: (1) size factor estimation, (2) gene-wise dispersion
estimation, (3) negative binomial GLM fitting, and (4) Wald tests.

```{r run-deseq}
dds <- DESeq(dds)
resultsNames(dds)
```

## 11.2 Extract and Shrink Results

`lfcShrink()` with the `ashr` method produces moderated log2 fold change
estimates that are more reliable for genes with low counts or high
variability (Stephens 2016).

```{r extract-results}
contrasts <- list(
  V1 = c("group", "Lyme_V1", "Control"),
  V2 = c("group", "Lyme_V2", "Control"),
  V5 = c("group", "Lyme_V5", "Control")
)

# Helper: extract, shrink, and annotate results
get_annotated_results <- function(dds, contrast, gene_annot) {
  res <- results(dds, contrast = contrast, alpha = 0.05)
  res_shrunk <- lfcShrink(dds, contrast = contrast, type = "ashr", res = res)
  as.data.frame(res_shrunk) %>%
    tibble::rownames_to_column("GeneID") %>%
    dplyr::mutate(GeneID_num = as.numeric(GeneID)) %>%
    dplyr::left_join(
      gene_annot %>% dplyr::select(GeneID, Symbol, Description, GeneType),
      by = c("GeneID_num" = "GeneID")) %>%
    dplyr::select(-GeneID_num) %>%
    dplyr::arrange(padj)
}
res_list <- lapply(contrasts, function(ct)
  get_annotated_results(dds, ct, gene_annot))
names(res_list) <- names(contrasts)
```

## 11.3 DEG Summary

```{r deg-summary}
deg_summary <- lapply(names(res_list), function(nm) {
  df <- res_list[[nm]]
  sig <- df %>% dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1)
  data.frame(
    Comparison = paste0(nm, " vs Control"),
    Total_DEGs = nrow(sig),
    Up   = sum(sig$log2FoldChange > 0),
    Down = sum(sig$log2FoldChange < 0)
  )
}) %>% bind_rows()
print(deg_summary)
```

## 11.4 Volcano Plots

```{r volcano, fig.width=16, fig.height=6}
make_volcano <- function(df, title) {
  df <- df %>%
    dplyr::mutate(sig = dplyr::case_when(
      is.na(padj) ~ "NS",
      padj < 0.05 & log2FoldChange > 1  ~ "Up",
      padj < 0.05 & log2FoldChange < -1 ~ "Down",
      TRUE ~ "NS"))
  top <- df %>% dplyr::filter(sig != "NS") %>% head(10)
  ggplot(df, aes(x = log2FoldChange, y = -log10(pvalue), color = sig)) +
    geom_point(alpha = 0.4, size = 1) +
    scale_color_manual(values = c(Up = "firebrick", Down = "steelblue", NS = "gray70")) +
    geom_text_repel(data = top, aes(label = Symbol), size = 2.5,
                    max.overlaps = 15) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
    labs(title = title, x = "log2 Fold Change", y = "-log10(p-value)") +
    theme(legend.position = "bottom")
}

grid.arrange(
  make_volcano(res_list$V1, "V1 (Acute) vs Control"),
  make_volcano(res_list$V2, "V2 (Post-Treatment) vs Control"),
  make_volcano(res_list$V5, "V5 (6 Months) vs Control"),
  ncol = 3
)
```

## 11.5 DEG Overlap Across Time Points

How many DEGs are shared across time points?

```{r deg-overlap}
sig_genes <- lapply(res_list, function(df) {
  df %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1) %>%
    dplyr::pull(GeneID)
})
cat("DEGs per time point: V1=", length(sig_genes$V1),
    " V2=", length(sig_genes$V2), " V5=", length(sig_genes$V5), "\n")
cat("V1 & V2:", length(intersect(sig_genes$V1, sig_genes$V2)), "\n")
cat("V1 & V5:", length(intersect(sig_genes$V1, sig_genes$V5)), "\n")
cat("V2 & V5:", length(intersect(sig_genes$V2, sig_genes$V5)), "\n")
cat("All 3:  ", length(Reduce(intersect, sig_genes)), "\n")
```

## 11.6 Save Results
```{r save-results}
for (nm in names(res_list)) {
  write.csv(res_list[[nm]],
            file.path(results_dir, paste0("DEG_", nm, "_vs_Control.csv")),
            row.names = FALSE)
}
cat("DEG tables saved to:", results_dir, "\n")
```

---
# 12. Downstream: 1st Generation -- ORA

Over-Representation Analysis tests whether our DEG list contains more
genes from a pathway than expected by chance (Khatri et al. 2012,
Chapter 8.1.1 of Mohrkeg).
We use `clusterProfiler::enricher()` with gene sets from MSigDB
(Chapter 8.2 of Mohrkeg), specifically the **Hallmark** (H) and
**Gene Ontology** (C5) collections.

## 12.1 Prepare MSigDB Gene Sets

```{r msigdb-setup}
msigdb <- msigdbr(species = "Homo sapiens")
cat("MSigDB entries:", nrow(msigdb), "\n")
cat("Collections:", paste(unique(msigdb$gs_collection), collapse = ", "), "\n")
```

## 12.2 ORA: Hallmark Gene Sets (V1)
The Hallmark collection contains 50 well-defined biological processes.

```{r ora-hallmark}
# Build term-to-gene mapping (Chapter 8.2 / 8.4.1 of Mohrkeg)
term2gene_H <- msigdb %>%
  dplyr::filter(gs_collection == "H") %>%
  dplyr::select(gs_name, gene_symbol)
# Significant DEGs (our gene list)
sig_symbols_v1 <- res_list$V1 %>%
  dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                !is.na(Symbol) & Symbol != "") %>%
  dplyr::pull(Symbol) %>% unique()
# Background universe = all tested genes
universe_v1 <- res_list$V1 %>%
  dplyr::filter(!is.na(Symbol) & Symbol != "") %>%
  dplyr::pull(Symbol) %>% unique()
cat("DEGs for ORA:", length(sig_symbols_v1), "\n")
cat("Universe:", length(universe_v1), "\n")
ora_hallmark_v1 <- enricher(
  gene       = sig_symbols_v1,
  universe   = universe_v1,
  TERM2GENE  = term2gene_H,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  qvalueCutoff  = 1
)

ora_h_df <- as.data.frame(ora_hallmark_v1)
cat("Significant Hallmark pathways:", nrow(ora_h_df), "\n")
print(ora_h_df %>% dplyr::select(ID, GeneRatio, p.adjust, Count) %>% head(10))
```

```{r ora-hallmark-dotplot, fig.height=7}
if (nrow(ora_h_df) > 0) {
  dotplot(ora_hallmark_v1, showCategory = 15) +
    ggtitle("ORA: Hallmark Pathways Enriched in V1 DEGs")
}
```
## 12.3 ORA: Gene Ontology (C5) for V1
```{r ora-go}
term2gene_C5 <- msigdb %>%
  dplyr::filter(gs_collection == "C5") %>%
  dplyr::select(gs_name, gene_symbol)
term2name_C5 <- msigdb %>%
  dplyr::filter(gs_collection == "C5") %>%
  dplyr::select(gs_name, gs_description) %>%
  dplyr::distinct()
ora_go_v1 <- enricher(
  gene       = sig_symbols_v1,
  universe   = universe_v1,
  TERM2GENE  = term2gene_C5,
  TERM2NAME  = term2name_C5,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  qvalueCutoff  = 1
)
ora_go_df <- as.data.frame(ora_go_v1)
cat("Significant GO terms:", nrow(ora_go_df[ora_go_df$p.adjust < 0.05, ]), "\n")
```

```{r ora-go-dotplot, fig.height=9}
if (nrow(ora_go_df) > 0) {
  dotplot(ora_go_v1, showCategory = 20) +
    ggtitle("ORA: Top GO Terms Enriched in V1 DEGs")
}
```

## 12.4 ORA Across All Time Points

```{r ora-all-timepoints}
ora_results <- list()
for (tp in names(res_list)) {
  sig_sym <- res_list[[tp]] %>%
    dplyr::filter(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1 &
                  !is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()
  univ_sym <- res_list[[tp]] %>%
    dplyr::filter(!is.na(Symbol) & Symbol != "") %>%
    dplyr::pull(Symbol) %>% unique()

  ora_results[[tp]] <- enricher(
    gene = sig_sym, universe = univ_sym,
    TERM2GENE = term2gene_H,
    pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 1
  )
}

cat("Significant Hallmark pathways per time point:\n")
for (tp in names(ora_results))
  cat("  ", tp, ":", nrow(as.data.frame(ora_results[[tp]])), "\n")
```

---

# 13. Downstream: 2nd Generation -- GSEA

Gene Set Enrichment Analysis (Subramanian et al. 2005) uses the **full
ranked gene list** (by log2 fold change), not just the significant genes.
This avoids arbitrary cutoffs and can detect subtle coordinated changes
that ORA misses (Chapter 8.1.2.1 of Mohrkeg).

This avoids arbitrary cutoffs and can detect subtle coordinated changes
that ORA misses (Chapter 8.1.2.1 of Mohrkeg).
ora_v1 <- enricher(
  gene = sig_symbols_v1,
  universe = universe_symbols,
  TERM2GENE = term2gene,
  TERM2NAME = term2name,
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH"
## 13.1 Build Ranked Gene Lists
```{r gsea-ranked-lists}
make_ranked_list <- function(res_df) {
  res_df %>%
    dplyr::filter(!is.na(Symbol) & Symbol != "" & !is.na(log2FoldChange)) %>%
    dplyr::arrange(dplyr::desc(log2FoldChange)) %>%
    dplyr::distinct(Symbol, .keep_all = TRUE) %>%
    dplyr::select(Symbol, log2FoldChange) %>%
    tibble::deframe()
}
ranked_lists <- lapply(res_list, make_ranked_list)
cat("Ranked list sizes:", sapply(ranked_lists, length), "\n")
```
## 13.2 GSEA: Hallmark (V1)
```{r gsea-hallmark-v1}
gsea_hallmark_v1 <- GSEA(
  geneList  = ranked_lists$V1,
  TERM2GENE = term2gene_H,
  pvalueCutoff = 1,
  eps = 0
)

gsea_h_df <- as.data.frame(gsea_hallmark_v1)
cat("GSEA Hallmark results:", nrow(gsea_h_df), "terms tested\n")
cat("Significant (padj<0.05):", sum(gsea_h_df$p.adjust < 0.05), "\n")
gsea_h_df %>%
  dplyr::filter(p.adjust < 0.05) %>%
  dplyr::select(ID, NES, pvalue, p.adjust) %>%
  dplyr::arrange(dplyr::desc(abs(NES))) %>%
  head(15) %>%
  print()
```

```{r gsea-hallmark-dotplot, fig.height=8}
if (sum(gsea_h_df$p.adjust < 0.05) > 0) {
  dotplot(gsea_hallmark_v1, showCategory = 20,
          split = ".sign") + facet_grid(~ .sign) +
    ggtitle("GSEA: Hallmark Pathways (V1 vs Control)")
}
```

## 13.3 GSEA Enrichment Plots for Key Pathways

These "mountain plots" show how the genes in a pathway are distributed
across the ranked list. A peak at the left = upregulated, right =
downregulated.

```{r gsea-enrichment-plots, fig.height=4, fig.width=8}
library(enrichplot)
key_pathways <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_COMPLEMENT"
)
for (pw in key_pathways) {
  if (pw %in% gsea_h_df$ID) {
    print(gseaplot2(gsea_hallmark_v1, geneSetID = pw, title = pw))
  }
}
```

## 13.4 GSEA Across All Time Points

```{r gsea-all-timepoints}
gsea_results <- list()
for (tp in names(ranked_lists)) {
  gsea_results[[tp]] <- GSEA(
    geneList  = ranked_lists[[tp]],
    TERM2GENE = term2gene_H,
    pvalueCutoff = 1,
    eps = 0
  )
# Compare NES across time points for significant pathways
nes_comparison <- lapply(names(gsea_results), function(tp) {
  df <- as.data.frame(gsea_results[[tp]])
  if (nrow(df) > 0) {
    df %>%
      dplyr::select(ID, NES, p.adjust) %>%
      dplyr::mutate(timepoint = tp)
  }
}) %>% bind_rows()
cat("Pathways tested across all time points:\n")
nes_wide <- nes_comparison %>%
  dplyr::select(ID, NES, timepoint) %>%
  tidyr::pivot_wider(names_from = timepoint, values_from = NES)
print(nes_wide %>% head(15))
```


```{r nes-heatmap, fig.height=8, fig.width=7}
# Heatmap of NES values across time points
sig_any <- nes_comparison %>%
  dplyr::filter(p.adjust < 0.05) %>%
  dplyr::pull(ID) %>% unique()
if (length(sig_any) > 0) {
  nes_mat <- nes_comparison %>%
    dplyr::filter(ID %in% sig_any) %>%
    dplyr::select(ID, NES, timepoint) %>%
    tidyr::pivot_wider(names_from = timepoint, values_from = NES) %>%
    tibble::column_to_rownames("ID") %>%
    as.matrix()
  nes_mat[is.na(nes_mat)] <- 0
  pheatmap(nes_mat,
           color = colorRampPalette(c("steelblue", "white", "firebrick"))(100),
           cluster_cols = FALSE,
           main = "GSEA NES: Hallmark Pathways Across Time Points",
           fontsize_row = 7)
}
```

---
# 14. Save All Downstream Results
```{r save-downstream}
# ORA results
for (tp in names(ora_results)) {
  df <- as.data.frame(ora_results[[tp]])
  if (nrow(df) > 0)
    write.csv(df, file.path(results_dir, paste0("ORA_Hallmark_", tp, ".csv")),
              row.names = FALSE)
}
# GSEA results
for (tp in names(gsea_results)) {
  df <- as.data.frame(gsea_results[[tp]])
  if (nrow(df) > 0)
    write.csv(df, file.path(results_dir, paste0("GSEA_Hallmark_", tp, ".csv")),
              row.names = FALSE)
}
cat("All results saved to:", results_dir, "\n")
```


---
# 15. Summary of Findings
```{r findings-summary}
cat("============================================================\n")
cat("          ANALYSIS SUMMARY\n")
cat("============================================================\n\n")
cat("DATA:\n")
cat("  Reference genome: GRCh38.p13 (vs hg19 in original)\n")
cat("  Samples analyzed:", ncol(counts_filtered), "\n")
cat("  Genes after filtering:", nrow(counts_filtered), "\n\n")
cat("DIFFERENTIAL EXPRESSION (DESeq2, |lfc|>1, padj<0.05):\n")
print(deg_summary)
cat("\nORIGINAL STUDY (Limma/Voom, |FC|>1.5, padj<0.001):\n")
cat("  V1: 1235 DEGs | V2: 1060 DEGs | V5: 686 DEGs\n\n")
cat("PATHWAY ANALYSIS:\n")
cat("  ORA Hallmark (V1):", nrow(as.data.frame(ora_hallmark_v1)), "enriched\n")
cat("  ORA GO (V1):", sum(as.data.frame(ora_go_v1)$p.adjust < 0.05), "enriched\n")
cat("  GSEA Hallmark (V1):", sum(as.data.frame(gsea_hallmark_v1)$p.adjust < 0.05),
    "significant\n")
cat("\nKEY IMPROVEMENTS OVER ORIGINAL:\n")
cat("  1. Modern genome reference (GRCh38 vs hg19)\n")
cat("  2. Raw counts + DESeq2 (vs FPKM + Limma/Voom)\n")
cat("  3. Open-source pathway analysis (clusterProfiler + MSigDB vs IPA)\n")
cat("  4. Added GSEA (2nd generation, not in original study)\n")
cat("============================================================\n")
```
---

# Session Info

```{r session-info}
sessionInfo()
```
